Spinal V1 inhibitory interneuron clades differ in birthdate, projections to motoneurons and heterogeneity

Version 1

Abstract

Abstract

Spinal cord interneurons play a crucial role in shaping motor output, but their precise identity and circuit connectivity remain unclear. Focusing on the cardinal class of inhibitory V1 interneurons, we define the diversity of four major V1 subsets according to timing of neurogenesis, genetic lineage-tracing, synaptic output to motoneurons, and synaptic inputs from muscle afferents. Birthdating delineates two early-born (Renshaw and Pou6f2) and two late-born V1 clades (Foxp2 and Sp8) suggesting sequential neurogenesis gives rise to different V1 clades. Neurogenesis did not correlate with motoneuron targeting. Early-born Renshaw cells and late-born Foxp2-V1 interneurons both tightly coupled to motoneurons, while early-born Pou6f2-V1 and late-born Sp8-V1 interneurons did not. V1-clades also greatly differ in cell numbers and diversity. Lineage labeling of the Foxp2-V1 clade shows it contains over half of all V1 interneurons and provides the largest inhibitory input to motoneuron cell bodies. Foxp2-V1 subgroups differ in neurogenesis and proprioceptive input. Notably, one subgroup defined by Otp expression and located adjacent to the lateral motor column exhibits substantial input from proprioceptors, consistent with some Foxp2-V1 cells at this location forming part of reciprocal inhibitory pathways. This was confirmed with viral tracing methods for ankle flexors and extensors. The results validate the previous V1 clade classification as representing unique interneuron subtypes that differ in circuit placement with Foxp2-V1s forming the more complex subgroup. We discuss how V1 organizational diversity enables understanding of their roles in motor control, with implications for the ontogenetic and phylogenetic origins of their diversity.

Introduction

The spinal cord contains a diversity of interneurons that lend to its vast computational power with inhibitory interneurons critically modulating and patterning motoneuron firing to adjust timing and force of muscle contractions. We focus in here on a major group of ventral inhibitory interneurons known as V1 that originate from p1 progenitors, express the transcription factor (TF) Engrailed-1 (En1) and send ipsilateral axons to the ventral horn where they densely innervate motoneuron cell bodies and proximal dendrites (Alvarez et al., 2005). Based on differential expression of calcium buffering proteins and synaptology this early study proposed a diversity of phenotypes and circuit roles for V1 interneurons. Later electrophysiological and modeling studies showed that V1 interneurons play crucial roles in shaping motor output by modulating locomotor speed, governing flexion-extension at the level of central pattern generator (CPG) half-centers and/or last-order reciprocal inhibition of antagonistic motoneurons, and providing recurrent feedback inhibition of motoneuron firing (Sapir et al., 2004; Zhang et al., 2014; Britz et al., 2015; Falgairolle and O’Donovan, 2019, 2021; Shevtsova et al., 2022). Accordingly, there is significant interest in defining the molecular identity and circuit organization of different subtypes of V1 interneurons. This is of further significance because recent findings suggest that early V1 disconnection from motoneurons in motor neurodegenerative diseases such as amyotrophic lateral sclerosis contributes to dysfunction and presages motoneuron death (Wootz et al., 2013; Salamatina et al., 2020; Allodi et al., 2021).

Prior work revealed that V1 diversity can be organized into four major clades according to their positions and molecular identity at postnatal day 0 (P0) (Bikoff et al., 2016). Each clade was defined by expression of unique transcription factors. Co-expression of V-maf musculoaponeurotic fibrosarcoma oncogene homologs A and B (MafA/MafB) defines a clade composed of Renshaw cells, the only homogenous functional V1 subgroup (Benito-Gonzalez and Alvarez, 2012; Stam et al., 2012; Bikoff et al., 2016). The other three clades are respectively defined by expression at P0 of the TFs P.O.U. domain (Pou6f2), Forkhead box P2 (Foxp2) and Specificity protein 8 (Sp8) (Bikoff et al., 2016). Within these three clades, additional diversity was uncovered by further combinations of TF expression and positions (Bikoff et al., 2016; Gabitto et al., 2016; Sweeney et al., 2018). A recent harmonized atlas of several mouse spinal cord transcriptomic studies (Russ et al., 2021) identified seven possible V1 groups, including Renshaw cells, Pou6f2-V1s and three Foxp2-V1 groups. V1 genetic diversity parallels the functional diversity described in physiological and modeling studies, but whether V1 clades occupy specific functional niches in spinal motor circuits remains unclear in part because of lack of information about their synaptic inputs and outputs and their origins.

Here we clarify the origins, diversity, and synaptic relations with motoneurons of the four major V1 clades. V1 neurogenesis was previously divided into early (E9.5 to E10.5) and late phases (E11 to E12.5), each producing distinct V1 interneurons (Benito-Gonzalez and Alvarez, 2012). Early-generated V1 interneurons include MafB positive Renshaw cells, whereas late-generated V1 interneurons comprise Ia reciprocal inhibitory interneurons (IaINs) expressing Foxp2. It is now well accepted that temporal and spatial properties during neurogenesis intersect to create diversity from each spinal cord progenitor domain (Sagner and Briscoe, 2019; Deska-Gauthier et al., 2020; Deska-Gauthier and Zhang, 2021; Osseward et al., 2021; Sagner et al., 2021). Using further nodal intersections between transcriptomics and neurogenesis, one recent report described seven embryonic V1 groups defined by TFs with temporally restricted expression (Delile et al., 2019). However, because of the dynamic nature of TF expression in spinal interneurons, it is difficult to match embryonic TFs in this study to the V1 clades previously defined at P0. Using clade-defining TFs in combination with 5-ethynyl-2’-deoxyuridine (EdU) birthdating, we identified a relationship between neurogenesis timing and clade identity and uncovered additional diversity within Foxp2-V1 interneurons, the largest V1 clade. To study Foxp2-V1 interneurons we used intersectional genetics to lineage label their cell bodies and axons. This revealed subdivisions according to location and expression of additional TFs. We also found that Foxp2-V1s establish the highest density of V1 synapses on proximal somatodendritic membranes of limb-related lateral motor column (LMC) motoneurons. They also include subgroups receiving dense proprioceptive inputs with some having connectivity of reciprocal IaINs. Together with Renshaw cells they contribute the majority of V1 synapses on motoneuron cell bodies and proximal dendrites, while synapses from other clades (Pou6f2 and Sp8) have minimal representation. We conclude that V1 clades differ in time of neurogenesis, internal heterogeneity and synaptic targeting of motoneurons.

Results

V1 interneurons belonging to different clades have distinct timing of neurogenesis

To examine the birthdates of the different V1 clades, we lineage traced all V1 interneurons (INs) using en1cre/+::Ai9R26-tdTomato mice (En1cre-Ai9:tdT) (sometimes intersected with foxp2flpo/+::RCE:dual-EGFP: En1-Foxp2-RCE:GFP; see below) and pulse-labeled developing embryos by injecting pregnant females with EdU at specified 12 hours intervals between E9.5 to E12.5 (Figure 1A). All born pups carrying the genetic V1 lineage markers were analyzed at P5. This age was chosen to preserve as much as possible expression of V1-clade defining TFs (MafB, Pou6f2, Foxp2 and Sp8) for antibody detection. We analyzed spinal cords at E9.5 (n=3; 3 litters), E10 (n=6; 3 litters), E10.5 (n=5; 3 litters), E11 (n=5; 2 litters), E11.5 (n=6; 3 litters), E12 (n=3; 1 litter), and E12.5 (n=4; 2 litters). In all animals we confirmed the expected sequence of cell birthdates in the mammalian spinal cord, from ventro-lateral to dorso-medial locations (Figure 1B) (Altman and Bayer, 2001). One animal pulse-labeled at E11 was removed from the analyses because EdU labeled cells did not correspond to the expected distribution (marked with x in Figure 1D). To ensure only V1 cells that incorporated EdU at the time of injection were included in the analyses, we defined positive neurons as those with more than two-thirds of the nucleus showing homogeneous EdU fluorescence. Neurons with speckles or partial nuclear labeling (Figure 1C) were discarded. Incompletely labeled cells may arise from either partial dilution of EdU after one division cycle or by incorporating EdU during late S-phase or DNA repair events (Packard et al., 1973; Ferreira et al., 1997; Taupin, 2007). Using our labeling criteria, we found varying percentages of V1s are generated at different time points, with a peak at E11 in which 23.4% ±2.9 (mean±SD) of V1s incorporated EdU (Figures 1D,E). After adding together all EdU-labeled V1s at all time points we account for 71.4% of all V1 interneurons. If we include V1 neurons with speckles or partial nuclear labeling, we overrepresent V1s by more than double (255.9%), and obscure differences in EdU labeling at different time points. We nevertheless found overlap among animals’ pulse-labeled in contiguous 12-hour time-points. This is expected since our timed-pregnancies have a ±12-hour resolution. Furthermore, within single litters it is common to find animals with developmental differences of 6 to 12 hours. Finally, although EdU has a very short half-life in mice (1.4 ±0.7 mins; (Cheraghali et al., 1994)) and, therefore, its bioavailability should overlap little with injections separated by 12 hr, estimates of S-phase duration during neurogenesis vary between 4 and 17 hrs (Ponti et al., 2013) suggesting other opportunities to create overlap from EdU injections spaced by 12 hr. These sources of variability are inherent to the technique and explain the relatively larger standard deviations on the percentages of V1s labeled at time points with the largest change in birthdates, around peak V1 neurogenesis. In summary, using rigorous criteria for defining EdU labeling and averaging animals from different litters allows to derive the strongest conclusions possible when estimating differences in birthdates.

Neurogenesis order of V1 clades assayed by EdU birthdating.

A, Experimental design. Timed pregnant En1cre-Ai9:tdT females were EdU injected at one of seven time points between E9.5 and E12.5 and the spinal cords were harvested at P5. Tissue sections were processed for EdU (Click-iT™) and immunostained for transcription factors (TFs) representing major V1 clades. MafB was used to identify the MafA-Renshaw cell clade by location (“ventral MafB-V1 cells”). This is because MafA antibodies in our hands labeled few Renshaw cells. B, Patterns of EdU labeling at different embryonic times. Spinal neurons are born in a ventrolateral to dorsomedial sequence. Until E11, mostly ventral neurons are generated. C, Example of E11 EdU labeling in an En1-tdT spinal cord. EdU integrated in the DNA at the time of injection is diluted with subsequent cell divisions. To ensure we only counted V1 cells that incorporated EdU during S-phase after their final division, we only counted V1 cells with nuclei largely filled by EdU Click-iT™ reaction (arrows indicate four such strongly labeled V1 EdU+ cells). D, Percentages of V1 cells labeled with “strong” EdU in each mouse. The x-axis indicates individual animals (“<litter number>.<animal number>”). For the most part there was consistency in the percentage of EdU-labeled V1 interneurons at each time point, although we also noted variability between litters and among animals within a litter. One animal (459.2) showed the wrong EdU pattern for its injection age and was discarded (indicated by an X). E, EdU birthdating reveals a peak in V1 neurogenesis around E11 (error bars = SD; each dot represents one animal). At the time points flanking the peak there is a larger amount of variability, suggesting a fast-changing pace of V1 neurogenesis. F, Representative images of TF antibody staining combined with EdU labeling to determine birthdates of defined V1 clades. The time points represented were selected according to the maximal or near maximal generation of V1 interneurons in each clade. G-I, V1-clade neurogenesis quantification. Graphs represent the average ±SEM calculated from n = 3.9 ±0.3 mice per TF/date (not all TFs were tested in all mice). For each mouse average, we analyzed four ventral horns in Lumbar 4 or 5 segments. G, Percentage of V1s expressing each clade-specific TF labeled with EdU at each embryonic time point. Ventral MafB-V1s (Renshaw cells) and Pou6f2-V1s are mostly born before E11 (dorsal MafB-V1s are a subgroup of Pou6f2-V1s). Foxp2-V1 and Sp8-V1 interneurons have wider windows of neurogenesis, but most are generated after E11. H, Data normalized to the maximum percentage of V1s born in each group to compare peak generation in each clade. Ventral MafB-V1s, Pou6f2-V1s, Foxp2-V1s and Sp8-V1s have progressively later times of peak generation. I, Cum-sum graphs of V1-clades neurogenesis. Between 50% to 68% of all neurons in each V1 clade are labeled across all ages. By E11 almost all early-clade neurons are generated, while less than half of neurons in late-clades are.

To analyze birthdates of V1 interneurons identified by clade specific markers (Figure 1F-I), we first compared the number of V1s expressing clade-defining TFs at P5 with previous estimates at P0 (Bikoff et al., 2016). Overall, MafB-V1s were 9.4% ±1.6 (±S.D.) of the whole V1 population (n = 21 animals) which differs to 25% at P0 form Bikoff et al (2016). Small differences were found for Pou6f2-V1s 8.1% ±4.6 (n = 19) compared to 13% at P0; Foxp2-V1s 32.5% ±8.4 (n = 26) compared to 34% and Sp8-V1s 8.8% ±2.6 (n = 26) compared to 13%. Differences in Pou6f2-V1, Foxp2-V1 and Sp8-V1 cells might be due to small differences in expression due to age and/or immunocytochemical (ICC) sensitivity. In contrast, the large differences found with multi-clade MafB-V1s likely result from rapid downregulation of MafB in some V1 clades after birth. MafB is expressed at P0-P5 in three V1 clades. Ventral MafA-calbindin Renshaw cells, a subpopulation of dorsal Pou6f2-V1s, and subgroups of Foxp2-V1s distributed throughout the ventral horn. MafB is quickly downregulated after birth in Foxp2-V1s, but expression is maintained in Pou6f2-V1s and ventral Renshaw cells (see below). When considering only ventral MafB-V1s located in the Renshaw area (Renshaw V1 clade) we obtained at P5 a percentage not different to the percentage at P0 for the MafA-Renshaw cell V1 clade (5.5% ±1.7, n = 21 vs 5% (Bikoff et al., 2016)). We conclude that TF ICC detection at P5 provides an accurate representation of V1 clades previously defined at P0.

Birthdating divides V1 clades into two groups: early-born and late-born. Most early-born V1 cells are EdU-labeled before E11 and include Renshaw cells and Pou6f2-V1s. Most cells from the two other clades (Foxp2-V1s and Sp8-V1s) are later-born (after E11) (Figures 1F-I). These data derive from 68,562 V1s sampled from 29 animals in 4 ventral horns per animal/TF/Edu time point (average 197.0 V1s per ventral horn). V1 clades differ in their peak times and the spread of birthdates, with Foxp2-V1s and Sp8-V1s showing the larger spreads and later peaks (Figure 1G-H). Pou6f2-V1s, dorsal MafB-V1s, and ventral MafB-V1s display narrower spreads of birthdates and earlier peaks. Normalizing to peaks reveals that ventral MafB-V1 Renshaw cells are generated the earliest (E10.5 peak), followed by Pou6f2-V1s including dorsal MafB-V1s (E11.0 peak), Foxp2-V1s (E11.0-E11.5 peak) and Sp8-V1s (E12 peak). Around 50% of all V1s are born by E11, including most cells in early-born clades and a proportion of Foxp2-V1s and Sp8-V1s (Figure 1I). Less than 10% of Foxp2-V1s and Sp8-V1s are generated before E10.5 and almost no ventral MafB-V1s or Pou6f2-V1s (including the dorsal MafB-V1s) are generated after E11 (Figure 1I). After E12, only Sp8-V1s are generated. Cumulative graphs through all ages show that overall, EdU labeling accounted for 67.3% of all ventral (Renshaw) MafB-V1s, 62.4% of Pou6f2-V1s, 68.2% of dorsal MafB-V1s, 49.8% of Foxp2-V1s and 55.8% of Sp8-V1s. This suggests significant representation from all clades in our EdU analyses (Figure 1I).

Next, we examined whether clade-specific V1 interneurons with different birthdates settle in different positions. We plotted the positions of V1 interneurons from each clade identified by the TFs MafB, Pou6f2, Foxp2 and Sp8 and that are born at different embryonic times (Figure 2). We compared time windows between E10 and E12 because few V1s were found with strong EdU at E9.5 and E12.5 which reduced the accuracy of cellular density plots. V1s within these clades showed subpopulations located at different positions according to birthdate. The earliest-born MafB-V1s accumulate ventrally in the “Renshaw-cell area”. Later-born MafB-V1s occupy more dorsal positions because they are a subgroup of Pou6f2-V1s. Pou6f2-V1s born at different time points always settle in dorsal positions. In contrast, Foxp2-V1s are a large group whose generation lasts the whole V1 neurogenesis period. They show large differences in localization according to birthdate. Foxp2-V1s born at E10 occupy dorsal locations while those born between E10.5 and E11.5 settle laterally: close to the lateral motor column (LMC). There is a dorsal to ventral progression of Foxp2-V1 cell born from E10.5 to E11. Foxp2-V1s born at E12 are located ventro-medially. In summary, Foxp2-V1s follow a clockwise rotation in positioning according to their time of neurogenesis. Finally, early born Sp8-V1 cells are located ventrally while later born Sp8-V1 cells are located more dorsally and medially.

V1 interneurons born at different embryonic times settle in different locations.

For each panel in the top row, blue open dots indicate all V1 interneurons positive for each TF at P5, and filled red dots indicate those that also have strong EdU. Each plot is from one representative animal injected at each of the indicated embryonic times (four ventral horns from each animal were analyzed with all cell locations superimposed in the diagram). The bottom rows show cellular density profiles of the V1 interneurons positive for each TF that have strong EdU from all animals EdU-injected at the indicated embryonic times; n indicates the number of animals in each plot. Four ventral horns were analyzed per animal. Contour plots are derived from 2D kernel density estimates of interneuron positions; lines encompass 10% increments. Only timepoints with representative numbers of TF+/EdU+ cells are shown. Blue directional arrows highlight the difference in settling locations for V1s in each clade born at this time point with respect to the previous time point. A, MafB-V1s are divided into two groups: ventral MafB-V1s corresponding to Renshaw cells and later-born dorsal MafB-V1s that correspond to a subgroup of Pou6f2-V1s. B, Pou6f2-V1s are born between E10 and E11.5 and always settle dorsally. C, Foxp2-V1s show large variations in location according to birthdate. Cells of increasingly older birthdates settle dorsally, laterally, ventro-laterally and then ventro-medially. D, A few Sp8-V1s are born early and are located ventrally, but the majority of Sp8-V1s are born later and locate dorsally before moving more medially those born the latest. The directional changes in settling positions according to birthdate is different for each V1 clade.

In summary, different V1 clades are generated through overlapping windows of neurogenesis but with distinct peaks which allow classification into early-born (Renshaw cells and Pou6f2-V1s) and late-born clades (most of Foxp2-V1s and Sp8-V1s). Within clades, subgroups with different birthdates settle at specific positions. This is most evident in the largest Foxp2-V1 clade suggesting significant cellular heterogeneity.

MafB-V1s genetic labeling reveals two main types in the mature spinal cord

We used MafB antibodies to identify the Renshaw cell clade because we were unable to label enough Renshaw cells with MafA antibodies at P5. MafB, however, is also expressed in subpopulations of Pou6f2-V1 and Foxp2-V1 cells at P0 (Bikoff et al., 2016). To better identify these cells, we used a genetic detection strategy by introducing a mafbGFP reporter allele (Moriguchi et al., 2006) in En1cre-Ai9:tdT mice (Figure 3). We also used two different antibodies against MafB of differing sensitivity and specificity (Figure 3 Supplemental). The MafB antibody with the highest specificity (Novus NB600-266)(Supplemental Figure 3.1) displayed weaker labeling of Renshaw cells at P5 in spinal cord sections. In contrast, the antibody with highest sensitivity at P5 (Sigma HPA005653) also showed cross-reactivity with other targets in tissue sections and Western blots. Each antibody was directed against different regions of the mouse mafB gene: aa18-167 for the Sigma antibody and aa100-150 for the Novus antibody. These target sequences are shared with MafA and c-Maf, but the overlap is larger with the Sigma antibody immunogen. Renshaw cells co-express MafA, but Western blots ruled out any cross-reactions with MafA for either of the MafB antibodies (Supplemental Figure 3). However, both antibodies produced a double band between the 37 and 50 kDa markers in Western blots, with only the lower band diminishing with a lower mafB gene dose in heterozygotes or absence in knockouts (Supplemental Figure 3). The upper band was not affected by gene dose. In a parallel western blot, the upper band corresponded to c-Maf. This suggests that in Western blots both MafB antibodies cross reacted with c-Maf. In tissue sections only the Sigma antibody cross reacted with other targets, likely c-Maf. This would explain the larger number of dorsal horn interneurons positive for the Sigma antibody compared to the Novus antibody. Many interneurons at this location are known to express c-Maf (Hu et al., 2012; Frezel et al., 2023). Despite these limitations, MafB detection was similar for both antibodies in Renshaw cells and Pou6f2-V1 cells. Expression of the mafB gene was further confirmed in these cells with the mafBGFP allele. We preferentially used the more sensitive Sigma antibody because it optimally recognized MafB in Renshaw cells at P5. Intriguingly, immunostaining with either antibody could not label any V1 cells in mature spinal cords despite continuous expression of MafB-Green Fluorescent Protein (GFP was detectable after P15 only after amplification of fluorescence with antibodies). This suggests that in addition to progressive downregulation of gene expression, the amount of detectable protein is further reduced at older ages by control of mafB mRNA translation and/or higher TF epitope masking in mature chromatin.

MafB-V1s visualized in a MafB:GFP mouse model.

A, In mature (>P15) mice, MafB-GFP+ cells in lamina VII are mostly contained in the V1 lineage (tdTomato) and distribute to the most dorsal and ventral regions of the distribution of V1s. Those that are also calbindin-IR fall in the typical ventral region occupied by Renshaw cells. In addition, there are many dorsal horn non-V1 MafB-GFP+ neurons. The small cells throughout white and gray matter are microglia. B, Compared to P15, a few more ventral horn neurons express MafB-GFP at P5, including some motoneurons and microglia whose MafB labeling is weaker. Within MafB-GFP+ V1 neurons the two groups located at the most dorsal and most ventral regions correspond to the V1 neurons that retain MafB-GFP at P15. Like P15, the ventrally located group expresses calbindin. The dorsal group expressed Pou6f2 at this age. Little-to-no Foxp2-IR or Sp8-IR is found in either dorsal or ventral groups of MafB-GFP+ V1s. C, Quantitation of MafB-V1 neurons in P5 mice. Around 13% of all V1s express MafB-GFP+, and the percentages located in the Renshaw cell area (“ventral”) or dorsal lamina VII (“dorsal”) are evenly split (n = 17 mice, 4 ventral horns each. Bars show SD). D, At P5, over half of the MafB-GFP+ V1 cells have detectable levels of MafB (Sigma) immunoreactivity in both the ventral and dorsal groups (n = 9 mice, 4 ventral horns each. Bars show SD). E, MafB-GFP+ / Pou6f2-IR V1 cells are exclusively in the dorsal MafB-GFP group, while MafB-GFP+ / calbindin-IR V1 cells are exclusively in the Renshaw cell area, Negligible number of MafB-GFP+ V1 cells express Foxp2 or Sp8. (n=4-5 mice, 4 ventral horns each. Bars show SD). F, EdU birthdating reveals similar proportions of EdU+ neurons at P5 for dorsal MafB-GFP vs MafB-IR neurons labeled at each embryonic time (n = 2 mice per timepoint per condition, 4 ventral horns each. Bars show SD). The mismatch in birthdates between MafB genetic and antibody labeling at E11 in the ventral group could arise because some E11-born ventral MafB-GFP V1s quickly downregulate MafB to levels undetectable with antibodies.

MafB-V1s are best identified in mafBGFP/+::En1cre-Ai9:tdT mice, particularly after amplification of GFP fluorescence with antibodies. In these animals, GFP is present in a variety of spinal neurons in the dorsal and ventral horns and in small microglia. In mature (P15) animals, two groups of MafB-V1 interneurons (GFP + tdtomato, tdT) can be identified: one occupies the dorsal region of the V1 cell distribution; the other is located ventrally, is calbindin-immunoreactive (-IR) and corresponds to Renshaw cells (Figure 3A). These two V1 cell groups are also identifiable at P5, but at this age weak GFP is also visible in other V1 cells located in the middle of the ventral horn and that likely belong to the Foxp2-V1 clade. Their numbers vary from animal to animal. This suggests mafB gene expression is in the process of downregulation in these cells and explains the high interanimal variability at P5 and the absence of expression at P15. We focused our analyses in dorsal and ventral MafB-V1 populations defined by boxes of 100 µm width at the level of the central canal (dorsal) or the ventral edge of the gray matter (ventral). Together, MafB-V1s in these two regions constitute 13.2% ±2.8 (mean±SD) of all V1s at P5 (n=17 mice) with similar representation in dorsal (7.1% ±2.4) and ventral (6.0% ±1.2) groups (Figure 3B). These percentages did not change at P15. At P5, on average, 54.2% ±19.4 of dorsal and 70.5% ±10.1 of ventral MafB-V1s had detectable MafB-IR (Figure 3C, n = 9 mice tested). No MafB protein was detected at P15. We tested V1 clade-immunomarkers in ventral and dorsal MafB-V1 cells at P5 (calbindin for Renshaw cells, and Pou6f2, Foxp2 and Sp8 for other V1 clades) (Figure 3D). Dorsal MafB-V1s expressed Pou6f2 (41.1% ±7.4, n=5 mice tested), negligible calbindin (1.5% ±1.8, n=4 mice) and no Foxp2 or Sp8 (0%, n=5 mice). Most ventral MafB-V1s expressed calbindin-IR (79.3% ±8.8, n=4 mice) while expression of other clade markers was negligible (Pou6f2, 0%; Foxp2, 0.4% ±0.9; Sp8. 0.7% ±1.5, n=5 mice) (Figure 1E). Using this genetic model, we confirmed birth dates of dorsal (Pou6f2) and ventral (Renshaw) V1s (Figure 3F, n=2 mice per EdU injection date) and found an excellent match in birthdates of dorsal MafB-V1s identified by either MafB-IR or MafB-GFP. However, we found more ventral MafB-V1s born at E11 using genetic labeling compared to antibody staining, although this difference was not statistically significant. In conclusion, we confirmed both MafB-V1 populations and their early birth dates using a genetic model. These two classes constitute discreet V1 populations. The functional significance of adult dorsal Pou6f2-MafB-V1s is yet unexplored.

Genetic labeling of the Foxp2-V1 lineage reveals twice the number of neurons relative to postnatal Foxp2 expression

To label the Foxp2-V1 interneuron lineage independent of developmental regulation of Foxp2 expression, we utilized an en1 and foxp2 intersection model by generating en1cre/+::foxp2flpo/+::R26RCE-dualGFP/Ai9-tdTmice. In these animals we expected that neurons expressing en1 and not foxp2 would be labeled with tdT from the R26-Ai9 cre reporter, and that neurons that expressed both en1 and foxp2 would be labeled with GFP from the R26-dual (Cre&Flpo) RCE:GFP and not with tdT because the Ai9 cassette is flanked by FRT sites that can be removed by Flpo. We confirmed either tdT or GFP labeling in most V1 interneurons, but a few expressed both fluorescent proteins (Figure 4A). We believe this is the consequence of Flpo recombination inefficiency in the larger FRET-flanked Ai9 cassette compared to the much shorter FRET-stop signal in the RCE:dualGFP. We thus interpret “yellow” V1 neurons as cells that express the foxp2 gene either for a short period of time, weakly, or both. For mapping purposes, we included “yellow” cells into the Foxp2-V1 clade since we used only the RCE:dualGFP reporter in many follow-up analyses. The distribution of Foxp2-V1s (green) and non-Foxp2 V1s (red, not green) overlapped in the ventral horn (Figure 4B). Their proportions were constant across postnatal ages (n=6 ventral horns from 1 mouse at P0, P15 and adult and 36 ventral horns from 6 mice at P5; Figure 4C) and across spinal segments from lower thoracic to sacral level at P15 analyzed in 2 mice (n=6 ventral horns per segment and mouse, Figure 4D). Respectively in each mouse, 51 and 52% of V1 cells were EGFP labeled on average, 34% and 39% were tdT labeled and 8% and 14% had both labels. The whole population of Foxp2-V1 interneurons (green+yellow) cells represents between 59% to 66% of all V1s. However, the data presented before suggest that only between 32% to 34% of V1s express Foxp2-IR at P0 and P5. This indicates that Foxp2 expression is downregulated in around half of lineage-labeled Foxp2-V1 cells before P5. Correspondingly, around half (48.9% ±3.8; n=3 mice) of genetically identified Foxp2-V1 cells were Foxp2 immunolabeled at P5 (Figure 4E-G). The locations of Foxp2-V1 cells with and without Foxp2 immunoreactivity at P5 overlapped in the ventral horn (Figure 4F).

Genetic mouse models used to label the Foxp2-V1 lineage.

A, P5 mouse intersection of En1::Cre and Foxp2::Flpo with two reporter alleles (Ai9-tdTomato and RCE:dualEGFP) in the Rosa26 locus. Foxp2-V1s express GFP (green) and non-Foxp2-V1s tdTomato (red). A few “yellow cells” correspond with Foxp2-V1 neurons that failed to remove the tdTomato Ai9 reporter. B, Density contours demonstrate high spatial overlap between V1 neurons expressing Foxp2 (EGFP+) or not (tdT+ and EGFP-) (n=6 ventral horns). C, Expression of the lineage label is stable throughout postnatal development. This means that there is no additional foxp2 gene upregulation in new V1s after P0 (n=6 ventral horns per age in one animal, except for P5 in which 6 animals and 36 ventral horns are included; error bars show SEM). D, In P15 mice, expression of the lineage labels is also uniform across the lower spinal cord segments from thoracic 13 to sacral 1 (n = 6 ventral horns in each segment from 2 mice; error bars show SEM). E, The En1::Cre, Foxp2::Flpo, RCE:dualEGFP model detects more Foxp2-V1s in the postnatal spinal cord than Foxp2 antibody staining at P5. F, Foxp2-V1s with or without postnatal expression of Foxp2 occupy roughly similar areas in the P5 spinal cord (n=3 mice, 3 ventral horns per animal). G, Right, Foxp2-V1s make up roughly 60% of all V1s when measured with genetic labeling, but only about 30% of V1s maintain detectable Foxp2 expression at P5. Left, around half of Foxp2-V1s maintain detectable levels of Foxp2 at P5 (n is same as F; error bars show SD). H, FLTG reporter mice reveal lineage-labeled non-V1 Foxp2 cells in the spinal cord. eGFP is expressed in Foxp2-V1s and tdTomato in non-V1 Foxp2 cells. The zoomed images of the highlighted region with and without NeuN-IR demonstrate that non-V1 Foxp2 cells include non-neuronal cell types with astrocyte morphologies. Only neurons (NeuN-IR) are included in the contour plots which show non-V1 Foxp2 neurons located in the medial ventral horn and a few in the deep dorsal horn. (n = 2 animals at P5, 6 ventral horns in each). I, Most Foxp2-neurons (red, non-V1s or green, V1s) are in the ventral horn and their number in 30 μm thick L4-5 sections decreases with age as the neuropil matures and expands in size. Dorsal horn Foxp2-neurons maintain their numbers despite the growth of the spinal cord. Each point is one animal analyzed through 6 ventral horns (errors bars are SD). J, The percentage of Foxp2-lineage-labeled cells that are Foxp2-V1s remains constant throughout postnatal development (n=2 mice, 6 ventral horns each; error bars show SD).

In summary, lineage labeling of cells that express Foxp2 at any time during development result in doubling the size of the Foxp2-V1 clade compared to estimates based on Foxp2 expression at P0 (Bikoff et al., 2016). Percentages remain constant throughout postnatal development and into adulthood suggesting no new expression of Foxp2 in V1 cells after birth.

Foxp2-V1 neurons are only half of all Foxp2 expressing spinal neurons but non-V1 Foxp2 cells distribute to very different regions

Foxp2 protein is found postnatally in many non-V1 cells (Figures 1F and 4E). To study these populations, we used RC::FLTG reporter mice. These mice carry a dual-conditional allele with a FRT-flanked stop and loxP-flanked tdT::STOP preventing transcription of eGFP. Therefore in en1Cre::foxp2Flpomice, cells that express only foxp2 are labeled with tdT, while additional en1 expression (Foxp2-V1 clade) results in eGFP fluorescence and removal of tdT. We analyzed two mice at P5 and two at P10, and the sections were counterstained with NeuN antibodies for neuronal confirmation (Figure 4H). We found several non-overlapping populations of Foxp2 cells. Most non-V1 Foxp2 cells are located medially in the ventral horn and can either be neurons or astrocytes (by morphology and lack of NeuN). In addition, glial cells and cells in the central canal and spinal cord midline were strongly labeled. This distribution suggests Foxp2 is transiently expressed in some progenitors, other than p1 (no eGFP astrocytes are observed in En1/Foxp2 derivatives). A few other non-V1 cells correspond to deep dorsal horn neurons (NeuN+). The number of Foxp2 neurons per section diminished in the ventral horn from P5 to P20, as expected, because of the reduction in cellular density as the spinal cord matures and grows. This decrease did not occur among dorsal horn cells (Figure 4I), which suggests some dorsal horn neurons might upregulate Foxp2 postnatally.

In summary, there are at least three broad different types of Foxp2 neurons in the spinal cord: 1) medio-ventral non-V1 neurons that express Foxp2 postnatally and/or at the progenitor stage, sharing labeling with glial cells; 2) dorsal horn non-V1 neurons in which Foxp2 expression increases during postnatal development; 3) V1 neurons that upregulate Foxp2 in embryo and then remain a stable population postnatally (see also (Benito-Gonzalez and Alvarez, 2012). Overall, Foxp2-V1 interneurons comprise 55.7% ±4.7 (mean±SD) of all genetically labeled Foxp2 neurons in the spinal cord (n=4 mice, 2 P5 and 2 P20, Figure 4J).

Foxp2-V1 neurons from lower thoracic to sacral segments follow motoneuron numbers in a 2:1 or 3:1 ratio

To obtain insights into the possible function of Foxp2-V1 interneurons we analyzed possible shifts in localization and numbers according to spinal cord segment in the lumbosacral region concerning control of lower body and hindlimbs. We examined spinal cord sections from segments Th13 to S1 in P20 mice (n=5) expressing eGFP in Foxp2-V1 interneurons combined with choline acetyltransferase (ChAT) immunoreactivity to identify the motor pools. Motoneurons were defined as any Chat-IR neuron in lamina IX (Figure 5A). Spinal segments were identified by the characteristic distribution and size of the somatic lateral, hypaxial and medial motor columns (LMC, HMC and MMC), and by the presence of autonomic sympathetic (Th13-L2) or parasympathetic (S1) neurons. We did not attempt to distinguish Lumbar 4 from 5 because of their similarity. From these sections we constructed cell plots for each animal (4 ventral horns per animal/segment, Figure 5B) and transformed these into density plots (Figure 5C) by combining all cell plots from all animals analyzed in each segment (n=3-5 mice depending on segment). Foxp2-V1 interneurons are located throughout the ventral horn but accumulate laterally. In segments in which the LMC expands, Foxp2-V1 interneurons border medially the LMC. Contour density plots indicate that the highest density of Foxp2-V1 interneurons lies adjacent to the LMC, suggesting a close relationship between Foxp2-V1 interneurons and the control of limb musculature. This is consistent in segments where the LMC emerges (L2), disappears (L6) or where the LMC reaches its maximal size (L4/5). Correspondingly, Foxp2-V1 neuron numbers significantly increase in segments innervating limb muscles (L3-L6) compared to segments involved with axial (Th13, S1) and hypaxial muscles (L1) (p<0.0001, one-way ANOVA followed by post-hoc Bonferroni t-tests summarized in Figure 4D and Supplement Table S3). The limb-innervating LMC is responsible for most of the change in motoneuron numbers across different spinal cord segments. Consistent with V1 interneurons and motoneurons increasing in number in parallel, the ratio of Foxp2-V1 interneurons per motoneuron remained relatively constant from Th13 to L5 (2:1 to 3:1 ratio) (Figure 5E). Differences between Th13 and L6 are non-significant (post-hoc Bonferroni t-tests, details in Supplemental Table S3). The larger ratio at S1 was significant compared to L3 and L4/5 (see Supplemental Table S3 for details). Estimated ratios at S1 were highly variable in the three animals studied likely because sampling issues: the number of motoneurons quickly diminishes in S1 depending on exact section level. Examination of cell plots in segments lacking LMC limb motoneurons show that most Foxp2-V1 interneurons are located dorsally and distal to the motor pools with relatively lower density close to MMC motoneurons. Foxp2-V1 interneurons located further away from the motor pools might have roles other that direct modulation of motoneuron firing. Finally, a sparse group of Foxp2-V1 interneurons is dispersed in the medial ventral horn in all segments with the best representation from L2 to L6. They correspond to the latest born subgroup (and they have a different genetic make-up, see below), suggesting a unique identity.

The spatial distribution of Foxp2-V1 interneurons is closely associated with shifting motor columns throughout thoracic, lumbar and sacral levels of the spinal cord.

A, Foxp2-V1 lineage labeling and ChAT antibody staining for motoneuron identification in adult mouse spinal cords from thoracic to sacral levels. Foxp2-V1 neurons accumulate at the lateral edge of the ventral spinal cord but their locations shift as the LMC expands from L3 to L6. In addition, there is a distinct group of Foxp2-V1 interneurons dispersed at medial locations in all lumbar segments. B, Plots of Foxp2-V1 and ChAT-IR motoneuron cell body positions in x,y coordinates with 0,0 at the top of the central canal (n=4 ventral horns, 1 representative animal). C, Contour plots of kernel Foxp2-V1 cell density estimations. The highest density of Foxp2-V1 neurons cluster close to LMC motoneurons from L2 to L5 (contours enclose 10% increments, closer lines indicate higher density). Motoneuron numbers progressively increase from Th13 to L5 before dropping in number. D, The number of Foxp2-V1s per 50 μm thick section (ventral horn) significantly increases in lower lumbar segments from L3 to L6 compared to S1 (dots represent individual mice; n = 3-5 mice in different segments, each mouse estimate is from 6 ventral horns; bars show SEM; one-way ANOVA, post-hoc Bonferroni corrected t-tests: summary statistical results are in the right-hand table. See Table S3 for more details). E, The ratios of Foxp2-V1 neurons to MNs remain relatively constant at roughly 2.5:1 with no significant changes throughout the lumbar cord. Significance was only found for L3-L5 compared to S1 (*p<0.05; **p<0.01; post-hoc Bonferroni tests). High variability in S1 is likely due to the sharp rostro-caudal decrease in motoneuron numbers in S1 sections and possible sampling differences among animals.

Foxp2-V1 interneurons project to LMC, HMC, and MMC motoneurons, but not to autonomic motoneurons

To gain further insight into the functional roles of Foxp2-V1 interneurons. we examined their synapses on different types of motoneurons and compared them to other V1 groups at P20: a time point after critical windows of synapse proliferation and pruning on V1 neurons (Mentis et al., 2006; Siembab et al., 2010). We examined mice in which Foxp2-V1 axons contain eGFP, and non-Foxp2-V1 axons express tdT (n=3) (Figure 4A). Yellow axons were included in the eGFP/Foxp2+ group. We also analyzed mice in which all V1 axons express tdT through the Ai9 reporter (n=2). In these animals we immunostained axons with calbindin antibodies to identify synapses from Renshaw cells. Sections were further immunostained using antibodies against ChAT (to identify motoneuron cell bodies in lamina IX) and with antibodies against the vesicular GABA/glycine amino acid transporter (VGAT) to reveal synaptic vesicle accumulations in genetically labeled axon varicosities contacting ChAT motoneurons. Motoneurons were sampled in different motor columns from Th13 to S1 segments (Figures 6A-B) and examined at high magnification for synaptic contacts from Foxp2-V1, non-Foxp2-V1, all V1 and Renshaw-V1 axons (Figures 6C1 and 6C2). We estimated synaptic contact densities on 3D reconstructions of cell body surfaces using rigorous methodology (Figure 6C3). We calculated overall V1 synaptic densities (green and red axons in eGFP/tdT dual color mice and all red axons in tdT single color mice) on motoneurons located in motor columns and segments. This included HMC motoneurons in the Th13 segment (ventral body musculature); LMC motoneurons in segments L1/2 (hip flexors), L4/5 (divided into dorsal and ventral pools, innervating distal and proximal leg muscles, respectively) and the dorsal L6 pool (intrinsic foot muscles); MMC neurons in segments Th13, L1/2, L3/4 and S1 (innervating axial trunk musculature and the tail at sacral levels) and finally preganglionic autonomic cells (PGC, sympathetic at Th13 and L1/2 and parasympathetic at S1). We analyzed 4 to 9 motoneurons per animal. Initially, we kept the data separated by mouse identity to check for possible differences due to mouse and/or genetics (Figure 6D). A mixed-effects nested ANOVA revealed significant differences in V1 synapse density over different types of motoneurons (p<0.0001), and no influence of mouse or genetics (statistics details in supplemental table S4 and Figure 6D table). Post-hoc Bonferroni t-tests demonstrated that HMC and lower lumbar LMC motoneurons receive significantly more V1 synapses than MMC motoneurons, while LMC motoneurons in L1/2 and L6 had V1 synaptic densities not significantly different to MMC motoneurons. PGC neurons received very low densities of V1 input, significantly lower than LMC or MMC motoneurons.

Limb and axial motoneurons are densely innervated by Foxp2-V1s and Renshaw cells.

A, Motor column identifications in P20 mice following labeling with ChAT antibodies from lower thoracic to upper sacral spinal cord: PGC = preganglionic cell column; MMC = medial motor column; HMC = hypaxial motor column; LMCd/v = lateral motor column (dorsal/ventral). B, Schematic depiction of the rostro-caudal span of each motor column in the spinal cord segments studied. C, Synapse quantification. Axons of Foxp2 and non-Foxp2 V1 interneurons were respectively labeled with eGFP and tdT in en1Cre::foxp2flpo::R26 RCE:dualGFP/Ai9tdT mice. In en1Cre::R26 Ai9tdT mice we identified V1-Renshaw cell axons using calbindin antibodies. Synaptic locations were labeled with VGAT antibodies and the postsynaptic motoneurons with ChAT antibodies. Synapse densities were analyzed in a ribbon of membrane at mid-cell body level (7 optical planes, 1 µm z-step). C1, Single optical plane of a L4/5 LMCv motoneuron surrounded by genetically labeled Foxp2-V1 and non-Foxp2-V1 axons. Inhibitory synapses on ChAT-IR motoneurons are VGAT+. C2, Single optical image of a L4/5 LMCv ChAT-IR motoneuron receiving synapses from putative V1 Renshaw cells (genetically labeled V1 axons with calbindin-IR and VGAT). C3-4, Method for estimating synapse densities on motoneuron cell bodies using C2 as example. C3, V1-VGAT (red arrowheads) and V1-CB-VGAT synapses (yellow arrowhead) are marked (VGAT-IR is not shown for clarity), and the cell body contour annotated with regions corresponding to dendrite exits. This process was repeated in 7 consecutive mid-cell body optical planes (cross-sections with well-defined nucleus and nucleolus). C4, A membrane surface slab is reconstructed in 3D (two different rotations shown). The surface area corresponding to dendrite exits is subtracted from the total surface area of the slab to calculate the available surface area on the motoneuron cell body. V1-VGAT synapses (red), V1-CB-VGAT synapses (yellow), and CB-VGAT synapses (green) are marked. A similar process was followed for calculating Foxp2-V1 synapse density in dual labeled animals. D, Quantification of total V1-VGAT synapse densities on motoneuron cell bodies in different motor columns (n = 21-30 motoneurons per motor column, taken from 5 animals with 4-9 motoneurons per animal per motor column). Each data point is one motoneuron color coded by mouse origin. Average synaptic densities ±SD indicated to the right of scatter plots. A nested ANOVA indicated significant differences among motor column/segments (p < 0.0001) with no inter-animal variability (p = 0.4768). The table summarizes all post-hoc pairwise comparisons for average V1 synaptic densities of each motor column and segment (Bonferroni corrected t-tests) (Statistical details are in Supplementary Table S4). Colors indicate increased (>1, red) or decreased ratios (<1, blue) of column motoneurons vs row motoneurons. PGC neurons receive significantly fewer V1 synapses than MMC or LMC motoneurons. The LMC (ventral and dorsal) in lower lumbar (L4/L5) had significantly more V1 contacts than MMC motoneurons or L6 dorsal LMC. E, Comparison of synaptic densities from Foxp2-V1 and non-Foxp2-V1 neurons (top) or Renshaw cells (bottom). All motoneurons sampled in 2 to 3 animals for each comparison were pooled together. Densities of V1-VGAT synapses from Foxp2-V1s, non-Foxp2 V1s, or calbindin (CB)+ V1s (Renshaw cells) (n = 6-17 motoneurons sampled per motor column/segments, average = 12.1 ±2.9 SD) were compared using a two-way ANOVA for axon type vs motor column and segment. Foxp2-V1 vs non-Foxp2-V1 synapses: significant differences in density were found for type of synapse (p=0.001), motor column location (p<0.0001), and their interaction (p<0.0001). Significant differences after post-hoc Bonferroni tests are indicated (*p<0.05; ****p<0.0001). In general, synapses from Foxp2-V1 axons have higher density than non-Foxp2-V1 axons on HMC and LMC columns at all spinal segments except for L1/L2 LMC. MMC motoneurons receive similar synaptic densities from both types of V1 axons, except at the sacral level in which non-Foxp2 V1 synapses predominate. PGC neurons receive very low densities of V1 axons and there are no significant differences between either type in any region. Foxp2-V1 vs CB+V1 synapses: significant density differences were found for type of synapse (p<0.0001), motor column location (p<0.0001), and their interaction (p<0.0001). Significant differences between Foxp2-V1 and CB+V1 synapses after post-hoc Bonferroni tests are indicated (*p<0.05; ****p<0.0001). Synapses from Foxp2-V1 axons have higher density than CB+V1 axons in HMC and LMC columns at all spinal segments except for L1/L2 LMC. MMC motoneurons receive similar synaptic densities from both types of V1 axons in upper lumbar regions, but Foxp2-V1 synapse predominate in lower lumbar. In S1 the density of CB+/V1 synapses is significantly higher. The low synaptic densities estimated in PGC neurons for Foxp2-V1s and CB+ V1s are not significantly different. Details of all statistical comparisons are Supplementary tables S5 and S6. F, Comparing the numbers of Foxp2 and CB+ (Renshaw) V1 synapses to the total number of V1 synapses, we estimated their respective percentages. From these estimates we calculated that the remainder belongs to non-Foxp2 and non-CB+ Renshaw cells. The large majority of V1 synapses on the cell bodies of LMC, HMC and MMC motoneurons are either from Renshaw cells or Foxp2-V1s. Asterisks denote significant differences as found in E.

Next, we examined possible differences between Foxp2-V1 and non-Foxp2-V1 neurons (Figure 6E, top graph). In this case we pooled all motoneurons from 3 mice (n=6 to 16 motoneurons per motor column/segment). We found significant differences according to motoneuron identity (p<0.0001), type of axon (p=0.0107), and their interaction (p<0.0001) (two-way ANOVA, statistics details in supplemental Table S5). This was followed by pair-wise comparisons of synaptic densities according to the type of V1 axon for each motoneuron type (Figure 6E, top graph). For HMC and LMC motoneurons we found significantly higher synaptic densities of Foxp2-V1 axons compared to non-Foxp2-V1 axons in all segments, except for L1/2. The synaptic densities of both types of axons are not significantly different in MMC motoneurons, except for S1 MMC motoneurons that received a significantly higher density of non-Foxp2-V1 synapses. Synapses on PGC neurons were at low density and were highly variable with no differences found between either type of V1 axon. We conclude that motoneurons controlling the hindlimb receive more synapses from Foxp2-V1 interneurons. Foxp2-V1 and non-Foxp2-V1 interneurons equally contact the cell bodies of motoneurons controlling axial musculature.

Finally, we compared synapse densities from Foxp2-V1 interneurons to those originating from Renshaw cells, another V1 interneuron that preferentially targets the proximal region of motoneurons. Renshaw cell axons were identified by the presence of calbindin-IR in tdT V1 axons (in this case we pooled 10 to 17 motoneurons from two mice). Like above, we found significant differences according to motoneuron identity (p<0.0001), type of axon (p<0.0001) and their interaction (p<0.0001) (two-way ANOVA, statistics details in supplemental Table S6). Post-hoc pairwise comparisons revealed that Renshaw cell synapses occurred at significantly lower densities compared to Foxp2-V1s in all LMC motor groups, except for L1/2 (Figure 6E, lower graph). MMC motoneurons showed similar densities of Renshaw cell and Foxp2-V1 synapses in Th13 and L1/2, higher density of Foxp2-V1 synapses in L3/4, and much higher density of calbindin+V1 synapses in S1. The identity of calbindin+ V1 axons in S1 is unclear because of the relatively higher proportions of calbindin-IR V1 interneurons in sacral segments. Further confirmation of Renshaw cell identity for calbindin+ V1s in S1 is required. PGC neurons received almost no calbindin+ V1 axons (Figure 6E, bottom graph). After calculating the synaptic densities originating from Foxp2-V1 and Renshaw cell axons, we estimated the remaining V1 synapses as non-Foxp2 and non-Renshaw V1. Plotting these three categories of synapses by their percent contributions to all V1 synapses shows that the majority of synapses on the cell bodies of HMC, LMC and MMC motoneurons originate from either Foxp2-V1s or Renshaw cells (Figure 6E). PGC neurons mostly receive inputs from V1s that are non-Foxp2 and non-Renshaw cells, but these synapses are of very low density and high variability. The data suggest that the Foxp2-V1 clade is a major source of inhibitory inputs to motoneuron cell bodies where they likely strongly modulate motoneuron firing. However, given the very large difference in cell numbers between Foxp2-V1 interneurons and Renshaw cells, the data also suggest that the amount of divergence in the Renshaw cell output is likely very high. Foxp2-V1 interneurons preferentially target limb motoneurons, consistent with conclusions from cell density plot analyses in Figure 5. There is a correlation between the densities of Foxp2-V1 synapses and their cell density in proximity to different motor columns and segments (see Figure 5). This suggests that Foxp2-V1 synapses on motoneurons likely originate from V1 interneurons that are clustered spatially close to the LMC. Motoneurons in spinal segments with Foxp2-V1 interneurons located further dorsally from motor pools (e.g., S1) receive a relatively lower density of synapses from Foxp2-V1 interneurons. Pou6f2, Sp8 and other possible V1 clades either do not target motoneurons directly, or they target them sparsely or on distal dendrites. This suggests functional differences among V1 clades regarding the strength of direct modulation of motoneuron firing through proximal synapses.

Foxp2-V1 interneurons clustered near LMC motoneurons are genetically distinct

Birthdate, spatial organization, and synapse densities on different motor columns all suggest that Foxp2-V1 interneurons are heterogeneous and that a laterally-located group close to LMC motor pools might modulate the output of limb motoneurons. To identify potential genetic differences among Foxp2-V1 interneurons, we examined the published TF expression profiles of V1 interneurons (Bikoff et al., 2016), and selected two TFs highly enriched in Foxp2-V1 interneurons for further study: Orthopedia homeobox (Otp) and Foxp4. We used antibodies to reveal V1 interneurons expressing Otp and/or Foxp4 in the P5 spinal cord (all analyses were performed in Lumbar 4 and 5 segments). We first used two animals with Foxp2 and non-Foxp2 V1 interneurons labeled respectively with eGFP and tdT. Both Otp and Foxp4 are almost exclusively expressed by Foxp2-V1 interneurons at P5 with negligible expression in non-Foxp2 V1 interneurons (Figure 7A). Otp was expressed in around 50% of lineage-labeled Foxp2-V1 interneurons and Foxp4 was expressed in around 20%. Foxp4-IR / Foxp2-V1 interneurons always co-localized with Otp-IR, indicating they are a subpopulation of the Otp group. To examine the relationship between these groups with V1 cells that retain Foxp2 expression at P5, we generated different combinations of paired immunolabelings for Otp, Foxp4 and Foxp2 (Figure 7B) in 3 mice with eGFP expression in Foxp2-V1 interneurons and liberating the red channel for TF co-localization. We constructed cell density contours and calculated the percentages of lineage-labeled Foxp2-V1 cells (eGFP) expressing different combinations of Otp, Foxp4 and Foxp2 (Figure 7C). To summarize the most salient results, a large group of Foxp2-V1 cells that co-expresses Otp and Foxp2 at P5 (44% of Foxp2-V1 cells) is localized laterally close to the LMC. It includes a smaller subgroup that also expresses Foxp4 (23% of Foxp2-V1 cells) and that is located more ventrally. Many ventromedial Foxp2-V1 cells express Foxp2, but never Otp or Foxp4 at P5, and they occupy the locations of very late-born Foxp2-V1 cells (Figure 2). A second group of cells expressing only Foxp2 is located more dorsally.

Subgroups of Foxp2-V1 interneurons defined by transcription factor expression at P5 and birthdate.

A, Otp (blue) and Foxp4 (white) expression in lineage-labeled Foxp2-V1s (eGFP, green) and non-Foxp2-V1s (tdT, red). The boxed area is shown at higher magnification with different color combinations for clarity. It shows that Foxp4-IR cells in the Foxp2-V1 population always also expressed Otp (arrows). Quantification: 49.8-51.2% of Foxp2-V1s express Otp and 20.7-21.7% Foxp4 (n=2 mice each examined in 3 ventral horns in L4/5). Very few non-Foxp2-V1 cells express either TF (Otp: 1.3-4.2% and Foxp2: 0.5-2.9%). B, Images of P5 spinal cords containing Foxp2-V1 lineage labeling (eGFP, omitted for clarity) and double immunolabeled for Otp/Foxp2, Foxp4/Foxp2 and Otp/Foxp4. C, Quantification of Foxp2-V1 interneurons with different combinations of TF expression at P5. For each combination, the left panels show cell distributions and the right graphs show the percentage of Foxp2-V1s with each combination (n = 3 mice each analyzed in 6 ventral horns). To discern groups with significantly different numbers of cells, the data was analyzed with one-way ANOVAS followed by Bonferroni corrected pair-wise comparisons (Statistical details in Supplemental Tables S7, S8 and S9. ****p<0.0001; ***p<0.001; **p<0.01; *p<0.05). There are five main groups defined by TF expression patterns: Otp-Foxp2 (44%), Otp-Foxp2-Foxp4 (23%), Otp only (15%), Foxp2 only (9%) and no TF labeling (32% by subtracting all other groups). Some groups associate with specific locations: Otp-Foxp2 cells and the smaller proportion of Otp only cells are located laterally; Otp-Foxp4-Foxp2 cells are located latero-ventrally; medial cells either contain only Foxp2 or nothing; some dorsal cells are either Foxp2 only or do not express any of these TFs. D, Example image of EdU birthdating in en1Cre::foxp2flpo::R26 RCE:dualGFP/Ai9tdT dual-color mice pulse labeled at E11. Foxp2-V1s (green and yellow cells) are born between E10 and E12, with peak birthdate around E11 and after non-Foxp2 V1s (tdTomato only). The lower graph’s data is normalized to highlight the time of peak neurogenesis for each population (n=15.0 ± 4.1 ventral horns from one mouse per time point from E10 to E11 and 2 mice per time from E11.5 to E12.5; total 9 mice, bars show SEM). There is no difference between green (eGFP only) and the smaller population of yellow cells (eGFP and tdTomato). E, EdU labeling in the Foxp2-V1 dual-color genetic model combined with Otp and Foxp4 antibody staining. F, V1s expressing Foxp2, Otp, and Foxp4 at P5 are mostly born between E10.5 and E11.5 with neurogenesis time courses largely overlapping. The normalized plot indicates that peak neurogenesis for all three populations occurs at E11.5, although a marginally higher number of Otp V1 cells are born earlier (n=4 ventral horns from 3.27 ±1.34 mice per timepoint; error bars show SEM). Contour plots to the right show settling locations of Otp and Foxp4-IR populations born at each time point. G, Schematic of the L4-5 ventral horn summarizing Foxp2-V1 subgroups according to location and combinatorial expression of Otp, Foxp4, and Foxp2 at P5. LMCD: dorsal lateral motor column; LMCV: ventral lateral motor column; MMC: medial motor column.

Next, we tested whether Otp and Foxp4 expressing populations match the dorsal to ventral progression of birthdates observed in laterally-located Foxp2-IR V1s (Figure 2). We first investigated possible differences in the birthdate time course of Foxp2-V1 cells labeled genetically (Figure 7D) in comparison with V1 cells identified by Foxp2 expression at P5 (Figure 1G-H and 7F). The time course of neurogenesis was similar, but a few differences were also noted. Like Foxp2-IR V1 cells, most cells in the genetically labeled Foxp2-V1 lineage were born between E10.5 and E12.0, with few born at E10.0 or E12.5 (n=1-2 mice per time point). The peak of neurogenesis for the lineage-labeled population occurred at E11.0; that is 12 hr. earlier than the peak of neurogenesis for V1 cells expressing Foxp2 at P5. This can be explained by a proportion of the earliest born cells in the Foxp2-V1 clade downregulating Foxp2 expression by P5. Next, we used Otp and Foxp4 to subdivide lineage-labeled lateral Foxp2-V1 cells. To increase sample size, we pooled spinal cords from mice in which we genetically labeled both Foxp2-V1s (eGFP) and non-Foxp2-V1s (tdT) (Figure 7E) together with mice having all V1s lineage-labeled with tdT (n = 4 at E10, 5 at E10.5, 4 at E11, 3 at E11.5 and 2 at E12.5). Pooling data from both genetic models is justified by the above results showing that V1 cells expressing Otp or Foxp4 at P5 are all contained within the Foxp2-V1 lineage. The neurogenesis curves of V1 cells expressing Otp, Foxp4 and Foxp2 at P5 largely overlapped. All three groups had a peak at E11.5 with almost no cells generated at E10.0 or E12.0. The locations of Otp and Foxp4 expressing V1s generated between E10.5 and E11.5 were lateral for Otp and ventral for Foxp4. The location of Foxp4 cells generated at E10.5, E11.0 and E11.5 did not change significantly, but the location of Otp cells generated at E11.5 shifted ventrally compared to those born at E10.5 or E11.0 (Figure 7F). We interpret this result as suggesting that generation of Foxp4-Otp Foxp2-V1 interneurons overlaps with the generation of other Otp cells in the same lineage, but also outweighs other groups towards the end of Otp Foxp2-V1 neurogenesis.

The results can be summarized by proposing at least four groups of Foxp2-V1 cells according to location, TF expression, and birthdate (Figure 7G). Group I is located dorsally, lacks Otp, and has variable expression of Foxp2 at P5. Many are likely generated during early Foxp2-V1 cell neurogenesis (before E10.5). Groups II and III are laterally located and express Otp with or without Foxp2 (group II) or express Otp, Foxp4 and Foxp2 together (group III). These two groups together represent the largest class of Foxp2-V1 cells in L4/5 and are generated during a 24-hr. period from E10.5 to E11.5 with a slight shift in the balance of group II vs III cells at later times of neurogenesis. Finally, medial Foxp2-V1 cells (group IV) are generated very late (after E12: see Figure 2) and lack expression of Otp or Foxp4, but many retain expression of Foxp2 at P5. We failed to identify in this medial Foxp2-V1 group expression of markers for late-born spinal interneurons located in the medial ventral horn, like NeuroD2 and Prox1 (Delile et al., 2019; Osseward et al., 2021), despite presence of many neurons positive for these TFs in the vicinity.

Otp-expressing Foxp2-V1 cells receive proprioceptive synapses

Previous studies reported that Foxp2-V1 interneurons include a large class of proprioceptive interneurons, some of which could represent reciprocal Ia inhibitory interneurons (IaINs) because they receive convergent synapses from excitatory proprioceptive sensory axons (VGLUT1+) and inhibitory Renshaw cells (calbindin+ V1 axons) (Benito-Gonzalez and Alvarez, 2012). Likely candidates are Otp Foxp2-V1 interneurons in groups II and III because their localization matches that of electrophysiologically identified IaINs in the cat (Jankowska and Lindstrom, 1972; Alvarez et al., 1997). To examine this question, we first analyzed the types of Foxp2-V1 interneurons receiving proprioceptive inputs. The spinal cords of two mice at P5 (to preserve TF expression) containing Foxp2-V1 lineage-labeled cells (eGFP) were dual or triple immunolabeled for VGLUT1 and Otp and/or Foxp2 (Figure 8A). We analyzed 6 ventral horns at L4/L5 in each animal using high magnification confocal microscopy (60x), tiling the whole ventral region containing all Foxp2-V1 cells. Cells were annotated as receiving no synapses (I in Figure 8A) or low/medium and high density of VGLUT1 synapses (respectively, II and III in Figure 8A). We also noted whether these synapses were located proximally (on cell body and primary dendrites) or more distally (VGLUT1 synapses far from cell bodies on higher order dendrites). In general, cells with proximal VGLUT1+ synapses had higher densities than those with only distal synapses. Overall, we found that 63.0% and 74.3% of Foxp2-V1 interneurons received VGLUT1 synapses in each animal (Figure 8B). Foxp2-V1 cells with no VGLUT1 synapses were found throughout the ventral horn, but those receiving VGLUT1 synapses had a lateral positioning bias (Figure 8C).

Many Foxp2-V1 interneurons receive proprioceptive inputs, and some are reciprocal Ia inhibitory interneurons (IaINs).

A, Left, low magnification confocal image of Foxp2-V1 lineage-labeling (eGFP) combined with VGLUT1 antibody staining to identify primary proprioceptive afferent synapses. The boxed area is shown at high magnification (right) to demonstrate variability of VGLUT1 synapse density on Foxp2-V1 interneurons (I = absent, II = medium or low, III = high). For simplicity and rigor, we classified Foxp2-V1 interneurons as receiving or not receiving VGLUT1 synapses. B, Percentage of Foxp2-V1 interneurons with VGLUT1 synapses (both proximal and distal). Each dot is one animal estimate from 6 or 7 ventral horns with respectively 591 and 525 Foxp2-V1 interneurons sampled. Line represents the average of both animals, 69.6%. C, Distribution of Foxp2-V1 interneurons with and without VGLUT1 synapses (green and blue, respectively). The positioning of Foxp2-V1 interneurons receiving VGLUT1 synapses is biased laterally. D, Foxp2-V1 interneurons with (green) and without (blue) VGLUT1 synapses according to expression of Otp (left graph) or Foxp2 (right graph). Each dot represents one mouse, and the lines represent averages. The numbers of sections and genetically labeled Foxp2-V1 interneurons sampled in each mouse are as in B. In each mouse, this includes 237 and 236 Otp-IR cells and 256 and 225 Foxp2-IR cells. E, Experimental design to label spinal neurons that receive inputs from tibialis anterior (TA) muscle primary afferents and also connect to lateral gastrocnemius (LG) motoneurons, forming Ia reciprocal inhibitory connections from TA to LG. TA sensory afferents are labeled anterogradely with CTB followed by antibody detection of CTB and the presynaptic marker VGLUT1. Interneurons premotor to LG motoneurons are labeled by monosynaptic retrograde labeling with RVΔG-mCherry. F, Foxp2-V1 IaIN with the most TA/VGLUT1 contacts (31) in our sample (n = 5). Left, low magnification image of Foxp2-V1 interneurons (eGFP, green), RV-mCherry labeling (red) of LG muscle afferents in the dorsal horn and interneurons presynaptic to the LG and of TA afferents anterogradely labeled with CTB (white). The Foxp2-V1 interneuron contains mCherry (yellow cell, inside box). This cell is magnified in two panels to the right, one showing Foxp2-V1 and RV-mCherry and the other RV-mCherry and CTB labeling. Arrows in the zoomed image show examples of CTB synapses (confirmed with VGLUT1) on the dendrites of this neuron. Far right image is the 3D reconstructed cell (Neurolucida) with CTB/VGLUT1 synapses indicated on its dendrites by yellow stars. The axon initial trajectory is indicated (the axon is lost at the section cut surface). The blue area highlights lamina IX. G, Low magnification showing TA-CTB afferents (white) and LG-RV-mCherry labeled interneurons (red). Transsynaptically labeled interneurons are categorized according to position and Foxp2-V1 lineage labeling: images below show superimposed Foxp2-V1 eGFP (green) and additional VGLUT1 immunolabeling (blue). The location of the LMC is indicated. Contralateral interneurons were found in LX (as the one in this section) and in LVIII (the other two in this animal, not shown). H, High magnification images of two LVII LG-coupled interneurons (RV-mCherry, red) receiving synapses from TA afferents (dendrites in boxed regions are shown at high magnification demonstrating CTB-TA labeling and VGLUT1 content). The most medial interneuron belongs to the Foxp2-V1 lineage (see inset with +eGFP). I, Neurolucida neuronal reconstructions showed that the Foxp2-V1 interneuron contained a medium number of TA/VGLUT1 synapses (13 contacts) in our sample of putative IaINs derived from the Foxp2-V1 lineage (n=5), while the non-Foxp2-V1 interneuron contained the largest (115) of any LVII reconstructed interneuron with mCherry, including many proximal synapses. However, one reason for this difference is that the dendrites projecting toward regions with high densities of TA CTB/VGLUT1 synapses were out of this tissue section. J, Intersection of otp and en1 using the dual color strategy with simultaneous expression of the Ai9 tdT and RCE-DC eGFP reporters. V1 cells that expressed otp are labeled with eGFP (green). In these cells the Ai9-tdT reporter is effectively deleted by flpo recombination dependent on otp expression level: strong (eGFP only) and weak (eGFP and tdT). V1 cells that express tdT only (red) never express otp. In addition, the sections were labeled with antibodies for Otp (light blue) to reveal cells that retained expression of Otp at P5 and ChAT (deep blue) to localize the motor pools. Most V1s express eGFP, with tdT-only cells being a minority. Right, cell plot positions of some of the cell types identified in these sections (one mouse 6 ventral horns in L4/L5). Most cells are Otp-V1 and include green (eGFP only) and pink cells (eGFP and tdT). K, Quantification of cells with eGFP only (green), tdT only (red), or both (pink) (n=12 ventral horns from 2 mice, error bars show SD). 88.6% of V1s expressed eGFP, most without tdT (61.1% of V1s) and of these around half (53.3%) retained Otp expression by P5. No Otp expression (antibody staining) was detected in tdT cells, and very few cells co-expressing tdT and eGFP (“weak” Otp expression) were Otp-IR. L, Detection of Pou6f2, calbindin and Otp in genetically labeled Otp-V1 cells. Left confocal image and right cell plot (n=6 ventral horns from 1 mouse). Pou6f2-Otp-V1 cells concentrate in a dorsal band within the ventral horn. Calbindin-IR Otp1-V1 cells concentrate in the Renshaw region (ventral most region) and others are in the dorsal region of the ventral horn. Otp-V1 cells retaining Otp expression at P5 occupy all dorsoventral positions in the lateral spinal cord. M, Quantification of 6 ventral horns at L4/L5 from one animal; 397 Otp-V1 cells sampled. Most (72.5%) did not express Pou6f2 or calbindin, 12.8% expressed Pou6f2 and 14.4% calbindin. Only one Otp-V1 neuron in the whole sample co-expressed Pou6f2 and calbindin (0.25%).

Next, we examined Otp and Foxp2 expression in genetically labeled Foxp2-V1 interneurons receiving VGLUT1 synapses (Figure 8D). On average (n=2 mice, 12 ventral horns and 1,116 Foxp2-V1 cells analyzed with similar representation in both animals), 42.5% of Foxp2-V1 interneurons had VGLUT1 contacts and were Otp+ while 22.1% were Otp(-), that means, 65.8% of Foxp2-V1 interneurons targeted by VGLUT1 synapses express Otp. On average only 9.3% of Foxp2-V1 interneurons that were Otp+ lacked VGLUT1 synapses. Thus, we detected VGLUT1 synapses on 82.0% of Otp+ Foxp2-V1 interneurons and 64.9% received these synapses proximally (cell body and primary dendrites) in addition to also having VGLUT1 synapses on more distal dendrites. Sections immunolabeled for Foxp2 revealed that 38.2% of Foxp2-V1 interneurons had VGLUT1 contacts and expressed Foxp2 while 17.4% were Foxp2(-). Moreover, only 14.0% of Foxp2-V1 interneurons retaining Foxp2 expression at P5 lacked VGLUT1 synapses. This suggests that 68.7% of interneurons in the Foxp2-V1 lineage that were targeted by VGLUT1 synapses retain Foxp2 expression at P5, and that 73.3% of Foxp2-V1 interneurons with genetic and antibody labeling received VGLUT1 synapses. Of these cells, 60.2% received VGLUT1 inputs proximally. Moreover, in one of the two animals we also analyzed Otp and Foxp2 co-localization in Foxp2-V1 interneurons receiving VGLUT1+ synapses. We found that 85.6% of Otp+ cells in the Foxp2-V1 lineage and receiving VGLUT1 synapses also express Foxp2 at P5, while 89.8% of dual genetic and antibody labeled Foxp2-V1 cells receiving VGLUT1 synapses are also Otp+. In conclusion, most laterally positioned Foxp2-V1 interneurons that express Otp and Foxp2 at P5 receive VGLUT1 synapses. VGLUT1 synapses in the ventral horn at P5 originate solely from parvalbumin+ proprioceptors, most likely Ia afferents (Alvarez et al., 2004).

Foxp-V1 interneurons integrated in reciprocal inhibitory circuits between antagonistic muscles

To examine whether some of these cells have inputs and outputs embedded in reciprocal inhibitory circuits, we combined anterograde labeling of sensory afferents with cholera toxin subunit B (CTB) injected in the tibialis anterior (TA) muscle with retrograde monosynaptic labeling of premotor interneurons using glycoprotein (G) deleted mCherry rabies virus (RVΔG-mCherry) injected in the antagonistic lateral gastrocnemius muscle (LG) of mice with eGFP labeled Foxp2-V1 interneurons (Figure 8E). The sections were immunolabeled for CTB, VGLUT1, mCherry and eGFP. To obtain transcomplementation of RVΔG-mCherry with glycoprotein in LG motoneurons we first injected the LG muscle with an AAV1 expressing B19-G at P4. To optimize muscle targeting and avoid cross-contamination of nearby muscles we performed RVΔG and CTB injections at P15. Analyses were done at P22, a timepoint that is after developmental critical windows through which Ia(VGLUT1+) synaptic numbers increase and mature on V1-IaINs (Siembab et al., 2010).

Unfortunately, motoneuron infection from muscle and transsynaptic retrograde labeling using RVΔG is known to be inefficient after P10 (Stepien et al., 2010). Additionally, at older ages transsynaptic transport is slower and more temporally spread such that fewer interneurons are recovered at single time points after injection. We chose 7 days post-injection for analyses to avoid as much as possible the cell degeneration that occurs at longer survival times after RV infection. We traded the low yield of these experiments for higher specificity when identifying synaptic inputs from TA sensory afferents on Foxp2-V1 interneurons that are premotor to the LG motor pool. We injected 5 animals that were analyzed in serial sections from L2 to L6 segments. All mice showed consistent TA anterograde labeling that in the ventral horn of the L4/L5 segments occupies the dorsal third of LVII and LIX (Figures 8F and 8G). This distribution matches the well-known musculotopic trajectories of central Ia afferents axons in the ventral horn (Ishizuka et al., 1979). In agreement with the known rostro-caudal trajectories of Ia afferent terminal axon collaterals entering the spinal cord, TA-CTB VGLUT1 labeled synapses were found in all lumbar segments, but caudal lumbar segments had the largest density of labeling in the ventral horn. Additionally, there were dense projections to medial LV and to discreet regions in LIII and LIV in all segments in all animals. In contrast, only 3 mice showed transsynaptic transport of RVΔG-mCherry from the LG motor pool to interneurons, with large variability from animal to animal. In the best animal we recovered 51 transsynaptically labeled interneurons with no evidence of degenerative phenotypes. These cells were found at the same locations, and in similar proportions as was reported by other groups using injections in younger animals where more cells were labeled (Stepien et al., 2010; Tripodi et al., 2011; Ronzano et al., 2022). The interneuron sample included cells in the Renshaw area (n=6 or 11.8%), LVII (15, 29.4%), dorsal horn medial LV (7, 13.7%), dorsal horn LI to LIV (20, 39.2%), and the contralateral spinal cord (3, 5.9%: 1 in LX and 2 in LVIII). Pooling cells with transsynaptic labeling from all three animals, we identified 8 LVII Foxp2-V1 interneurons. Their dendritic arbors were reconstructed in Neurolucida following the mCherry labeling. Five of these cells received more than 1 TA synaptic contact (CTB+ and VGLUT1+) although with large differences in number (5, 9, 12, 17, and 31 synapses). Most synaptic contacts occurred on dendrites, particularly those crossing areas with many CTB-labeled TA afferents. Therefore, the presence and direction of cut dendrites in the section strongly influences the total number of synapses detected in different cells. VGLUT1 contact densities on spinal interneurons are highly dependent on dendritic trajectories with respect to VGLUT1 synaptic fields (Siembab et al., 2016). The Foxp2-V1 cell shown in Figure 8F received the most contacts, which were concentrated on a dendritic segment crossing a field with high density of TA afferent synapses. Within single sections we found LG-coupled Foxp2-V1 and non-Foxp2 LVII interneurons receiving TA synapses on dendrites located in areas with high densities of TA/VGLUT1 synapses (Figure 8G and 8H). Non-Foxp2 IaINs could be derived from V2b or even non-Foxp2-V1 cells. It is known that several genetic subclasses of interneurons contribute to the full repertoire of IaINs controlling different leg joints (Zhang et al., 2014). These results provide proof-of-principle that some Foxp2-V1 interneurons are in synaptic circuits capable of exerting reciprocal Ia inhibition between antagonistic muscles. Clearly, a technique with higher yield and that maintains high specificity is necessary. Additionally, analysis of further extensor-flexor pairs in different joints in both directions will need to be performed to reveal a complete picture of IaIN organization.

Other cell types transsynaptically labeled from the LG included V1-Renshaw cells which did not receive any TA synapses since their dendrites are far away from TA projection areas. Medial LV LG-coupled interneurons (possibly Ia/Ib interneurons not derived from either V2b or V1 classes) are embedded within a region of high density of TA/VGLUT1+ synapses. A few reconstructed neurons at this location (n=3) showed they receive the highest densities of TA synapses with more than 50 contacts on relatively smaller dendrites. Some other LG-coupled interneurons reside in superficial laminae of the dorsal horn (Figure 8G) where they were contacted by CTB-labeled TA afferent synapses and by RVΔG-mCherry-labeled LG afferents. Muscle afferents ending in superficial laminae are likely non-proprioceptive (i.e., Type III(Aδ) and IV(C) afferents (Ling et al., 2003)). It has also been shown that interneurons at this location can be transsynaptically labeled in the anterograde direction from sensory afferents that incorporate AAV1-G and RVΔG-mCherry from the muscle injection (Zampieri et al., 2014). In our experiments few mCherry-labeled primary afferents were located below LIV suggesting that ventral interneurons, including Foxp2-V1 cells, were most likely transsynaptically labeled in the retrograde direction from LG motor pools.

The timing of these tracing experiments prevented analysis of Otp expression in Foxp2-V1 interneurons because of Otp downregulation at these postnatal times. To best target Otp-V1 interneurons we tried a genetic approach by breeding en1cre/+::Ai9-R26tdTomato mice with new Otpflpo/+ animals crossed to RCE:dual-eGFP. The Otpflpo/+ mice were generated by inserting Flpo-pA into the ATG in exon 1 of the Otp gene (Supplemental Figure 8A). After crossing Otp-flpo with RCE.fsf.GFP reporter mice, we show that genetically labeled Otp cells are distributed throughout the ventral horn of P0 mice and that 98.1% ±2.1% (mean±SD; n=3 mice) of Otp+ cells at P0 are genetically labeled (Supplemental Figure 8B). In dual color en1cre/+::Ai9-R26tdTomato Otpflpo/+::RCE:dual-eGFP mice, we found at P5 a surprisingly large proportion of V1s labeled (n = 2 mice, 6 L4/L5 ventral horns analyzed in each). The sections were also immunostained for ChAT (for best identification of the LMC) and for Otp to reveal expression at P5 (Figure 8J). Of 1,278 V1 interneurons analyzed in both animals we found 60.8% were eGFP (en1 and Otp intersection), 27.8% were “yellow” (en1 and “weak“ otp) and surprisingly only 11.4% were tdT only (en1 and no otp). As expected, P5 Otp expression detected with antibodies was absent in most tdT labeled V1 cells (93.5% of cells) and “yellow” tdT+eGFP V1 cells (85.4%), but also in a significant proportion of eGFP labeled V1 cells (46.7%) (Figure 8K). This suggests that otp is expressed by many V1 cells during embryonic development but is downregulated in around half of them by P5. Cells of the Otp-V1 lineage are located at all V1 medio-lateral and dorso-ventral locations of the V1 distribution (Figure 8J)

V1-cells transiently expressing Otp in embryo included cells of other clades. To examine this we generated a en1cre/+::otpflpo/+::RCE:dual-eGFP mouse in which only cells at the intersection of en1 and Otp express eGFP. L4/L5 sections from this animal were immunolabeled simultaneously for Pou6f2, calbindin and ChAT and is some serial sections we immunolabeled for Otp. We plotted all Otp-V1 cells (n=519) in 6 sections from this animal according to Pou6f2 or calbindin immunoreactivity (-IR) and then added the locations of Otp-IR cells from the Otp-V1 lineage in 2 further serial sections (Figure 8L). Pou6f2 was detected in 13.1% of Otp-V1 cells at P5. These cells occupy the dorsal band of the V1 cell distribution typical of the Pou6f2 clade. In addition, 14.7% of Otp-V1 cells were calbindin+ and this included many in the Renshaw cell ventral region and few others located more dorsally. One rare dorsal Otp-V1 cell contained both Pou6f2 and calbindin (included in both percentages above). It is well-known that calbindin in V1s becomes restricted to Renshaw cells throughout postnatal development, mostly in the second postnatal week (Siembab et al., 2010; Lane et al., 2021). By limiting the analysis to ventral Otp-V1 interneurons in the Renshaw area we estimated that 8.3% of them are Renshaw cells. Otp-V1 cells retaining Otp expression at P5 distribute to all dorso-ventral locations but are restricted laterally (Figures 8L and 8M). They represent 36.5% of all lineage-labeled Otp-V1 neurons.

In conclusion, the Otp-V1 lineage includes cells from several V1 clades that downregulate Otp expression before birth. In addition, many medial non-V1 Foxp2 cells also express Otp. Therefore, to specifically target the lateral group of proprioceptive Otp-Foxp2-V1 cells tightly associated to the LMC a triple genetic intersection or alternatively postnatally timed Otp-dependent recombination is necessary.

Discussion

A comprehensive inventory of spinal premotor interneurons and circuits requires detailed cataloging of its core components, ideally using multimodal information such as genetic subtypes, timing of neurogenesis, settling positions, incorporation within spinal circuits and their electrophysiological properties and functional roles in the circuit during motor behaviors. In the present work we focus on V1 inhibitory interneurons, a major and heterogenous group of inhibitory interneurons with ipsilateral synaptic projections throughout the ventral horn (Alvarez et al., 2005). Previously, the diversity of V1 interneurons in mice was organized into four clades deduced from Bayesian statistical analyses of combinatorial expression of 19 TFs and their cell locations in the spinal cord (Bikoff et al., 2016; Gabitto et al., 2016). Each group divides into further subgroupings organized in a hierarchical fashion. Here we show that the four V1 clades differ in neurogenesis windows and targeting of motoneurons. This suggests unique early developmental pathways and specific placings in spinal circuits validating the significance of V1 clade divisions as unique functional subtypes.

Two V1 clades, Renshaw cells and Pou6f2-V1 cells have narrow windows of neurogenesis with almost all cells generated before E11, while the Foxp2-V1 and Sp8-V1 clades are generated in a wider temporal window with most born after E11. Within the “early group” many Renshaw cells precede Pou6f2-V1 cells and within the “late group” Sp8-V1 cells lag behind Foxp2-V1 cells. These data confirm a previous report on earlier generation of MafB-V1 Renshaw cells compared to Foxp2-V1 IaINs (Benito-Gonzalez and Alvarez, 2012). Sequential generation of these two V1 cell types seem to be an intrinsic property of p1 progenitors and can be replicated in vitro using mouse embryonic stem cells (mESC) to derive p1 and V1 cells (Hoang et al., 2018). The present data now extends the view of sequential determination of cell fate by time of neurogenesis to all V1 clades. However, the results also suggest that while some clades with few neurons and limited diversity have narrow windows of neurogenesis, larger V1 clades like Foxp2-V1 interneurons with wider neurogenesis windows include subtypes that differ in location and synaptic inputs.

Overall, our data agree with a previous report that examined spinal cord subtypes based on the intersection of neurogenesis and transcriptomics by analyzing mouse spinal cords from E9.5 to E13.5 (Delile et al., 2019). This study divided V1 interneurons into seven groups. Groups V1.5 and V1.2 had early birthdates. The V1.5 group’s gene expression profile includes neurod1,2 and 4, neurod1, prox1, tcf4, lhx1 and 5, pou2f2 and hes1, and although it does not clearly match the TF repertoire of the Pou6f2-V1 clade, it occupies a similar birthdate window. In contrast, the V1.2 group shares the gene expression profile of Renshaw cells that includes calb, mafB and onecut2. One difference is that we have consistently found Renshaw cells to be the first born V1 cells and this study place them after the V1.5 group. One possible explanation is that some TFs used to define V1.2 cells, like mafB, are upregulated sometime after neurogenesis (Benito-Gonzalez and Alvarez, 2012). V1.1, V1.4 and V1.3 groups follow in neurogenesis timing, and all three express foxp2; V1.1 and V1.4 have similar birthdates while V1.3 neurogenesis is slightly delayed. We show similar heterogeneity of birthdates in Foxp2-V1 interneurons. Two subgroups expressing the TF Otp at P5 with or without Foxp4 and/or Foxp2 co-expression have intermediate times of neurogenesis and form a lateral group that is closely related to LMC motoneurons. They both receive VGLUT1/proprioceptive inputs and some form reciprocal inhibitory circuits between antagonistic motor pools. Many Foxp2-V1s located dorsomedially are generated earlier, and Foxp2-V1s located ventromedially are generated last. Ventromedial Foxp2-V1s are not targeted by VGLUT1/proprioceptive inputs which indicates different circuit roles. Finally, groups V1.6 and V1.7 are generated at the end of V1 neurogenesis, and it is tempting to speculate they include the Sp8-V1 clade and the late generated ventromedial group of Foxp2-V1 interneurons. However, it is presently difficult to match gene expression profiles of V1.6 and V1.7 (neurod1, 2, and 6, Nfia, Nfib, Nfix, prox1, tcf4, hex6, cbln2 and slit2) to our V1 clades. One problem is that the high throughput sequencing in Delile’s paper is based on samples collected at early embryonic times (E9.5 to E13.5), but gene expression in V1 groups changes throughout embryonic and postnatal development. For example, our efforts to generate a genetic model for Otp-V1s show that the postnatal restriction of this TF to the Foxp2-V1 clade does not occur until after embryogenesis. A similar situation occurs with the Sp8-V1 clade that is best defined by V1 interneurons retaining Sp8 expression at P0 (Bikoff et al., 2016), as Sp8 expression is widespread within ventral spinal progenitors during embryogenesis. On the other hand, we obtained preliminary evidence that the medial Foxp2-V1 group upregulates Foxp2 expression in late embryo (after E14) by difference to the lateral group that upregulates Foxp2 expression as they emerge from progenitors (Benito-Gonzalez and Alvarez, 2012). Thus, Foxp2 expression might not be captured in this group of Foxp2-V1 interneurons when examining V1 interneuron gene expression from E9.5 to E13.5. Despite these differences, there is general agreement in that Renshaw cells are early born and that several Foxp2 groups are generated later and can also be separated by differences in neurogenesis.

In addition to differences in neurogenesis, we uncovered differences among V1-clades in motoneuron synaptic targeting which supports the view that they constitute unique functional subsets. Renshaw cells and Foxp2-V1 interneurons are the major sources of V1 synapses on motoneuron cell bodies and proximal dendrites, while the Pou6f2-V1 and Sp8-V1 clades either do not provide much input to the motoneuron or these occur quite distally. We previously reported that the density of Sp8-V1 synapses on motoneuron cell bodies and proximal dendrites is between one and two orders of magnitude less dense compared to Renshaw cells synapses at this location (Bikoff et al., 2016). This analysis included a variety of motoneurons innervating flexor and extensor muscles at various hindlimb joints ruling out the possibility that each interneuron preferentially target specific motor pools. Unpublished data from our group of intracellular fills of dorsal MafB-V1 interneurons belonging to the Pou6f2 also show no axon projections to LIX. Given that V1 axons have limited projections outside the ventral horn, it is fair to conclude that Pou6f2-V1 and Sp8-V1 cells likely modulate activity in ventral premotor spinal networks and perhaps modulate synaptic integration in motoneuron distal dendrites traversing LVII, but they do not establish proximal synapses for efficient modulation of motoneuron firing and excitability. That role seems exclusive to Renshaw cells and Foxp2 interneurons within V1 interneurons, although these cells can also exert modulation of other premotor network elements. Within the Foxp2-V1 clade it is not possible at present to define which subgroup provides proximal synapses to motoneurons. However, given the known placement of IaINs (Jankowska and Lindstrom, 1972) and the known proximal location of reciprocal inhibitory synaptic inputs (Burke et al., 1971), it is reasonable to expect that many Foxp2-V1 synapses on the cell bodies and proximal dendrites of motoneurons arise from Foxp2-V1 lateral groups (II and III) which occupy spinal cord spaces typical of IaINs and receive VGLUT1/proprioceptive projections.

Intriguingly, hierarchical gene clustering of V1 subgroups shows that Renshaw cells (V1.2) and the three groups of Foxp2 V1s (V1.1, V1.2, V1.4) are closely related but distant from the V1.5, V1.6, V1.7 subgroups (Delile et al., 2019) which may include the Pou6f2-V1 and Sp8-V1 clades. This suggests a closer genetic relationship between Renshaw cells and Foxp2-V1 cells than with other V1 groups. Birthdate order is therefore not predictive of synaptic coupling with motoneurons, but hierarchical closeness according to early genetic programs is. This agrees with the general view that early genetic programs tightly define axonal projections. For example, spinal interneurons of different V-subtypes differ in their axons being mainly ipsilateral, contralateral, bilateral, ascending or descending (Goulding, 2009). Relationships between V1 clades may extend to phylogeny. Most genetic subclasses of spinal interneurons increase in genetic and functional diversity from zebrafish to mice (Wilson and Sweeney, 2023). V1 interneurons are present in all vertebrate species from fishes with swimming locomotion to mammals with limbed terrestrial locomotion, but they show low diversity in zebrafish (Kimura and Higashijima, 2019) and large heterogeneity in mice (Bikoff et al., 2016). Similarly, reciprocal inhibition of ipsilateral muscle flexors and extensors at limb joints first appears in limbed amphibia, a species that lacks recurrent inhibitory circuits (Czeh, 1977). As of today, recurrent inhibition of motoneurons by Renshaw cells has only been detected in mammals and in developing hindlimb motor pools of chick embryos (Wenner and O’Donovan, 1999; Alvarez and Fyffe, 2007). It is thus possible that these V1-clades are phylogenetically more recent and were sequentially added to the V1 interneuron repertoire in parallel with the evolution of limb motor control. Their significance is further highlighted by recent studies in ALS mouse models showing that V1 synapses located specifically on the cell bodies and proximal dendrites of motoneurons, along with Renshaw cells and Foxp2-V1 interneurons, are preferentially affected during early progression of the disease (Wootz et al., 2013; Salamatina et al., 2020; Allodi et al., 2021)

The birthdate distributions of Renshaw cells and Foxp2-V1 cells occupy the whole period of V1 neurogenesis which suggests that birthdate order might not be a predictor of phylogenetic position. In addition, the axial MMC (shared with fish) is also densely innervated by Foxp2-V1s and Renshaw cells, and thoracic axial motoneurons in the cat are modulated by Renshaw recurrent inhibitory axons (Saywell et al., 2013). This suggests that axial motoneurons in limbed mammals acquire inhibitory controls that are non-existent in aquatic vertebrates. Despite the similarities between Renshaw cells and Foxp2-V1 interneurons, they have very different electrophysiological and firing properties in the mature state. Renshaw cells are burst-tonic firing cells, but Foxp2-V1 interneurons are tonic fast-spiking (Bikoff et al., 2016). This suggests differences in the ion channel ensembles they express at maturity. This divergence develops postnatally. Indeed, the firing properties of Renshaw cells in early embryos (E12) (Boeri et al., 2018; Boeri et al., 2021) greatly differ from those expressed postnatally.

In conclusion, the results suggest that previously defined V1 clades show differences in birthdate, heterogeneity, projections to motoneurons and/or premotor networks, and collectively may represent interneurons that differ in their relationship to the evolution of limb function.

Molecular modulation of V1 neurogenesis

In a remarkable in vitro replication of the V1 neurogenesis sequence using mouse embryonic stem cells (mESC), it was found that 24 hr. treatment with a high concentration of retinoic acid and low smoothened agonist induced mESC-derived p1 progenitors. These progenitors sequentially generated V1 interneurons with genetic profiles of Renshaw cells at day 5 and Foxp2-V1s at day 8 in vitro (Hoang et al., 2018). Lengthening the retinoic acid treatment favored differentiation into cells with Renshaw cell characteristics at the expense of Foxp2-V1s. This suggests that Renshaw cells and Foxp2-V1s derive from a similar pool of p1 progenitors and their fates might depend on differences in the morphogenetic signals present at the embryonic times when they are born. For example, in early embryos retinoic acid is highly expressed by the mesoderm adjacent to the developing spinal cord (Novitch et al., 2003) and by motoneurons inside the spinal cord (Sockanathan et al., 2003), while later retinoic acid signals are attenuated by opposing actions from fibroblast growth factor family members (Diez del Corral et al., 2003). Another consideration, however, is that although Hoang et al. also differentiated V1 cells with gene expression profiles typical of Pou6f2 and Sp8 cells, these clades were underrepresented. It is thus possible that p1 progenitors giving rise to Pou6f2 and Sp8 V1 cells have different signal requirements during embryogenesis.

Significance of Foxp2-V1 interneuron diversity

Genetic lineage labeling of Foxp2-V1 interneurons reveals approximately double the number of V1 interneurons than when this clade is defined by postnatal Foxp2 protein expression. Not surprisingly we uncovered high heterogeneity. We defined four groups according to TF combination/position (Figure 7G), but that also reflect birthdate order. Of special interest were groups II and III which were born at mid neurogenesis times and expressed Otp at P5. These cells were closely related to the limb-controlling LMC. Foxp2-V1 interneurons at this location also receives proprioceptive VGLUT1 inputs and some were in reciprocal inhibitory circuits between the TA and LG. There were also non-Foxp2-V1s cells in LVII close to the LMC with similar connections (input from TA proprioceptive afferents, output to the LG motor pool). This agrees with reciprocal inhibition being mediated by more than one genetically defined group of inhibitory interneurons (Zhang et al., 2014). Proprioceptive TA input densities on different dendrites of single neurons varied. This agrees with the idea that afferent inputs are established on interneuron dendrites according to their location relative to the trajectories of specific primary afferents, a concept established for the major inputs of Renshaw cells (proprioceptive and motor) (Mentis et al., 2006; Benito-Gonzalez and Alvarez, 2012; Siembab et al., 2016). Similarly, motoneurons with genetically altered dendritic arbor structure show altered proprioceptive inputs (Vrieseling and Arber, 2006). In this context, the two groups of Otp-Foxp2-V1 interneurons defined by Foxp4 expression and ventral location (group III) and lack of Foxp4 more dorsal (group II), are likely coupled to primary afferents from different muscle groups.

In conclusion, Foxp2-V1 interneurons are a highly diverse group of interneurons. Types I, II/III, and IV may be broadly related to different motor functions while types II and III may be closely related to proprioceptive pathways and differ in muscle, joint, flexor, and/or extensor specificities.

Materials and methods

All animal procedures were carried out in accordance with NIH guidelines and were approved by the Emory University Institutional Animal Care and Use Committee (Atlanta, GA). Animal experimentation also follow ARRIVE guidelines.

Animal models

To genetically lineage-label all or subclasses of V1-interneurons we used eight transgenic mouse models (Table 1). These mice were crossed for intersectional genetic labeling combining a line in which all V1 interneurons express Cre with lines expressing Flpo dependent on Foxp2 or Otp or in which MafB expressing cells express GFP.

V1 and Foxp2-V1 model

en1Cre/+ heterozygotes (Sapir et al 2004) were crossed with Rosa26-Frt-lox-STOP-lox-tdTomato-WPRE-frt homozygotes (Ai9R26-tdT, JAX#007909; B6;129S6-Gt(ROSA)26Sort M14(CAG-tdtomato)Hze/J) to obtain en1cre/+::Rosa26lsl-tdT/+ and after backcrossing, en1cre/+::Rosa26lsl-tdT/lsl-tdT animals (in both animals all V1 cells are lineage-labeled with tdT). Similarly, foxp2flpo/+animals (Bikoff et al 2015) were crossed with RCE:dual-EGFP homozygotes (RCE:FRT JAX#010675; Gt(ROSA)26Sortm1(CAG-EGFP)Fsh; initially donated by Dr. Gordon Fishell, Harvard University) to produce foxp2flpo/+::RosadualEGFPl/+and foxp2flpo/+::RosadualEGFP/dualEGFP mice. Crossing en1Cre/+::Rosa26lsl-tdT/+ or en1cre/+::Rosa26lsl-tdT/lsl-tdT mice with foxp2flpo/+ or foxp2flpo/+::RosadualEGFP/+ or foxp2flpo/+::RosadualEGFP/dualEGFP we obtained mice for experiments with the following genotypes: en1cre/+::foxp2flpo/+::Rosa26lsl-tdT/+(V1 red), en1cre/+::foxp2flpo/+::Rosa26+/dualEGFP (Foxp2-V1 green) or en1cre/+::foxp2flp/+::Rosa26lsl-tdT/ dualEGFP (dual color, Foxp2-V1 green and non-Foxp2-V1 red). In some experiments we substituted the reporter lines Ai9-lsl-tdT and RCE:dualEGFP for the RC::FLTG line. This line has in the R26 locus a frt-flanked STOP and loxP-flanked tdTomato::STOP preventing transcription of EGFP (JAX#026932,B6.Cg-Gt(ROSA)26Sortm1.3(CAG-tdTomato,-EGFP)Pjen/J). In this line, Flpo recombination in cells with Foxp2 expression induces expression of tdT, while cells with additional Cre recombination (V1 expressing Foxp2 cells) will express EGFP and remove the tdT reporter.

V1-Otp model

Similar breeding schemes and reporter lines were used to combine en1cre/+ and otpflpo/+ mice to study Otp-V1 interneurons. These mice were generated as described (Bikoff et al., 2016) and summarized in Supplementary Figure 8. Briefly, Flpo, a codon-optimized version of Flp recombinase was inserted into the ATG in the 1st exon of the Otp genomic locus, generating a null allele. Positive ES cell clones were screened by Southern blot and microinjected into blastocysts, and the resulting chimeric mice were crossed to C57BL/6J females. The neomycin selectable cassette was removed using Protamine::Cre mice (Jax#03328).

V1-MafB model

en1cre/+::Rosa26lsl-tdT/lsl-tdTanimals were crossed with mafBGFP/+ knock-in mice (Moriguchi et al 2006) to label cells expressing MafB and assess their overlap with V1 interneurons.

The en1, foxp2, otp and mafB genes carry the recombinases or GFP inserted into the genomic loci resulting in null alleles. The animals are maintained and bred in heterozygosis, with homozygotes being knockouts for each of these genes. We generated foxp2flpo/flpo and mafBgfp/gfp homozygotes to test antibody specificities. foxp2 knockout mice survive postnatally, but mafB knockouts die at birth. Thus, foxp2 knockout mice were harvested at P5 and mafB knockout mice as late embryos. All animals were bred in our colonies at Emory University, and the resulting litters were genotyped using a combination of standard tail PCR or Transnetyx and fluorescent phenotyping of neonates (each gene combination results in specific patterns of labeling along the body).

Timed pregnancies

Female mice were caged with males at the beginning of the dark period (7:00 PM), and the next morning (7:00 AM) vaginal plugs were checked. A positive plug was considered E0.5; however, since the exact time of mating is unknown, this procedure has an approximate error of ±12 hr. Moreover, we found embryos within single litters that differ by 6-12 hr. in developmental stage.

Tissue preparation

Mouse pups of different postnatal (P) ages (P0, P5, P15, P30, adult) were anesthetized with an overdose of Euthasol (>100 mg/Kg i.p.) and after transcranial vascular rinsing with vascular rinse and heparin they were perfusion-fixed with 4% paraformaldehyde in 0.1M phosphate buffer (PB). The spinal cords were then dissected and removed, postfixed in 4% paraformaldehyde for either 2 hr, 4 hr, or overnight. Postfixation times depend on antigens targeted in immunocytochemistry experiments. In general, calcium buffering proteins, choline acetyltransferase and synaptic vesicle markers require longer postfixation times and transcription factors reauire shorter postfixation times. After postfixation, the tissues were cryoprotected in 0.1M PB with 30% sucrose and prepared for sectioning. Generally, transverse spinal cord sections were obtained in a sliding freezing microtome at 50 µm thickness and collected free-floating. Embryos, P0, and sometimes P5 spinal cords were cut in a cryostat at 20 µm thickness from tissue blocks snap-frozen in OCT.

Birthdating experiments

5-ethynyl-2’-deoxyuridine (EdU; Invitrogen) was injected i.p. at a dose of 50 mg/kg weight in timed-pregnant females. The data reported were obtained from 18 pregnant females successfully injected at gestation days E9.5, 10, 10.5, 11, 11.5, or 12 after crossing with appropriate males to generate pups with genetic labels for all V1, Foxp2-V1 or MafB-V1 interneurons. The spinal cords were collected after P5 perfusion-fixation with paraformaldehyde as above. P5 was chosen for analyses to maximize transcription factor antigenicity.

EdU Click-iT reaction

Fifty micrometer thick transverse spinal cord sections were obtained in a freezing, sliding microtome from lower lumbar segments (4 and 5) and processed free floating with Click-iT Alexa Fluor 488 (C10337, Invitrogen) or Click-iT Alexa Fluor 647 Imaging Kits (C10340, Invitrogen) depending on other genetic fluorophores present in the animal. Sections were washed twice (5 min) with 3% bovine serum albumin (BSA, Fisher) in 0.01M phosphate buffered saline (PBS) and then then permeabilized with a solution of 0.5% Triton X-100 (Fisher) in 0.01M PBS at room temperature for 20 min. During this time, the Click-iT reaction cocktail was prepared per manufacturer instructions and applied to the sections for 30 min at room temperature protected from light. Sections were washed with 3% BSA in 0.01M PBS at the conclusion of the incubation. Antibody labeling followed the EdU Click-iT reaction. In one pup EdU labeling did not correspond to the target time point and was discarded (#459.2 in Figure 1D). The littermate (#459.1 in Figure 1D) displayed correct EdU labeling for the injection time.

Immunohistochemistry

The characteristics, RRID numbers, and dilutions of all primary antibodies used are summarized in Table 2. Genetic labels, tdT and EGFP, were always amplified with antibodies to aid in visualization. After blocking the sections with normal donkey serum (1:10 in PBS + 0.3% of Triton-X-100; PBS-TX), we incubated the section in different primary antibody cocktails diluted in PBS-TX. Chicken antibodies were used to detect EGFP, and mouse or rabbit antibodies were used to detect tdT, depending on the hosts of primary antibody combinations. In birthdating experiments we used in serial sections either rabbit anti-MafB (Sigma), goat anti-Foxp2 (Sant Cruz), rabbit anti-Pou6f2 (Sigma), goat anti-Sp8 (Santa Cruz), guinea pig anti-Otp (Jessell lab; Bikoff et al., 2016), rabbit anti-Otp (Jessell lab; Bikoff et al., 2016), rabbit anti-Foxp4 (Jessell lab; Bikoff et al., 2016), or rabbit anti-calbindin (Swant). Depending on color combination for triple or quadruple fluorescent labeling with genetic reporters (EGFP or tdTomato) and the EdU Click-it reaction (Alexa 488 or Alexa 647), these markers were revealed with either FITC, Cy3 or Cy5 conjugated species-specific donkey-raised secondary antibodies, or with biotinylated secondary antibodies followed by streptavidin Alexa-405 (for all secondary antibodies dilution was from 1:100 to 1:200; all secondary reagents were obtained from Jackson ImmunoResearch). After immunoreactions, the sections were mounted on slides and cover-slipped with Vectashield (Vector Laboratories). Similar immunocytochemical protocols were followed in other experiments.

Analysis

Confocal images (10X and 20X) were obtained with an Olympus FV1000 microscope. Image confocal stacks were fed into Neurolucida (MicroBrightField) for counting and plotting cells. We analyzed 4 ventral horns per animal in lower lumbar segments (L4-L5). Cells were classified according to genetic labeling, TF immunoreactivity, and EdU labeling. EdU labeled cells were classified as strongly labeled (at least two thirds of the nucleus uniformly labeled) or weakly labeled (speckles or partial nuclear labeling). From Neurolucida plots we estimated: (1) the percentage of V1 INs labeled with EdU (strongly or weakly); (2) the percentage of V1 cells labeled with TF antibodies; (3) the percentage of V1 cells genetically labeled with MafB-GFP; (4) the percentage of V1 INs genetically labeled with Foxp2; (5) the percentage of V1 cells with different genetic or immunocytochemical markers and incorporating weak or strong EdU at different time points; (6) the cumulative numbers of EdU weakly or strongly labeled cells for all V1 cells and for each marker across all embryonic times.

Cell location/density analyses

From Neurolucida plots we constructed cell density profiles for each V1 interneuron type and birthdate, assigning cartesian coordinates to the nucleus location with respect to the dorsal edge of the central canal which was defined as position (0,0). Coordinates were exported as .csv files and plotted using custom Matlab scripts to display the position of each individual cell (Bikoff et al., 2016). Distribution contours were constructed in Matlab using the kde2d function (Matlab File Exchange), which estimates a bivariate kernel density over a set of grid points. We plotted density contours containing from 5% (inner contours) to 95% (most outer contour) of the cell population in 10% increments. Density contours were superimposed to hemicord schematic diagrams in which the distance from central canal to lateral, dorsal, or ventral boundaries was adjusted depending on age and segment from direct measurements obtained in Neurolucida.

Analyses of lineage-labeled Foxp2-V1, Otp-V1 and MafB(GFP)-V1 cells across ages and markers

The spinal cords of animals carrying different combinations of genetic labels were prepared as above for amplification with immunolabeling of their fluorescent reporters and combination with different markers (for the characteristics of the different samples in terms of number of animals and sections analyzed see the text in results).

Analysis of Foxp2-V1 interneurons in different segments of the adult spinal cord

We used en1cre/+::foxp2flpo/+: :Rosa26+/dualEGFP mice and amplified the EGFP signal as above. The position of motoneurons was revealed using a goat anti-choline acetyltransferase (ChAT, Millipore) antibody laminae cytoarchitectonics with a rabbit recombinant anti-NeuN antibody (Novus) (not shown in figures). ChAT and NeuN immunoreactivities were respectively detected with Cy3 and Cy5 conjugated donkey anti-goat IgG antibodies (Jackson ImmunoResearch). Confocal microscopy images were obtained (as above), plotted in Neurolucida and cell density contours generated. We analyzed the number of Foxp2-V1 cells and cholinergic motoneurons per section and calculated their ratios (a motoneuron was defined as any cholinergic immunoreactive cell in Lamina IX). Density contours were used to qualitatively compare their distributions in different segments.

Analyses of transcription factor expression in Foxp2-V1 interneurons at P5

To examine Otp and Foxp4 immunoreactivity in lineage-labeled Foxp2 vs. non-Foxp2 V1 cells, we used dual color en1cre/+::foxp2flp/+::Rosa26lsl-tdT/ dualEGFP mice. The sections were immunolabeled with guinea pig anti-Otp (Jessell Lab, CU) and rabbit anti-Foxp4 (Jessell Lab, VU). Otp immunoreactive sites were revealed with biotinylated donkey anti-guinea pig followed by streptavidin Alexa-405 and Foxp4 immunoreactivity was detected with donkey anti-rabbit IgG secondary antibody conjugated to Cy5. These antibodies were combined with immunocytochemical amplification of the EGFP signal (chicken anti-EGFP) and tdTomato (mouse anti-RFP) respectively with FITC and Cy3 conjugated species-specific IgG antibodies raised in donkey. To analyze the populations of lineage-labeled Foxp2-V1 cells expressing Otp, Foxp4 and Foxp2 at P5, we used single color en1cre/+::foxp2flp/+::Rosa26+/ dualEGFP mice and amplified EGFP fluorescence with antibodies as above, combined with guinea pig anti-Otp (Jessell, CU)/rabbit anti-Foxp4 (Jessell, CU), rabbit anti-Otp (ThemoFisher)/goat anti-Foxp2 (Santa Cruz) and rabbit anti-Foxp4 (Jessell, CU)/goat anti-Foxp2 (Santa Cruz) antibodies. Triple immunostains were revealed with species-specific secondary antibodies as above using the Cy3 and Cy5 channels for transcription factors. A few sections were immunolabelled with rabbit antibodies against NeuroD2 (Abcam) or Prox1 (Millipore/Sigma). In this case immunoreactivities were revealed with Cy3-conjugated anti-rabbit IgG antibodies. All sections were imaged using confocal microscopy and confocal stacks analyzed in Neurolucida as above. All analyses were done in Lumbar 4 and 5 segments. From cellular plots we estimated co-localizations and cells distributions as explained above.

Analyses of V1 clade markers in MafB-V1 and Otp-V1 interneurons at P5

For these analyses we used en1cre/+::otpflpo/+::Rosa26+/dualEGFP or en1cre/+::Rosa26lsl-tdT/+::mafbGFP/+mice. To analyze Otp-V1 interneurons, EGFP fluorescence was amplified with chicken anti-EGFP antibodies as above, and one of the following additional primary antibodies was added in serial sections: rabbit anti-calbindin (Swant), guinea pig anti-Otp (Jessell CU), guinea pig anti-Pou6f2 (Jessell CU), goat anti-Foxp2 (Santa Cruz), or goat anti-Sp8 (Santa Cruz). All markers were detected in the Cy3 channel. When possible, they were combined with goat anti-ChAT antibodies (EMD Millipore) or mouse anti-NeuN antibodies (Millipore) labeled in the Cy5 channel. ChAT and NeuN immunoreactivities are not shown in results but were used to identify laminae and spinal cord segments according to motor column organization and cytoarchitectonics. In MafB-V1 mice we amplified both EGFP and tdT as above, and combined with the following antibodies: rabbit anti-calbindin (Swant), guinea pig anti-Pou6f2 (Jessell CU), goat anti-Foxp2 (Santa Cruz) and goat anti-Sp8 (Santa Cruz). All marker antibodies were revealed using Cy5 conjugated donkey species-specific anti-IgG secondary antibodies as above. Analyses were done as above: confocal images were imported into Neurolucida for cell plotting and the results expressed as number or proportion of neurons and their position analyzed using cell distribution density profiles. All analyses were done in Lumbar 4 and 5 segments.

Analyses of Foxp2-V1 and Foxp2-non-V1 cells at P5

For these analyses we generated two en1cre/+::foxp2flpo/+::Rosa26+/FLTG mice. EGFP (Foxp2-V1 cells) and tdTomato (Foxp2-non-V1 cells) were amplified with antibodies as above. The sections were counterstained with mouse anti-NeuN antibodies (Millipore) for segment and laminar localization. Sections from Lumbar 4 and 5 segments were imaged with confocal miscopy and analyzed in Neurolucida. Cell distribution plots, cell numbers, and cell density curves were obtained as above.

Analyses of VGLUT1 inputs on Foxp2-V1 interneurons

Analyses were done at P5 in en1cre/+::foxp2flp/+::Rosa26dualEGFP/+mice. The P5 age was selected to preserve TF immunoreactivity. Moreover, at this age VGLUT1 synapses in the ventral horn are specifically contributed by proprioceptive afferents, most likely Ia afferents (Alvarez et al., 2004). For these analyses spinal cord sections were obtained in a cryostat (20 µm thickness). EGFP was amplified with chicken anti-GFP (Aves) and combined with goat anti-Foxp2 (Santa Cruz), rabbit anti-Otp (ThermoFisher) and guinea-pig anti-VGLUT1 (Synaptic Systems). EGFP, Foxp2 and Otp immunoreactivities were revealed respectively with FITC, Cy3 and Cy5 conjugated species-specific secondary antibodies. VGLUT1 was revealed with biotinylated anti-guinea-pig IgG antibodies followed by streptavidin-Alexa 405. The sections were imaged at 10X and 60X with confocal microscopy. VGLUT1 contacts were analyzed only in high magnification images in which we tiled the whole ventral horn to sample every Foxp2-V1 present. Using Neurolucida, we plotted all Foxp2-V1 cells in each section and classified them according to the presence of VGLUT1 inputs and whether they were at high or low density in a qualitative assessment. Later we pooled all data into presence or abscence of VGLUT1 contacts. From these plots we estimated: (1) the percentage of lineage-labeled Foxp2-V1 interneurons with VGLUT1 contacts; (2) the percentage of these cells with Otp, Foxp2 or both TFs co-localized (for sample attributes with respect to number of animals, sections and cells analyzed, see results).

Identification of Foxp2-V1 interneurons interposed in reciprocal connections between the tibialis anterior muscle and the lateral gastrocnemius motor pool

For these analyses we combined retrograde monosynaptic tracing of RVΔG-mCherry from the LG muscles with anterograde tracing of muscle sensory afferent synapses using cholera toxin subunit b (CTb) from the TA. RVΔG-mCherry and CTb intramuscular injections were done at P15 to avoid critical windows of synaptic reorganization. To facilitate tracing at this age we applied RVΔG-mCherry at high titer (>109 TU/ml) and CTb was injected at high concentration (1%).

Production of RVΔG-mCherry

SADB19ΔG-mCherry rabies virus (RVΔG-mCherry) and the BHK-B7GG (B7GG) cells expressing B19 RV glycoprotein (RV-G) and nuclear GFP for virus propagation and amplification were donated by Dr. Edward Callaway (Salk Institute, La Jolla, CA). Sterile cell culture technique without antibiotics was used throughout all procedures. B7GG cells were placed in cell culture dishes containing DMEM in 10% fetal bovine serum (FBS) (culture medium) and incubated in 5%CO2 in humid air at 37°C for two hours. Once they adhered to the substrate, the medium was removed, the cultures washed with 10 ml warm PBS and fresh medium applied. Four plates were grown to 90% confluency, and then 4 ml of virus stock was added to each culture and incubated at 37°C/5%CO2 for 4 hours. After washing three times in warm PBS (to remove as much virus as possible) the cultures were incubated in fresh medium at 35°C/3%CO2 and monitored daily for fluorescence. Four plates of fresh B7GG cells were grown to 90% confluency, washed with 10ml PBS and detached with 6 ml of 0.25% trypsin for five minutes at room temperature (RT) with gentle rocking. A warm culture medium with 10% FBS was added to quench trypsin activity and the cells were dissociated by gentle trituration (approximately 12 passes). The cell suspensions were centrifuged at 800G for 3 mins to pellet cells. After removing the trypsin/culture medium they were re-suspended in warm culture medium. Twelve culture dishes containing 18 ml of culture medium were simultaneously inoculated with 2ml each of the B7GG cell suspension and incubated at 37°C/5%CO2. The culture medium was changed after 80-90% confluency and 4ml of viral supernatant added to each plate and incubated for 4 hours at 37°C/5%CO2., then culture medium was removed, the cells washed three times in PBS and 20ml fresh medium added to each plate. The cultures were incubated at 35°C/3%CO2 and monitored daily for expression of cytoplasmic mCherry and nuclear GFP. The cell culture medium containing RVΔG-mCherry was collected after 4 days, filtered through 0.45 µm membranes and placed on ice. RVΔG-mCherry was concentrated from 180 ml (10 plates) of cell supernatant via ultra-centrifugation, 2 hrs at 20,000G at 4°C. The supernatant was aspirated and the six viral pellets re-suspended in 200 µL each of cold Hanks Balanced Salt Solution (HBSS). All supernatants were combined and layered over 1.8 ml of 20% sucrose in HBSS in a 3 ml tube. The virus was pelleted through the sucrose cushion by ultra-centrifugation, at 20,000G for 2 hours at 4°C. The supernatant and sucrose cushion were gently poured off and any remining fluid aspirated. The pellet was re-suspended in 105 µL of HBSS by gentle agitation and 5 µL aliquots frozen at -80°C for later use.

Virus Titer

Serial dilutions of RVmCH (from one 5 uL aliquot of virus) from the preparation were inoculated onto 1 105 293T cells in a 48 well plate. As soon as mCherry was detectable (usually 2 days), the positive cells in the well (dilution) containing 10-100 cells were counted and a titer obtained. For these experiments we used virus at a titer of 1.92 E10 transfection units (TU)/ml.

Production of AAV1-G

Adeno-associated viruses were produced by the Emory Virus Core from AAV2 plasmids expressing the B19 RV glycoprotein under a CMV promoter. This plasmid was donated by Dr. Silvia Arber (Biozentrum, Basel)(Stepien et al., 2010). Virus titer was expressed in genomic copies (GC) and was 2.5XE12 vg/ml by qPCR for the lot used here.

Viral and tracer intramuscular injections

AAV1-G was injected at P4 in the LG (1-2 µL, undiluted), RVΔG-mCherry was injected in the same muscle at P15 (2-3 µL, undiluted) and CTB was injected in the TA at the same time (0.5 µL, 1% diluted is sterile saline). The animals were allowed to survive 7 days after the last injection (P22) at which time they were perfusion-fixed, and 50 µm thick sections were prepared as above.

Immunocytochemistry

Sections were incubated overnight in a cocktail of primary antibodies that included chicken anti-GFP (Aves), rabbit anti-DsRed (Clontech), guinea-pig anti VGLUT1 (Synaptic Systems) and goat anti-CTb (List labs). Primary antibodies were detected with species-specific antibodies coupled to FITC (for EGFP), Cy3 (for tdT), Cy5 (for CTb) or biotin (for VGLUT1). Biotinylated antibodies were exposed with streptavidin-Alexa 405.

Analyses

All sections with mCherry labeled cells were imaged at low (10X), medium (20X) and high (60X) magnification using confocal microscopy. Ia inhibitory interneurons were defined as neurons retrogradely labeled from the LG by RVΔG-mCherry and receiving inputs from TA CTb/VGLUT1 labeled boutons. Because this technique was relatively low yield. all analyses were qualitative. The numbers of animals and yields are reported in results.

Antibody specificities

The most frequently used antibodies in this study were tested in knockout mice: guinea-pig anti-VGLUT1 and rabbit anti-calbindin (Siembab et al., 2010), rabbit anti-Pou6f2, goat anti-Sp8, and guinea-pig anti-Otp (Bikoff et al., 2016), and goat anti-Foxp2 and two rabbit anti-MafB antibodies (this study, supplemental Figures 1 and 3). In addition, Foxp2 and MafB antibodies used here were further characterized using western blots (see below). Alternative antibodies against Pou6f2 or Otp were first confirmed in dual immunolabeling with validated antibodies. NeuroD2 and Prox1 antibodies were tested and used in recent literature (Osseward et al., 2021). Goat anti-ChAT antibodies gave the well-known patterns of cholinergic immunoreactive neurons in the spinal cord and coincide with genetically labeled neurons in ChAT-IRES-Cre-tdTomato mice. GFP, DsRed and RFP antibodies did not result in any immunolabeling in sections not expressing any of these fluorescent proteins. CTb antibodies resulted in no staining in naïve sections. Rabbit and mouse anti-NeuN antibodies gave identical results. The immunostaining of the mouse anti-NeuN monoclonal in the spinal cord has been amply characterized (Alvarez et al., 2004 and 2005).

Western Blots. Foxp2 and MafB antibodies were further characterized in western blots from spinal cord samples collected from wildtypes, heterozygotes (one null allele) and homozygous knockouts (both null alleles) (supplemental Figures 1 and 3). Lumbar spinal cords were dissected in oxygenated artificial cerebrospinal fluid and immediately homogenized using Cytoplasmic Extraction Reagent Kit from the NE-PERTM nuclear extraction kit (ThermoFisher) with a protease inhibitor cocktail added (5mg/ml. Complete Mini, Roche). The manufacturer’s instructions were followed to isolate nuclei and yield aliquots of nuclear proteins. A Bio-Rad DC protein assay was used to determine total nuclear protein content. Protein standards (Bio-Rad) were prepared in NER buffer from the NE-PER kit. Sample absorbance was read on a plate reader at 750 nm. Samples were stored at -80°C until use. For immunoblotting, the samples were prepared in standard SDS-PAGE sample buffer (5X) and 30 µL of nuclear protein from each spinal cord was added to each lane of a Bio-Rad precast 10% polyacrylamide gel. Bio-Rad Kaleidoscope molecular weight markers were added to one lane. Electrophoresis was carried out in Tris buffer (Bio-Rad) at 180V until the dye front reached the bottom of the gel. The proteins were then then transferred overnight onto PVDF membranes (Bio-Rad) using standard SDS-PAGE transfer buffer (Bio-Rad) and a constant 0.15 A current with gentle stirring at 4°C. For MafB antibodies, three immunostaining procedures were carried out successively on the same membrane. The membrane was washed three times with Tris Buffered Saline and Triton-X-100 (TBST) for 15 minutes and blocked with non-fat dry milk (Carnation, Nestle) for 1hr at room temp on a rocker. The first immunostaining employed a primary antibody against MafB (Novus) at a dilution 1:500 and worked best with 2.5% NFDM in TBST and incubated overnight at 4°C. The primary antibody was detected with donkey anti-rabbit IgG secondary antibodies conjugated to horseradish peroxidase (HRP, GE Health Sciences) and immunostained bands were revealed using enhanced chemiluminescence (ECL). The blot was then stripped for 20 minutes (Reblot Plus), washed 3 times for 15 minutes in TBS, blocked in 5% NFDM for one hour and re-probed with a primary antibody against MafA (Novus) at 1:500 in TBST and the immunoreactive bands detected via ECL as above. Finally, the blot was stripped a final time using ReBlot Plus (Millipore) for 20 minutes, blocked in 5% NFDM and re-probed with an antibody against MafB (Sigma). Secondary antibody and ECL were identical to the previous two primary antibodies. In a different sequence we substituted the MafB (Sigma) antibody for c-maf (Novus). Similar procedures were used to detect Foxp2 immunoreactivity (Santa Cruz) except that in this case the blot was probed only once, and we used anti-goat secondary antibodies coupled to HRP.

Statistical Analyses

All statistical analyses were performed in Prism (GraphPad ver 9). In all cases the samples passed the normality test. When comparing multiple groups we used one or two-way ANOVAs depending on the sample structure. To include consideration of repetitive measures in single animals (for example different motoneurons from single animals) we used Nested one-way ANOVAs. Post-hoc pair wise comparisons were always done using Bonferroni-corrected t-tests. When comparisons involved only two groups we used standard t-tests. All statistical details are provided in supplementary tables.

Figure composition

All images for presentation were obtained with an Olympus FV1000 confocal microscope and processed with Image-Pro Plus (Media Cybernetics) for optimization of image brightness and contrast. Frequently we used a high gaussian filter to increase sharpness. Figures were composed using CorelDraw (version X6) and graphs were obtained in Prism (GraphPad, ver 9).

Supporting information

eLife Assessment

This manuscript reports important findings regarding the development, anatomical placement and synaptic connectivity of a subtype of V1 spinal inhibitory interneurons. Using a combination of techniques, the authors show convincingly how V1 interneurons in the spinal cord, specifically those expressing the transcription factor Foxp2, differ in their birthdates, synaptic connectivity to motor neurons and their postnatal location. The study is an important addition to the literature on spinal cord interneurons and opens avenues for their functional assessment.

Reviewer #1 (Public Review):

To understand the spinal locomotor circuits, we need to reveal how various types of spinal interneurons work in the circuits. So far, the general roles of the cardinal groups of spinal interneurons (dI6, V0, V1, V2a, V2b, and V3) involved in locomotion have been roughly established but not fully understood. Each group is believed to contain some clades with more detailed functional differences. However, each character and function of these clades has not yet been elucidated.

In this study, Worthy et al. investigated clades of V1 neurons that are one of the main groups of inhibitory neurons in the spinal cord. Previous reports proposed four clades (Renshaw cells, FoxP2, sp8, and pou6f2) in V1 neurons defined by the expression of transcription factors. For V1 neurons in each of the four clades, the authors investigated the birth time and showed the postnatal location in the spinal cord according to the birth time. They found FoxP2-V1 located near LMC motor neurons that project to the limb. Using genetically labeled Foxp2-V1 mice, they showed that most of the synapses of V1 neurons on the cell bodies of motor neurons were from Foxp2-V1 and Renshaw cells. Furthermore, a higher proportion of Foxp2-V1 synapses is observed on LMC motor neurons than on axial motor neurons. They proposed that Foxp2-V1, which represents 60% of V1, can be further classified according to the expression of transcription factors Otp and Foxp4.

These results will be helpful for future analyses of the development and function of V1 neurons. In particular, the discovery of strong synaptic connections between Foxp2-V1 and LMC motor neurons will be beneficial in analyzing the role of V1 neurons in motor circuits that generate movement of the limbs.

The conclusions of this paper are well supported by the data obtained using widely used methods. However, for some analyses, the specificity of labeling V1 clades should be clearly described.

(1) In Figure 1, the MafB antibody (Sigma) was used to identify Renshaw cells at P5. However, according to the supplementary Figure 3D, the specificity of the MafB antibody (Sigma) is relatively low. The image of MafB-GFP, V1-INs, and MafB-IR at P5 should be added to the supplementary figure. The specificity of MaFB-IR-Sigma in V1 neurons at P5 should be shown. This image also might support the description of the genetically labeled MafB-V1 distribution at P5 (page 8, lines 28-32).

(2) The proportion of genetically labeled FoxP2-V1 in all V1 is more than 60%, although immunolabeled FoxP2-V1 is approximately 30% at P5. Genetically labeled Otp-V1 included other non-FoxP2 V1 clades (Fig. 8L-M). I wonder whether genetically labeled FoxP2-V1 might include the other three clades. The authors should show whether genetically labeled FoxP2-V1 expresses other clade markers, such as pou6f2, sp8, and calbindin, at P5.

Reviewer #2 (Public Review):

Summary:

This work brings important information regarding the composition of interneurons in the mammalian spinal cord, with a developmental perspective. Indeed, for the past decades, tools inspired from developmental biology have opened up promising avenues for challenging the functional heterogeneity in the spinal cord. They rely on the fact that neurons sharing similar mature properties also share a largely similar history of expression of specific transcription factor (TF) genes during embryogenic and postnatal development. For instance, neurons originating from p1 progenitors and expressing the TF Engrailed-1, form the V1 neuronal class. While such "cardinal" neuronal classes defined by one single RF indeed share numerous features - e.g., for the case of V1 neurons, a ventral positioning, an inhibitory nature and ipsilatetal projections - there is accumulating evidence for a finer-grained diversity and specialization in each class which is still largely obscure. The present work studies the heterogeneity of V1 interneurons and describes multiple classes based on their birthdate, final positioning, and expression of additional TF. It brings in particular a solid characterization of the Foxp2-expressing V1 interneurons for which authors also delve into the connectivity, and hence, possible functional implication. The work will be of interest to developmental biologists and those interested in the organization of the locomotor spinal network.

Strengths:

This study has deeply analyzed the diversity of V1 neurons by intersecting multiple criteria: TF expression, birthdate, location in the spinal cord, diversity along the rostro-caudal axis, and for some subsets, connectivity. This illustrates and exemplifies the absolute need to not consider cardinal classes, defined by one single TF, as homogeneous. Rather, it highlights the limits of single-TF classification, and exemplifies the existence of further diversity within cardinal class.

Experiments are generally well performed with a satisfactory number of animals and adequate statistical tests.

Authors have also paid strong attention to potential differences in cell-type classification when considering neurons currently expressing of a given TF (e.g., using antibodies), from those defined as having once expressed that TF (e.g., defined by a lineage-tracing strategy). This ambiguity is a frequent source of discrepancy of findings across studies.

Furthermore, there is a risk in developmental studies to overlook the fact that the spinal cord is functionally specialized rostro-caudally, and to generalize features that may only be applicable to a specific segment and hence to a specific motor pool. While motoneurons share the same dorso-ventral origin and appear homogenous on a ChAT staining, specific clusters are dedicated to specific muscle groups, e.g., axial, hypaxial or limb muscles. Here, the authors make the important distinction between different lumbar levels and detail the location and connectivity of their neurons of interest with respect to specific clusters of MN.

Finally, the authors are fully transparent on inter-animal variability in their representation and quantification. This is crucial to avoid the overgeneralization of findings but to rather provide a nuanced understanding of the complexities of spinal circuits.

Weaknesses:

The current version of the paper is VERY hard to read. It is often extremely difficult to "see the forest for the trees" and the reader is often drowned in methodological details that provide only minor additions to the scientific message. Non-specialists in developmental biology, but still interested in the spinal cord organization, especially students, might find this article challenging to digest and there is a high risk that they will be inclined to abandon reading it. The diversity of developmental stages studied (with possible mistakes between text and figures) adds a substantial complexity in the reading. It is also not clear at all why authors choose to focus on the Foxp2 V1 from page 9. Naively, the Pou6f2 might have been equally interesting. Finally, numerous discrepancies in the referencing of figures must also be fixed. I strongly recommend an in-depth streamlining and proofreading, and possibly moving some material to supplement (e.g. page 8, and elsewhere).

Second, and although the different V1 populations have been investigated in detail regarding their development and positioning, their functional ambition is not directly investigated through gain or loss of function experiments. For the Foxp2-V1, the developmental and anatomical mapping is complemented by a connectivity mapping (Fig 6s, 8), but the latter is fairly superficial compared to the former. Synapses (Fig 6) are counted on a relatively small number of motoneurons per animal, that may, or may not, be representative of the population. Likewise, putative synaptic inputs are only counted on neuronal somata. Motoneurons that lack of axono-somatic contacts may still be contacted distally. Hence, while this data is still suggestive of differences between V1 pools, it is only little predictive of function.

Third, I suggest taking with caution the rabies labelling (Figure 8). It is known that this type of Rabies vectors, when delivered from the periphery, might also label sensory afferents and their post-synaptic targets in the cord through anterograde transport and transneuronal spread (e.g., Pimpinella et al., 2022). Yet I am not sure authors have made all controls to exclude that labelled neurons, presumed here to be premotoneurons, could rather be anterogradely labelled from sensory afferents.

Fourth, the ambition to differentiate neuronal birthdate at a half-day resolution (e.g., E10 vs E10.5) is interesting but must be considered with caution. As the author explains in their methods, animals are caged at 7pm, and the plug is checked the next morning at 7 am. There is hence a potential error of 12h.

Reviewer #3 (Public Review):

Building on their previous work that defined four major subgroups, or clades, of V1 interneurons largely by their transcriptional signatures, they do meticulous yet comprehensive analysis of the birth timing of V1 interneurons by clade, and even intra-clade, subtypes. This analysis establishes new relationships between the molecular identity, settling position, and birth time with extraordinary precision.

These relationships are then explored from the lens of synaptic connectivity. Focusing on the FoxP2 clade, they show tight spatial correspondence between V1 and motor neuron position, and through detailed synaptic analysis, find the FoxP2 V1 clade, as compared to Renshaw cells and other V1s, are the major contributors to V1-to-limb motor neuron connectivity. Finally, by analyzing sensory-to-V1 connectivity too, they show that the FoxP2 clade exhibits Ia-reciprocal interneuron-like convergence of proprioceptive and Renshaw cell synapses.

Taking the development and connectivity analysis together, their work substantially advances our understanding of spinal interneurons and yields fundamental basic information about how cell type heterogeneity corresponds across developmental, molecular and anatomical features.

An additional strength of this study is that they generate new genetic tools for labeling interneuron subpopulations, and provide insider knowledge into antibody, genetic and viral labeling that often get tucked under the rug, providing a very useful resource for further studies.

My only criticism is that some of the main messages of the paper are buried in technical details. Better separation of the main conclusions of the paper, which should be kept in the main figures and text, and technical details/experimental nuances, which are essential but should be moved to the supplement, is critical. This will also correct the other issue with the text at present, which is that it is too long.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1:

Comment 1. In Figure 1, the MafB antibody (Sigma) was used to identify Renshaw cells at P5. However, according to the supplementary Figure 3D, the specificity of the MafB antibody (Sigma) is relatively low. The image of MafB-GFP, V1-INs, and MafB-IR at P5 should be added to the supplementary figure. The specificity of MaFB-IR-Sigma in V1 neurons at P5 should be shown. This image also might support the description of the genetically labeled MafB-V1 distribution at P5 (page 8, lines 28-32).

We followed the reviewer’s suggestion and moved analyses of the MafB-GFP mouse to a supplemental figure (Fig S3). The characterization of MafB immunoreactivities is now in supplemental Figure S2 and the related text in results was also moved to supplemental to reduce technicalities in the main text. We added confocal images of MafB-GFP V1 interneurons at P5 showing immunoreactivities for both MafB antibodies, as suggested by the reviewer (Fig S2A,B). We agree with the reviewer that this strengthens our comparisons on the sensitivity and specificity of the two MafB antibodies used in this study.

As explained in the preliminary response we cannot show lack of immunoreactivity for MafB antibodies in MafB GFP/GFP knockout mice at P5 because MafB global KOs die at birth. This is why we used tissues from late embryos to check MafB immunoreactivities (Figure S2C and S2D). We made this point clearer in the text and supplemental figure legends.

Comment 2. The proportion of genetically labeled FoxP2-V1 in all V1 is more than 60%, although immunolabeled FoxP2-V1 is approximately 30% at P5. Genetically labeled Otp-V1 included other nonFoxP2 V1 clades (Fig. 8L-M). I wonder whether genetically labeled FoxP2-V1 might include the other three clades. The authors should show whether genetically labeled FoxP2-V1 expresses other clade markers, such as pou6f2, sp8, and calbindin, at P5.

We included the requested data in Figure 3E-G. Lineage-labeled Foxp2-V1 neurons in our genetic intersection do not include cells from other V1-clades.

Reviewer 2:

Comment 1. The current version of the paper is VERY hard to read. It is often extremely difficult to "see the forest for the trees" and the reader is often drowned in methodological details that provide only minor additions to the scientific message. Non-specialists in developmental biology, but still interested in the spinal cord organization, especially students, might find this article challenging to digest and there is a high risk that they will be inclined to abandon reading it. The diversity of developmental stages studied (with possible mistakes between text and figures) adds a substantial complexity in the reading. It is also not clear at all why authors choose to focus on the Foxp2 V1 from page 9. Naively, the Pou6f2 might have been equally interesting. Finally, numerous discrepancies in the referencing of figures must also be fixed. I strongly recommend an in-depth streamlining and proofreading, and possibly moving some material to supplement (e.g. page 8, and elsewhere).

The whole text was re-written and streamlined with most methodological discussion (including the section referred to by the reviewer) transferred to supplemental data. Nevertheless, enough details on samples, stats and methods were retained to maintain the rigor of the manuscript.

The reasons justifying a focus on Foxp2-V1 interneurons were fully explained in our preliminary response. Briefly, we are trying to elucidate V1 heterogeneity, and prior data showed that this is the most heterogeneous V1 clade (Bikoff et al., 2016), so it makes sense it was studied further. We agree that the Pou6f2 clade is equally interesting and is in fact the subject of several ongoing studies.

Comment 2. … although the different V1 populations have been investigated in detail regarding their development and positioning, their functional ambition is not directly investigated through gain or loss of function experiments. For the Foxp2-V1, the developmental and anatomical mapping is complemented by a connectivity mapping (Fig 6s, 8), but the latter is fairly superficial compared to the former. Synapses (Fig 6) are counted on a relatively small number of motoneurons per animal, that may, or may not, be representative of the population. Likewise, putative synaptic inputs are only counted on neuronal somata. Motoneurons that lack of axo-somatic contacts may still be contacted distally. Hence, while this data is still suggestive of differences between V1 pools, it is only little predictive of function.

We fully answered the question on functional studies in the preliminary response. Briefly, we are currently conducting these studies using various mouse models that include chronic synaptic silencing using tetanus toxin, acute partial silencing using DREADDs, and acute cell deletion using diphtheria toxin. Each intervention reveals different features of Foxp2-V1 interneuron functions, and each model requires independent validation. Moreover, these studies are being carried out at three developmental stages: embryos, early postnatal period of locomotor maturation and mature animals. Obviously, this is all beyond the goals and scope of the present study. The present study is however the basis for better informed interpretations of results obtained in functional studies.

Regarding the question on synapse counts, we explained in the preliminary results fully why we believe our experimental designs for synapse counting at the confocal level are among the most thorough that can be found in the literature. We counted a very large number of motoneurons per animal when adding all motor column and segments analyzed in each animal. Statistical power was also enough to detect fundamental variation in synaptic density among motor columns.

We focus our analyses on motoneuron cells bodies because analysis of full dendritic arbors on all motor columns present throughout all lumbosacral segments is not feasible. Please see Rotterman et al., 2014 (J. of Neuroscience; doi: 10.1523/JNEUROSCI.4768-13.2014) for evaluation of what this entails for a single motoneuron. We agree with the reviewer that analyses of V1 synapses over full dendrite arbors in specific motoneurons will be very relevant in further studies. These should be carried out now that we know which motor columns are of high interest. Nevertheless, inhibitory synapses exert the most efficient modulation of neuronal firing when they are on cell bodies, and our analyses clearly suggest a difference in in cell body inhibitory synapses targeting between different V1 interneuron types that we find very relevant.

Comment 3. I suggest taking with caution the rabies labelling (Figure 8). It is known that this type of Rabies vectors, when delivered from the periphery, might also label sensory afferents and their postsynaptic targets in the cord through anterograde transport and transneuronal spread (e.g., Pimpinella et al., 2022). Yet I am not sure authors have made all controls to exclude that labelled neurons, presumed here to be premotoneurons, could rather be anterogradely labelled from sensory afferents.

Over the years, we performed many extensive controls and validation of rabies virus transsynaptic tracing methods. These were presented at two SfN meetings (Gomez-Perez et al., 2015 and 2016; Program Nos. 242.08 and 366.06). Our validation of this technique was fully explained in our preliminary response. We also pointed out that the methods used by Pimpinella et al. have a very different design and therefore their results are not comparable to ours. In this study we injected the virus at P15 into leg muscles, and not directly into the spinal cord. In our hands, and as cited in Pimpinella et al., the rabies virus loses tropism for primary afferents with age when injected in muscle. The lack of primary afferent labeling in key lumbosacral segments (L4 and L5) is now illustrated in a new supplemental figure (Figure S6). This figure also shows some starter motoneurons. As explained in the text and in our previous response, these are few in number because of the reduced infection rate when using this method in mature animals (after P10).

Comment 4. The ambition to differentiate neuronal birthdate at a half-day resolution (e.g., E10 vs E10.5) is interesting but must be considered with caution. As the author explains in their methods, animals are caged at 7pm, and the plug is checked the next morning at 7 am. There is hence a potential error of 12h.

We agree with the reviewer, and we previously explicitly discussed these temporal resolution caveats. We have now further expanded on this in new text (see middle paragraph in page 5). Nevertheless, the method did reveal the temporal sequence of neurogenesis of V1 clades with close to 12-hour resolution.

As explained in text and preliminary response this is because we analyzed a sufficient number of animals from enough litters and utilized very stringent criteria to count EdU positives.

Moreover, our results fit very well with current literature. The data agree with previous conclusions from Andreas Sagner group (Institut für Biochemie, Friedrich-Alexander-Universität Erlangen-Nürnberg), on spinal interneurons (including V1s) birthdates based on a different methodology (Delile J et al.

Development. 2019 146(12):dev173807. doi: 10.1242/dev.173807. PMID: 30846445; PMCID: PMC6602353). In the discussion we compared in detail both the data and methods between Delile article and our results. We also cite Sagner 2024 review as requested later in the reviewer’s detailed comments. Our results also confirmed our previous report on the birthdates of V1-derived Renshaw cells and Ia inhibitory interneurons (Benito-Gonzalez A, Alvarez FJ J Neurosci. 2012 32(4):1156-70. doi: 10.1523/JNEUROSCI.3630-12.2012. PMID: 22279202; PMCID: PMC3276112). Finally, we recently received a communication notifying us that our neurogenesis sequence of V1s has been replicated in a different vertebrate species by Lora Sweeney’s group (Institute of Science and Technology Austria; direct email from this lab) and we shared our data with them for comparison. This manuscript is currently close to submission. Therefore, we are confident that despite the limitations of EdU birthdating we discussed, the conclusions we offered are strong and are being validated by other groups using different methods and species. We also want to acknowledge the positive comments of reviewer 3 regarding our birthdating study, indicating it is one the most rigorous he or she has ever seen.

Reviewer 3:

Comment 1. My only criticism is that some of the main messages of the paper are buried in technical details. Better separation of the main conclusions of the paper, which should be kept in the main figures and text, and technical details/experimental nuances, which are essential but should be moved to the supplement, is critical. This will also correct the other issue with the text at present, which is that it is too long.

Similar to our response to comment 1 from Reviewer 2 we followed the reviewers’ recommendations and greatly summarized, simplified and removed technical details from the main text, trying not to decrease rigor.

Reviewer #1 (Recommendations For The Authors):

In Figure 1, the definition of the area to analyze MafB ventral and MafB dorsal is unclear. It should be described.

This has been clarified in both text and supplemental figure S3.

“We focused the analyses on the brighter dorsal and ventral MafB-V1 populations defined by boxes of 100 µm dorsoventral width at the level of the central canal (dorsal) or the ventral edge of the gray matter (ventral) (Supplemental Figure S3B).”

Problems with figure citation.

We apologize for the mistakes. All have been corrected.

Reviewer #2 (Recommendations For The Authors):

As indicated in the public review, I'd recommend to substantially revise the writing, for clarity. As such, the paper is extremely hard to read. I would also recommend justifying the focus on Foxp2 neurons.

Also, the scope of the present paper is not clearly stated in the introduction (page 4).

Done. We also modified the introduction such that the exact goals are more clearly stated.

I would also recommend toning down the interpretation that V1 clades constitute "unique functional subsets" (discussion and elsewhere). Functional investigation is not performed, and connectomic data is partial and only very suggestive.

We include the following sentence at the end of the 1st paragraph in the discussion:

“This result strengthens the conclusion that these V1 clades defined by their genetic make-up might represent distinct functional subtypes, although further validation is necessary in more functionally focused studies.”

Different post-natal stages are used for different sections of the manuscript. This is often confusing, please justify each stage. From the beginning even, why is the initial birthdating (Figure 1) done here at p5, while the previous characterization of clades was done at p0? I am not sure to understand the justification that this was chosen "to preserve expression of V1 defining TFs". Isn't the sooner the better?

The birthdating study was carried out at P5. P5 is a good time point because there is little variation in TF expression compared to P0, as demonstrated in the results. Furthermore, later tissue harvesting allows higher replicability since it is difficult to consistently harvest tissue the day a litter is born (P0). Also technically, it is easier to handle P5 tissue compared to P0. The analysis of VGUT1 synapses was also done at P5 rather than later ages. This has two advantages: TFs immunoreactivities are preserved at this age, and also corticospinal projections have not yet reached the lumbar cord reducing interpretation caveats on the origins of VGUT1 synapses in the ventral horn (although VGLUT1 synapses are still maturing at this age, see below).

Other parts of the study focus on different ages selected to be most adequate for each purpose. To best study synaptic connectivity, it is best to study mature spinal cords after synaptic plasticity of the first week. For the tracing study we thoroughly explain in the text the reasons for the experimental design (see also below in detailed comments). For counting Foxp2-V1 interneurons and comparing them to motor columns we analyze mature animals. For testing our lineage labeling we use animals of all ages to confirm the consistency of the genetic targeting strategy throughout postnatal development and into adulthood.

Figure 5: wouldn't it be worth quantifying and illustrating cellular densities, in addition to the average number of Foxp2 neurons, across lumbar segments (panel D & E)? Indeed, the size of - and hence total number of cells within - each lumbar segment might not be the same, with a significant "enlargement" from L2 to L4 (this is actually visible on the transverse sections). Hence, if the total number of cells is in the higher in these enlarged segments, but the total number of Foxp2-V1 is not, it may mean that this class is proportionally less abundant.

We believe the critical parameter is the ratio of Foxp2-V1s to motoneurons. This informs how Foxp2-V1 interneurons vary according to the size of the motor columns and the number of motoneurons overall.

The question asked by the reviewer would best be answered by estimating the proportion of Foxp2-V1 neurons to all NeuN labeled interneurons. This is because interneuron density in the spinal cord varies in different segments. We are not sure what this additional analysis will contribute to the paper.

Why, in the Rabies tracing scheme (Fig 8), the Rabies injection is performed at p15? As the authors explain in the text, rabies uptake at the neuromuscular junction is weak after p10. It is not clear to me why such experiments weren't done all at early postnatal stages, with a "classical" co-injection of TVA and Rabies.

First, we do not need TVA in this experiment because we are using B19-G coated virus and injecting it into muscles, not into the spinal cord directly.

Second, enhanced tracing occurs when the AAV is injected a few days before rabies virus. This is because AAV transgene expression is delayed with respect to rabies virus infection and replication. We have performed full time courses and presented these data in one abstract to SfN: Gomez-Perez et al., 2015 Program Nos. 242. We believe full description of these technical details is beyond the scope of this manuscript that has already been considered too technical.

Third, the justification of P15 timing of injections for anterograde primary afferent labeling and retrograde monosynaptic labeling of interneurons is fully explained in the text.

“To obtain transcomplementation of RVDG-mCherry with glycoprotein in LG motoneurons, we first injected the LG muscle with an AAV1 expressing B19-G at P4. We then performed RVDG and CTB injections at P15 to optimize muscle targeting and avoid cross-contamination of nearby muscles. Muscle specificity was confirmed post-hoc by dissection of all muscles below the knee. Analyses were done at P22, a timepoint after developmental critical windows through which Ia (VGLUT1+) synaptic numbers increase and mature on V1-IaINs (Siembab et al., 2010)”

Furthermore, CTB starts to decrease in intensity 7 days after injection because intracellular degradation and rabies virus labeling disappears because cell death. Both limit the time of postinjection for analyses.

Likewise, I am surprised not to see a single motoneuron in the rabies tracing (Fig 8, neither on histology nor on graphs (Fig 8). How can authors be certain that there was indeed rabies uptake from the muscle at this age, and that all labelled cells, presumed to be preMN, are not actually sensory neurons? It is known that Rabies vectors, when delivered from the periphery, might also label sensory afferents and their post-synaptic targets through anterograde transport and transneuronal spread (e.g., Pimpinella et al., 2022). This potential bias must be considered.

This is fully explained in our previous response to the second reviewer’s general comments. We have also added a confocal image showing starter motoneurons as requested (Figure S6A).

Please carefully inspect the references to figures and figure panels, which I suspect are not always correct.

Thank you. We carefully revised the manuscript to correct these deficiencies and we apologize for them.

Reviewer #3 (Recommendations For The Authors):

Figure 1: Data here is absolutely beautiful and provides one of the most thorough studies, in terms of timepoints, number of animals analyzed, and precision of analysis, of edU-based birth timing that has been published for neuron subtypes in the spinal cord so far. My only suggestion is to color code the early and late born populations (in for example, different shades of green for early; and blue for late, to better emphasize the differences between them). It is very difficult to differentiate between the purple, red and black colors in G-I, which this would also fix. The antibody staining for Pou6f2 (F) is also difficult to see; gain could be increased on these images or insets added for clarity.

The choice of colors is adapted for optimal visualization by people with different degrees of color blindness. Shades of individual colors are always more difficult to discriminate. This is personally verified by the senior corresponding author of this paper who has some color discrimination deficits. Moreover, each line has a different symbol for the same purpose of easing differentiation.

Figure 2: This is also a picture-perfect figure showing further diversity by birth time even within a clade. One small aesthetic comment is that the arrows are quite unclear and block the data. Perhaps the contours themselves could be subdivided by region and color coded by birth time-such that for example the dorsal contours that emerge in the MafB clade at E11 are highlighted in their own color. Some quantification of the shift in distribution as well as the relative number of neurons within each spatially localized group would also be useful. For MafB, for example, it looks as though the ventral cells (likely Renshaw) are generated at all times in the contour plots; in the dot plots however, it looks like the most ventral cells are present at e10.5. This is likely because the contours are measuring fractional representations, not absolute number. An independent measure of absolute number of ventral and dorsal, by for example, subdividing the spinal cord into dorsoventral bins, would be very useful to address this ambiguity.

We believe density plots already convey the message of the shift in positions with birthdate. We are not sure how we can quantify this more accurately than showing the differences in cellular density plots. We used dorsoventral and mediolateral binning in our first paper decades ago (Avarez et al., 2005). This has now been replaced by more rigorous density profiles that describe better cell distributions. Unfortunately, to obtain the most accurate density profiles we need to pool all cells from all animals precluding statistical comparisons. This is because for some groups there have very few cells per animal (for example early born Sp8 or Foxp2 cells).

Figure 3 and Figure 4: These, and all figures that compare the lineage trace and antibody staining, should be moved to the supplement in my opinion-as they are not for generalist readers but rather specialists that are interested in these exact tools. In addition, the majority of the text that relates to these figures should be transferred to the supplement as well. Figure 5: Another great figure that sets the stage for the analysis of FoxP2V1-to-MN synaptic connectivity, and provides basic information about the rostrocaudal distribution of this clade, by analyzing settling position by level. I have only minor comments. The grid in B obscures the view of the cells and should be removed. The motor neuron cell bodies in C would be better visible if they were red.

We moved some of the images to supplemental (see new supplemental Fig S4). However, we also added new data to the figure as requested by reviewers (Fig 3E-G). We preserved our analyses of Foxp2 and non-Foxp2 V1s across ages and spinal segments because we think this information is critical to the paper. Finally, we want to prevent misleading readers into believing that Foxp2 is a marker that is unique to V1s. Therefore, we also preserved Figures 3H to 3J showing the non-V1 Foxp2 population in the ventral horn.

Figure 6: Very careful and quantitative analysis of V1 synaptic input to motor neurons is presented here. For the reader, a summary figure (similar to B but with V1s too) that schematizes V1 FoxP2 versus Renshaw cell connectivity with LMC, MMC, and PGC motor neurons are one level would be useful.

Thanks for the suggestion. A summary figure has now been included (Figure 5G).

Figure 7: The goal of this figure is to highlight intra-clade diversity at the level of transcription factor expression (or maintenance of expression), birth timing and cell body position culminating in the clear and concise diagram presented in G. In panels A-F however, it takes extra effort to link the data shown to these I-IV subtypes. The figure should be restructured to better highlight these links. One option might be to separate the figure into four parts (one for each type): with the individual spatial, birth timing and TF data for each population extracted and presented in each individual part.

We agree with the reviewer that this is a very busy figure. We tried to re-structure the figure following the suggestions of the reviewer and also several alternative options. All resulted in designs that were more difficult to follow than the original figure. We apologize for its complexity, but we believe this is the best organization to describe all the data in the simplest form.

Figure 8: in A-D, the main point of the figure - that V1FoxP2Otp preferentially receive proprioceptive synapses is buried in a bunch of technical details. To make it easier for the reader, please:

(1) add a summary as in B of the %FoxP2-V1 Otp+ cells (82%) with Vglut1 synapses to make the point stronger that the majority of these cells have synapses.

We added this graph by extending the previous graph to include lineage labeled Foxp2-V1s with OTP or Foxp2 immunoreactivity. It is now Figure 7B.

(2) Additionally, add a representative example that shows large numbers of proximal synapses on an FoxP2-V1 Otp+.

The image we presented before as Figure 8A was already immunostained for OTP, so we just added the OTP channel to the images. Now all this information is in panels that are subparts of Figure 7A.

(3) Move the comparison between FoxP2-V1 and FoxP2AB+V1s to the supplement.

We preserved the quantitative data on Foxp2-V1 lineage cells with Foxp2-immunoreactivity but made this a standalone figure, so it is not as busy.

(4) Move J-M description of antibody versus lineage trace of Otp to supplement as ending with this confuses the main message of the paper (see comment above).

All results for the Otp-V1 mouse model have now been placed in a supplemental figure (Figure 5S).

Discussion: A more nuanced and detailed discussion of how the temporal pattern of subtype generation presented here aligns with the established temporal transcription factor code (nicely summarized in Sagner 2024) would be helpful to place their work in the broader context of the field.

This aspect of the discussion was expanded on pages 20 and 21. We replaced the earlier cited review (Sagner and Briscoe, 2019, Development) with the updated Sagner 2024 review and further discussed the data in the context of the field and neurogenesis waves throughout the neural tube, not only the spinal cord. We previously carefully compared our data with the spinal cord data from Sagner’s group (Delile et, 2019, Development). We have now further expanded this comparison in the discussion.

Version 2

Abstract

Abstract

Spinal cord interneurons play critical roles shaping motor output, but their precise identity and connectivity remain unclear. Focusing on the V1 interneuron cardinal class we defined four major V1 subsets according to neurogenesis timing, genetic lineage-tracing, synaptic output to motoneurons, and synaptic inputs from muscle afferents. Birthdate delineates two early born (Renshaw and Pou6f2) and two late born (Foxp2 and Sp8) V1 clades, showing that sequential neurogenesis produces different V1 subsets. Early born Renshaw cells and late born Foxp2-V1 interneurons are tightly coupled to motoneurons, while early born Pou6f2-V1 and late born Sp8-V1 interneurons are not, indicating that timing of neurogenesis does not correlate with motoneuron targeting. V1 clades also differ in cell numbers and diversity. Lineage labeling shows that the Foxp2-V1 clade contains over half of all V1 interneurons, provides the largest inhibitory input to motoneuron cell bodies and includes subgroups that differ in birthdate, location, and proprioceptive input. Notably, one Foxp2-V1 subgroup, defined by postnatal Otp expression is positioned near the lateral motor column and receives substantial input from proprioceptors, consistent with an involvement in reciprocal inhibitory pathways. Combined tracing of ankle flexor sensory afferents and interneurons monosynaptically connected to ankle extensors confirmed placement of Foxp2-V1 interneurons in reciprocal inhibitory pathways. Our results validate previously proposed V1 clades as unique functional subtypes that differ in circuit placement, with Foxp2-V1 cells forming the most heterogeneous subgroup. We discuss how V1 organizational diversity enables understanding of their roles in motor control, with implications for their diverse ontogenetic and phylogenetic origins.

Introduction

The spinal cord contains a diversity of interneurons which lend to its vast computational power. Many of these interneurons are inhibitory and critically modulate and pattern motoneuron firing adjusting the timing and force of muscle contractions. We focus here on one major group of ventral inhibitory interneurons known as V1 that originate from p1 progenitors, express the transcription factor (TF) Engrailed-1 (En1), and send ipsilateral axons throughout the ventral horn including lamina IX (LIX), where they densely innervate motoneuron cell bodies and proximal dendrites (Alvarez et al., 2005). This early study proposed a diversity of phenotypes and circuit roles for V1 interneurons based on differential expression of calcium buffering proteins, location and synaptology. Electrophysiological and modeling studies then showed that V1 interneurons play crucial roles shaping motor output and modulating locomotor speed, governing flexion-extension at the level of central pattern generator (CPG) half-centers and/or last-order reciprocal inhibition of antagonistic motoneurons and providing recurrent feedback inhibition of motoneuron firing (Sapir et al., 2004; Zhang et al., 2014; Britz et al., 2015; Falgairolle and O’Donovan, 2019, 2021; Shevtsova et al., 2022). However, the relations between these multiple functions and V1 heterogeneity is not fully clear. Accordingly, there is significant interest in defining the molecular identity, circuit organization and motoneuron connections of different V1 interneuron subtypes. This is of further significance because recent findings suggest that early disconnection of V1 synapses from motoneurons contributes to dysfunction and presages motoneuron death in amyotrophic lateral sclerosis (ALS) (Wootz et al., 2013; Salamatina et al., 2020; Allodi et al., 2021; Mora et al., 2024; Montaña-Rosell et al., 2024).

Prior work revealed that V1 interneurons are organized into at least four major clades according to their positions and molecular identity at postnatal day 0 (P0) (Bikoff et al., 2016). Each clade was defined by expression of unique transcription factors. Co-expression of V-maf musculoaponeurotic fibrosarcoma oncogene homologs A and B (MafA/MafB) defines the V1 clade that corresponds with Renshaw cells (Benito-Gonzalez and Alvarez, 2012; Stam et al., 2012; Bikoff et al., 2016). The other three clades are respectively defined by expression at P0 of the TFs P.O.U. domain class 6 homeobox 2 (Pou6f2), Forkhead box P2 (Foxp2) and Specificity protein 8 (Sp8) (Bikoff et al., 2016). Within these three clades, additional diversity was uncovered by further combinations of TF expression and positions (Bikoff et al., 2016; Gabitto et al., 2016; Sweeney et al., 2018). A recent harmonized atlas of several mouse spinal cord transcriptomic studies (Russ et al., 2021) identified seven possible V1 groups, including Renshaw cells, Pou6f2-V1s and three Foxp2-V1 groups. V1 genetic diversity parallels the functional diversity described in physiological and modeling studies, but whether V1 clades occupy specific functional niches in spinal motor circuits remains unclear, partly because of a lack of information about their synaptic inputs and outputs and their origins.

Here we clarify the origins, diversity, and synaptic relations with motoneurons of four major V1 clades. V1 neurogenesis was previously divided into early (E9.5 to E10.5) and late phases (E11 to E12.5), each producing distinct V1 interneurons (Benito-Gonzalez and Alvarez, 2012). It is now well accepted that temporal and spatial properties intersect during neurogenesis to create cellular diversity from each spinal cord progenitor domain (Sagner, 2024; Deska-Gauthier et al., 2020; Deska-Gauthier and Zhang, 2021; Osseward et al., 2021; Sagner et al., 2021; Deska-Gauthier et al., 2024). Using nodal intersections between transcriptomics and neurogenesis, one report described seven embryonic V1 groups defined by TFs with temporally restricted expression (Delile et al., 2019). However, because of the dynamic nature of TF expression in spinal interneurons, it is difficult to match embryonic TFs to V1 clades previously defined at P0. Using clade-defining TFs in combination with 5-ethynyl-2’-deoxyuridine (EdU) birthdating, we identified a relationship between neurogenesis timing and clade identity and uncovered additional diversity within Foxp2-V1 interneurons, the largest V1 clade. To further study Foxp2-V1 interneurons, we used intersectional genetics to lineage label their cell bodies and axons. This revealed subdivisions according to location and expression of additional TFs. We also found that Foxp2-V1s establish the highest density of V1 synapses on limb-related lateral motor column (LMC) motoneurons. They also include subgroups receiving dense proprioceptive inputs, with some having connectivity typical of reciprocal IaINs. Together with Renshaw cells, they contribute the majority of V1 synapses on motoneuron cell bodies and proximal dendrites, while synapses from other clades (Pou6f2 and Sp8) have minimal representation. We conclude that V1 clades differ in time of neurogenesis, internal heterogeneity, and synaptic targeting of motoneurons.

Results

V1 interneurons belonging to different clades have distinct timing of neurogenesis

To examine the birthdates of the different V1 clades, we lineage traced all V1 interneurons (INs) using en1cre/+::Ai9R26-lsl-tdTomato mice (sometimes intersected with foxp2flpo/+:: RCE:dual-EGFP) and pulse-labeled developing embryos by injecting pregnant females with EdU at different 12 hour intervals from E9.5 to E12.5 (Figure 1A). Pups were analyzed at P5. This age was chosen to maximize expression of V1 clade defining TFs (MafB, Pou6f2, Foxp2 and Sp8) for antibody detection. We analyzed spinal cords at E9.5 (n=3; 3 litters), E10 (n=6; 3 litters), E10.5 (n=5; 3 litters), E11 (n=5; 2 litters), E11.5 (n=6; 3 litters), E12 (n=3; 1 litter), and E12.5 (n=4; 2 litters). In all animals, we confirmed the expected ventrolateral to dorsomedial sequence of cell birthdates in the mammalian spinal cord (Figure 1B) (Altman and Bayer, 2001). One animal pulse-labeled at E11 was removed from the analyses because the pattern of EdU labeling did not correspond to the expected distribution (marked with x in Figure 1D). We defined positive neurons as those with more than two-thirds of the nucleus showing homogeneous EdU fluorescence to ensure only V1 cells in S-phase at the time of injection were included in analyses. Neurons with speckles or partial nuclear labeling (Figure 1C) may arise from EdU dilution in successive division cycles, or incorporation of EdU during very late S-phase or DNA repair events (Packard et al., 1973; Ferreira et al., 1997; Taupin, 2007).

Neurogenesis order of V1 clades assayed by EdU birthdating.

A, Experimental design. Timed pregnant En1cre-Ai9:tdT females were EdU injected at one of seven time points between E9.5 and E12.5, and the spinal cords harvested at P5. Tissue sections were processed for EdU (Click-iT™) and immunostained for representative transcription factors (TFs) of major V1 clades. MafB was used to identify the MafA-Renshaw cell clade by location (“ventral MafB-V1 cells”). B, EdU labeling at different embryonic times. Spinal neurons are born in a ventrolateral to dorsomedial sequence. C, Example of E11 EdU labeling in an En1-tdT spinal cord. EdU integrated in the DNA at the time of injection is diluted with subsequent cell divisions. To ensure we sampled V1 cells that incorporated EdU during S-phase after their final division, we only counted V1 cells with nuclei filled by EdU Click-iT™ reaction (arrows). D, Percentages of V1 cells labeled with “strong” EdU in each mouse. The x-axis indicates individual animals (“<litter number>.<animal number>”). The percentage of EdU-labeled V1 interneurons at each time point was consistent, although we also noted variability between litters and among animals within a litter. One animal (459.2) showed the wrong EdU pattern for its injection age and was discarded (indicated by an X). E, EdU birthdating reveals a peak in V1 neurogenesis around E11 (error bars = SD; each dot represents one animal). At the time points flanking the peak there is a larger amount of variability, suggesting a fast-changing pace in V1 neurogenesis. F, Representative images of TF antibody staining combined with EdU labeling to determine birthdates of defined V1 clades. The time points represented were selected according to the maximal or near-maximal generation of V1 interneurons in each clade. G-I, V1-clade neurogenesis quantification. Graphs represent the average ±SEM calculated from n = 3.9 ±0.3 mice per TF/date (not all TFs were tested in all mice). For each mouse average, we analyzed four ventral horns in Lumbar 4 or 5 segments. G, Percentage of V1s expressing each clade-specific TF labeled with EdU at each embryonic time point. Ventral MafB-V1s (Renshaw cells) and Pou6f2-V1s are mostly born before E11 (dorsal MafB-V1s are a subgroup of Pou6f2-V1s). Foxp2-V1 and Sp8-V1 interneurons have wider windows of neurogenesis, but most are generated after E11. H, Data normalized to the maximum percentage of V1s born in each group showing peak generation in each clade. Ventral MafB-V1s, Pou6f2-V1s, Foxp2-V1s and Sp8-V1s have progressively later times of peak generation. I, Cum-sum graphs of V1-clades neurogenesis. Between 50% to 68% of all neurons in each V1 clade are labeled across all ages. By E11 nearly all neurons in early-clades are generated, while fewer than half of neurons in late-clades are.

Using our stringent criteria for EdU+ cell classification, we found different percentages of V1s generated at different time points. Peak V1 neurogenesis occurred at E11 with 23.4% ±2.9 (mean±SD) of V1s incorporating EdU (Figure 1D-E). After adding together all EdU-labeled V1s at all time points, we account for 71.4% of all V1 interneurons. If we include V1 neurons with speckles or partial nuclear labeling, we overrepresent V1s by more than double (255.9%) and obscure differences in neurogenesis timing. Nevertheless, some overlap occurs among animals’ pulse-labeled in contiguous 12-hour time-points. Given the fast diffusion and elimination of EdU in mice (1.4 ±0.7 and 24.1 ±2.9 mins half-times, respectively; Cheraghali et al., 1994) there should be little overlap in EdU bioavailability with injections separated by 12 hours, but other limitations can reduce the effective time-resolution of the technique. In our mating protocol, fertilization can occur anytime within a 12-hour window (see methods). It is also common within single litters to find animals with developmental differences of 6 to 12 hours. Finally, S-phase duration can vary between 4 and 17 hours depending on embryonic stage and progenitor type (Ponti et al., 2013). All these conditions create opportunities for labeling overlaps after EdU injections spaced by 12-hour intervals. We overcame time-resolution issues by using rigorous criteria for defining EdU labeling and using large sample sizes that included animals from different pregnancies at each time point. This allowed robust birthdate estimates, even within 12-hour intervals. Inter-animal variability was most noted at times around peak neurogenesis (E10.5 and E11.5, Figure 1E). This is best explained by rapid acceleration and deceleration of V1 generation at these times.

To analyze birthdates of V1 interneurons identified by clade-specific markers (Figure 1F-I), we first compared the number of V1s expressing clade-defining TFs at P5 with previous estimates at P0 (Bikoff et al., 2016). Antibody specificities are shown in Supplemental Figures S1 and S2, and in a previous paper (Bikoff et al., 2016). Overall, V1s expressing MafB at P5 were 9.4% ±1.6 (±S.D.) of the whole V1 population (n=21 animals) which differs from the 25% estimate at P0 from Bikoff et al (2016). Smaller differences were found for Pou6f2+ V1s (8.1% ±4.6 (n=19) compared to 13%) Foxp2+ V1s (32.5% ±8.4 (n=26) compared to 34%) and Sp8+ V1s (8.8% ±2.6 (n=26) compared to 13%). These small differences are attributable to differences in age and/or immunocytochemical (ICC) sensitivity. In contrast, the large differences found in multi-clade MafB-V1s likely result from rapid downregulation of MafB from P0 to P5 in some V1 clades. MafB is expressed at P0 in three V1 clades (Bikoff et al., 2016): ventral MafA-calbindin Renshaw cells, subpopulations of dorsally-located Pou6f2-V1s, and subgroups of Foxp2-V1s distributed throughout the ventral horn. MafB is quickly downregulated after birth in Foxp2-V1s, but expression is maintained in Pou6f2-V1s and ventral Renshaw cells (see below). When considering only ventral MafB-V1s located in the Renshaw area (Renshaw V1 clade), we obtained identical percentages at P5 and P0 for the MafA-Renshaw cell V1 clade (5.5% ±1.7, n = 21 vs 5.0% (Bikoff et al., 2016)). We conclude that TF ICC detection at P5 provides an accurate sampling of V1 clades previously defined at P0 (Bikoff et al., 2016).

Birthdating divided V1 clades into two groups: early and late born. Most early born V1 cells are EdU-labeled before E11 and include Renshaw cells and Pou6f2-V1s. Most cells from the two other clades (Foxp2-V1s and Sp8-V1s) are late born (after E11) (Figures 1F-I). These data derive from 68,562 V1s sampled from 29 animals in 4 ventral horns per animal/TF/EdU time point (average 197.0 V1s per ventral horn). V1 clades differ in birthdate peak times and temporal spread. Ventral MafB-V1 Renshaw cells, Pou6f2-V1s, and dorsal MafB-V1s display early peaks (at or before E11) and narrow spreads in their birthdates. Foxp2-V1s and Sp8-V1s show later peaks and larger spreads (Figure 1G-H). Normalizing to their peaks, reveals that ventral MafB-V1 Renshaw cells are generated the earliest (E10.5 peak), followed by Pou6f2-V1s including dorsal MafB-V1s (E11.0 peak), Foxp2-V1s (E11.5 peak) and Sp8-V1s (E12 peak). Around 50% of all V1s are born by E11, including most cells in early born clades and a proportion of Foxp2-V1s and Sp8-V1s (Figure 1I). Less than 10% of Foxp2-V1s and Sp8-V1s are generated before E10.5, and almost no ventral MafB-V1s or Pou6f2-V1s (including the dorsal MafB-V1s) are generated after E11 (Figure 1I). After E12, only Sp8-V1s are generated. Cumulative graphs show that EdU labeling accounted for 67.3% of ventral (Renshaw) MafB-V1s, 62.4% of Pou6f2-V1s, 68.2% of dorsal MafB-V1s, 49.8% of Foxp2-V1s and 55.8% of Sp8-V1s. This suggests that our analyses represent half or more of the cells in each of the clades being this and adequate representative sample to derive differences in birthdate based on EdU pulse-labeling at 7 discrete time points during a 96 hour neurogenesis period.

Next, we examined whether clade-specific V1 interneurons with different birthdates settle in different locations by plotting the positions of V1 interneurons from each clade born at different embryonic times (Figure 2). We analyzed E10 to E12 because few V1s incorporated EdU at E9.5 and E12.5 and this diminishes the accuracy of cellular density plot representations. The earliest-born MafB-V1s (E10) accumulate ventrally in the “Renshaw cell area”. Later-born MafB-V1s (E10.5 and E11) are a subgroup of Pou6f2-V1s and occupy dorsal positions. Pou6f2-V1s, overall, are born between E10 to E11 and settle dorsally. In contrast, Foxp2-V1 generation spans the entire V1 neurogenesis period and shows differences in location according to birthdate, following a clockwise rotation in positioning with time: Foxp2-V1s born at E10 occupy dorsal locations while those born between E10.5 and E11.5 settle adjacent to the lateral motor column (LMC). Among these, there is a dorsal to ventral progression of Foxp2-V1 cells born from E10.5 to E11. Finaly, the few Foxp2-V1s born at E12 are located ventromedially. Sp8-V1 neurons born at different times also show differences in location. Early born Sp8-V1 cells are ventrally located, while later born Sp8-V1 cells are located more dorsally and medially.

V1 interneurons born at different embryonic times settle in different locations.

Top row in each panel: blue open dots indicate all V1 interneurons positive for each TF at P5, and filled red dots indicate those with strong EdU. Each plot is from one representative animal injected at each of the indicated embryonic times (four ventral horns from each animal are superimposed in the diagram). Bottom rows: cellular density profiles of V1 interneurons positive for each TF and with strong EdU. Cellular density profiles represent all cells sampled in all animals EdU-injected at the indicated embryonic times; n indicates the number of animals in each plot (4 ventral horns analyzed per animal). Contour plots are derived from 2D kernel density estimates of interneuron positions; lines encompass 10% increments. Only timepoints with representative numbers of TF+/EdU+ cells are shown. Blue arrows highlight major directional change in settling locations for V1s born at each time point with respect to the previous time point. A, MafB-V1s are divided into two groups: ventral Renshaw cell MafB-V1s and dorsal MafB-Pou6f2-V1s with different birthdates. B, Pou6f2-V1s (taken as a whole) are born between E10 and E11.5 and always settle dorsally. C, Foxp2-V1s show variations in location according to birthdate. Cells of increasingly older birthdates settle dorsally, laterally, ventro-laterally and then ventro-medially. D, A few Sp8-V1s are born early and are located ventrally, but the majority are born later and settle dorsally. Directional changes in settling positions according to birthdate differ among V1 clades.

In summary, different V1 clades are generated through overlapping windows of neurogenesis, but with distinct peaks which allow classification into early (Renshaw cells and Pou6f2-V1s) and late born clades (most of Foxp2-V1s and Sp8-V1s). Within clades, cells with different birthdates settle at specific positions, perhaps suggesting cellular and functional heterogeneity. This is most evident for the large Foxp2-V1 clade. Birthdate-related settling positions do not follow the general rule of ventrolateral to dorsomedial positions with neurogenesis progression that is evident when considering all spinal neurons. This suggests specific migratory and settling behaviors for specific V1 subgroups, as shown earlier for Renshaw cells and these being predictive of cell-specific connectivity (Benito-Gonzalez and Alvarez 2012).

MafB-V1 genetic labeling reveals two main types in the mature spinal cord

We used two MafB antibodies to identify the Renshaw cell clade because we were unable to label enough Renshaw cells with MafA antibodies at P5. The two different MafB antibodies differ in sensitivity and specificity of labeling (Supplemental Figure S2), but both consistently labeled ventral MafB-V1 Renshaw cells and dorsal MafB-Pou6f2-V1 cells. For most studies we used the antibody that best labeled Renshaw cells, although this was also the less specific antibody. As explained in Supplemental Figure S2, the better labeling with this antibody might be due to cross reactivity with MafA, a related TF specifically expressed also in Renshaw cells. Neither MafB antibody labeled V1 cells after P5. To further confirm the identify of these cells and study them in older animals, we used a genetic detection strategy by introducing a mafbGFP reporter allele (Moriguchi et al., 2006) in en1::cre, Ai9 R26-lsl-tdT mice (Supplemental Figure S3). In these animals, GFP is present in many spinal neurons and microglia. Neurons belonging to the V1 group are easily identified by tdT labeling. The two main groups of MafB-V1 interneurons, ventral calbindin-immunoreactive (-IR) Renshaw cells and dorsal Pou6f2-V1s, are easily identified at both P5 and P15 (Supplemental Figures S2A,B and S3A). At P5, weak GFP is also visible in a few V1 cells located in the middle of the ventral horn that likely belong to the Foxp2-V1 clade; mafb gene expression in Foxp2-V1s cells is probably downregulating at P5 because fluorescence is weak, shows high inter-animal variability, and it is absent at P15. We focused the analyses on the brighter dorsal and ventral MafB-V1 populations defined by boxes of 100 μm dorsoventral width at the level of the central canal (dorsal) or the ventral edge of the gray matter (ventral) (Supplemental Figure S3B). MafB-V1s in these two regions together constitute 13.2% ±2.8 (mean±SD) of all V1s at P5 (n=17 mice) with similar representation in dorsal (7.1% ±2.4) and ventral (6.0% ±1.2) groups (Supplemental Figure S3C). These percentages did not change at P15. At P5, 54.2% ±19.4 of dorsal and 70.5% ±10.1 of ventral MafB-V1s had detectable MafB-IR (Supplemental Figure S3D, n = 9 mice tested). No MafB protein was detected by P15. However, mafb gene expression is present in adult animals, as reported by GFP (after antibody amplification). Since MafB protein is undetectable this suggests post-translational regulation of MafB protein translation and/or stability.

Once we defined the two main V1 populations detected by expression of mafb in postnatal spinal cords, we tested V1 clade immunomarkers at P5 (Supplemental Figure S3E). Dorsal MafB-V1s expressed Pou6f2 (41.1% ±7.4, n=5 mice tested), almost no calbindin (1.5% ±1.8, n=4 mice), and no Foxp2 or Sp8 (0%, n=5 mice). Most ventral MafB-V1s expressed calbindin-IR (79.3% ±8.8, n=4 mice) while expression of other clade markers was negligible (Pou6f2, 0%; Foxp2, 0.4% ±0.9; Sp8. 0.7% ±1.5, n=5 mice) (Supplemental Figure S3E). Using this genetic model, we then confirmed birthdates of dorsal (Pou6f2) and ventral (Renshaw) MafB-V1s (Supplemental Figure S3F, n=2 mice per EdU injection date). The data mostly agreed with birthdate estimates for dorsal and ventral MafB-immunoreactive V1 neurons, except that we found more ventral MafB-V1s born at E11 using genetic labeling compared to antibody staining. But this difference did not reach statistical significance. In conclusion, we confirmed both MafB-V1 populations and their early birth dates using a genetic model that provided robust and long-lasting labeling in mature spinal cords. These two V1 classes constitute discreet populations; however, while the connectivity of Renshaw cells is well known, the significance of adult dorsal Pou6f2-MafB-V1 interneurons remains unexplored.

Genetic labeling of the Foxp2-V1 lineage reveals twice the number of neurons relative to postnatal Foxp2 expression

To label the Foxp2-V1 lineage independent of the developmental regulation of Foxp2 expression, we utilized an en1 and foxp2 intersection by generating en1::cre, foxp2::flpo, R26 RCE:dualEGFP/Ai9-lsl-tdT mice. In these mice we expected that neurons expressing en1 and not foxp2 would be labeled with tdT from the R26-Ai9 cre reporter, and that neurons that expressed both en1 and foxp2 would be labeled with EGFP from the R26-RCE:dualEGFP and not with tdT because the Ai9 cassette is flanked by FRT sites removed by Flpo. We confirmed tdT or GFP labeling in most V1 interneurons, but a few expressed both fluorescent proteins (Figure 3A). We believe this is the consequence of Flpo recombination inefficiencies in the larger FRET-flanked Ai9 cassette compared to the much shorter FRET-stop signal in the RCE:dualEGFP. We thus interpret “yellow” V1 neurons as cells that express the foxp2 gene either for a short period of time, weakly, or both. For mapping purposes, we included “yellow” cells into the Foxp2-V1 clade since we used only the RCE:dualEGFP reporter in follow-up analyses. The whole population of Foxp2-V1 interneurons (green+yellow) cells represents between 59% to 66% of all V1s (Supplementary Figure 4C). The distribution of Foxp2-V1s (green) and non-Foxp2 V1s (red, not green) overlapped in the ventral horn (Figure 3B) and their proportions were constant across postnatal ages (n=6 ventral horns from 1 mouse at P0, P15 and adult and 36 ventral horns from 6 mice at P5; Figure 3C) and spinal segments from lower thoracic to sacral level (2 mice at P15 analyzed in n=6 ventral horns per segment and mouse, Figure 3D). In each mouse, 51 and 52% of V1 cells were GFP labeled, 34% and 39% were tdT labeled, and 8% and 14% had both labels. These data suggest a consistent population through different ages with no new expression of Foxp2 in V1 cells after birth. However, previous data based on protein detection at P0 suggest that only 32% to 34% of V1s express Foxp2-IR (Bikoff et al., 2016). Correspondingly, only around half of genetically identified Foxp2-V1 cells were Foxp2 immunolabeled at P5 in dual color R26 RCE:dualEGFP/Ai9-lsl-tdT mice (Figure 3F, n=2 mice, 3 ventral horns each) or in single color R26 RCE:dualEGFP mice (Supplementary Figure 4C, n=3 mice, 4 ventral horns each). The locations of lineage labeled Foxp2-V1 cells with and without Foxp2 immunoreactivity at P5 overlapped (Supplementary Figure 4B). Lineage labeled Foxp2-V1 cells did not express markers of other V1 clades, suggesting they uniquely represent the Foxp2-V1 clade. In contrast, subpopulations of tdT-only V1 cells expressed all other clade specific TFs (Pou6f2, MafB and Sp8) while consistently lacking Foxp2 immunoreactivity (Figure 3G).

Genetic mouse models to label the Foxp2-V1 lineage.

A, P5 mouse intersection of En1::Cre and Foxp2::Flpo with two reporter alleles (Ai9-tdTomato and RCE:dualEGFP). Foxp2-V1s express EGFP (green) and non-Foxp2-V1s tdTomato (red). A few “yellow cells” correspond with Foxp2-V1 neurons that failed to remove the tdTomato Ai9 reporter. B, Density contours demonstrate high spatial overlap between V1 neurons expressing Foxp2 (EGFP+) or not (tdT+ and EGFP-) (n=6 ventral horns). C, Expression of lineage labels are stable throughout postnatal development, suggesting no additional foxp2 gene upregulation in V1s after P0 (n=6 ventral horns per age in one animal, except for P5 in which 6 animals and 36 ventral horns are included; error bars show SEM). D, Lineage labeling in P15 mice is uniform across all spinal cord segments from thoracic 13 to sacral 1 (n = 6 ventral horns in each segment from 2 mice; error bars show SEM). E, Foxp2-V1 (EGFP) and non-Foxp2 V1 (tdT) lineage labeling with antibody staining for the transcription factors defining the four major V1 clades: Foxp2, Pou6f2, MafB, and Sp8. The Foxp2-V1 lineage contains all P5 Foxp2-expressing V1s and excludes almost all those expressing the markers of other clades. F, Percentages of Foxp2-V1s (EGFP) expressing the clade markers at P5. Around half of the Foxp2-V1s maintain expression of the Foxp2 protein at P5, and a minimal number of these cells express transcription factors definitive of other V1 clades. G, Percentages of non-Foxp2 lineage V1s (tdT) expressing the different V1 clade markers. Cells outside of the Foxp2-V1 lineage do not express Foxp2 at P5, and this subset contain V1s from the three other clades (for both plots n = 6.5 ± 2.6 mice, 4 ventral horns each; error bars show SD). H, FLTG reporter mice reveal lineage-labeled non-V1 Foxp2 cells in the spinal cord. EGFP is expressed in Foxp2-V1s and tdTomato in non-V1 Foxp2 cells. The zoomed images of the highlighted region with and without NeuN-IR demonstrate that non-V1 Foxp2 cells include non-neuronal cell types with astrocyte morphologies. Only neurons (NeuN-IR) are included in the cell density contour plots. non-V1 Foxp2 neurons are located in the medial ventral horn and a few in the deep dorsal horn (n = 2 animals at P5, 6 ventral horns in each). I, Most Foxp2-neurons (red, non-V1s or green, V1s) are in the ventral horn and their number in 30 μm thick L4-5 sections decreases with age as the neuropil matures and expands in size. Dorsal horn Foxp2-neurons maintain their numbers despite the growth of the spinal cord, suggesting de novo postnatal Foxp2 expression in this population. Each point is one animal analyzed through 6 ventral horns (errors bars are SD). J, Percentage of Foxp2-lineage-labeled cells that are Foxp2-V1s remains constant throughout postnatal development (n=2 mice, 6 ventral horns each; error bars show SD).

Non-V1 Foxp2 cells distribute to separate regions of the spinal cord

The Foxp2 protein is also expressed in non-V1 cells (Figure 1F and Supplementary Figure 4A). To study this population we used the RC-FLTG reporter mice carrying a dual-conditional allele with a FRT-flanked stop and loxP-flanked tdT::STOP preventing transcription of EGFP. Therefore in en1::Cre, foxp2::Flpo mice, cells that express only foxp2 are labeled with tdT, while en1 expression (Foxp2-V1 clade) results in EGFP fluorescence and removal of tdT. We analyzed two mice at P5 and two at P10, and the sections were counterstained with NeuN antibodies for neuronal confirmation (Figure 3H). We found several non-overlapping populations of non-V1 Foxp2 cells. Most are located medially in the ventral horn and can be either neurons or astrocytes (by morphology and lack of NeuN). In addition, cells in the central canal and spinal cord midline were strongly labeled. This distribution suggests Foxp2 is transiently expressed in some progenitors that are different from p1 (no EGFP astrocytes are present). A few non-V1 neurons (NeuN+) are in the deep dorsal horn. The number of Foxp2 neurons per section diminished in the ventral horn from P5 to P20. This is expected because of the decrease in cellular density that occurs as the spinal cord grows and matures. In contrast, dorsal horn Foxp2 neurons slightly increase in number (Figure 3I), suggesting some might upregulate Foxp2 postnatally. In summary, there are at least three broad classes of Foxp2 neurons in the spinal cord: 1) medioventral non-V1 neurons that express Foxp2 postnatally and/or at the progenitor stage, sharing labeling with glial cells; 2) dorsal horn non-V1 neurons in which Foxp2 expression increases during postnatal development; 3) V1 neurons that upregulate Foxp2 in embryo and then remain a stable population postnatally (see also Benito-Gonzalez and Alvarez, 2012). Overall, lineage labeled Foxp2-V1 interneurons comprise 55.7% ±4.7 (mean±SD) of all spinal cord Foxp2+ neurons at P5 (n=4 mice, 2 P5 and 2 P20, Figure 3J).

Foxp2-V1 neurons from lower thoracic to sacral segments follow motoneuron numbers in a 2:1 or 3:1 ratio

To gain insights into possible functions of Foxp2-V1 interneurons, we analyzed their localization and numbers across spinal cord segments in the lumbosacral region that govern the lower body and hindlimbs. We examined segments Th13 to S1 in P20 mice (n=5) expressing EGFP in Foxp2-V1 interneurons. This was combined with choline acetyltransferase (ChAT) immunoreactivity to identify the motor pools. Motoneurons were defined as Chat-IR neurons located in lamina IX (Figure 4A). Spinal segments were identified by the distribution and size of the somatic lateral, hypaxial, and medial motor columns (LMC, HMC and MMC), and presence of autonomic sympathetic (Th13-L2) or parasympathetic (S1) neurons. We did not attempt to distinguish Lumbar 4 from 5 because of their similarity. From these sections, we constructed cell plots for each animal (4 ventral horns per animal/segment, Figure 4B) and transformed these into density plots (Figure 4C) by combining all cell plots from all animals analyzed in each segment (n=3-5 mice depending on segment). Foxp2-V1 interneurons are located throughout the ventral horn, but many accumulate laterally. In segments where the LMC expands, Foxp2-V1 interneurons border the LMC medially. Contour density plots indicate that the highest density of Foxp2-V1 interneurons lies adjacent to the LMC, suggesting a close relationship between Foxp2-V1 interneurons and the motoneurons that control limb musculature. This is consistent in segments where the LMC emerges (L2), disappears (L6), or reaches its maximal size (L4/5). Correspondingly, Foxp2-V1 neuron numbers significantly increase in segments innervating limb muscles (L3-L6) compared to segments involved with axial (Th13, S1) and hypaxial muscles (L1) (p<0.0001, one-way ANOVA followed by post-hoc Bonferroni t-tests summarized in Figure 4D and Supplement Table S1). The limb-innervating LMC is responsible for most of the change in motoneuron numbers across different spinal cord segments. Consistent with the parallel increase in number of V1 interneurons and motoneurons, the ratio of Foxp2-V1 interneurons to motoneurons remained relatively constant from Th13 to L5 (Figure 4E). Differences between Th13 and L6 are non-significant (post-hoc Bonferroni t-tests, details in Supplemental Table S1). The larger ratio at S1 was significant compared to L3 and L4/5 (see Supplemental Table S1 for details) but the estimated ratios at S1 were highly variable in the three animals studied. This is probably best explained by the rapid diminution of motoneuron numbers in S1 depending on the exact section level. Cell plots in segments lacking LMC limb motoneurons show that most Foxp2-V1 interneurons are located dorsally and distal to the motor pools, with lower density close to MMC motoneurons. Foxp2-V1 interneurons located further away from the motor pools might have roles other than the direct modulation of motoneuron firing. Finally, a sparse group of Foxp2-V1 interneurons is dispersed in the medial ventral horn in all segments. They correspond to the latest born subgroup (and they have a different genetic make-up, see below), suggesting a unique identity.

Foxp2-V1 interneurons are closely associated with shifting motor columns throughout thoracic, lumbar and sacral levels of the spinal cord.

A, Foxp2-V1 lineage labeling and ChAT antibody staining for motoneuron identification in adult mouse spinal cords from thoracic to sacral levels. Foxp2-V1 neurons accumulate at the lateral edge of the ventral spinal cord but their locations shift when the LMC expands from L3 to L6. In addition, a distinct group of Foxp2-V1 interneurons is dispersed at ventro-medial locations in lumbar segments. B, Plots of Foxp2-V1 and ChAT-IR motoneuron cell body positions in x,y coordinates with 0,0 at the top of the central canal (n=4 ventral horns, 1 representative animal). C, Contour plots of kernel Foxp2-V1 cell density estimations. The highest density of Foxp2-V1 neurons cluster close to LMC motoneurons from L2 to L5 (contours enclose 10% increments, closer lines indicate steeper changes in density). Motoneuron numbers progressively increase from Th13 to L5 and drop in number in L6 and S1. D, Number of Foxp2-V1s per 50 μm thick section (ventral horn) significantly increases in lower lumbar segments from L3 to L6 compared to S1 (dots represent individual mice; n = 3-5 mice in different segments, each mouse estimate is from 6 ventral horns; bars show SEM). One-way ANOVA, post-hoc Bonferroni corrected t-tests are summarized in the right-hand table. See Table S1 for details). E, Ratios of Foxp2-V1 neurons to MNs remain constant at roughly 2.5:1 with no significant changes throughout the lumbar cord. Significance was only found for L3-L5 compared to S1 (*p<0.05; **p<0.01; post-hoc Bonferroni tests). High variability in S1 is likely due to the sharp rostro-caudal decrease in motoneuron numbers in S1.

Foxp2-V1 synapses target LMC, HMC, and MMC motoneurons, but not pre-ganglionic autonomic neurons

Next, we examined Foxp2-V1 synapses on motoneurons and compared them to other V1 groups at P20: a time point after the critical window of synapse proliferation and pruning for V1 interneurons (Mentis et al., 2006; Siembab et al., 2010). We examined three mice in which Foxp2-V1 axons express EGFP, and non-Foxp2-V1 axons tdT. Yellow axons were included in the EGFP/Foxp2+ group. We also analyzed two mice in which all V1 axons express tdT through the Ai9 reporter. In these animals we immunostained axons with calbindin antibodies to identify synapses from Renshaw cells. Sections were further immunostained using antibodies against ChAT (to identify motoneuron cell bodies in lamina IX) and against the vesicular GABA/glycine amino acid transporter (VGAT) to reveal presynaptic vesicle accumulations in genetically labeled axon varicosities. Motoneurons were sampled in different motor columns from Th13 to S1 segments (Figures 5A-B) and examined for synaptic contacts from Foxp2-V1, non-Foxp2-V1, all V1, and Renshaw-V1 axons (Figures 5C1 and 5C2). We used rigorous criteria to estimate synaptic contact densities on 3D reconstructions of cell body surfaces (Figure 5C3). We calculated overall V1 synaptic densities (green plus red axons in EGFP/tdT dual color mice, and all red axons in tdT single color mice) on motoneurons from the following motor columns and segments: HMC motoneurons in the Th13 segment (ventral body musculature); LMC motoneurons in L1/2 (hip flexors), L4/5 (divided into dorsal and ventral pools, innervating distal and proximal leg muscles, respectively) and the dorsal L6 pool (intrinsic foot muscles); MMC neurons in segments Th13, L1/2, L3/4 and S1 (innervating axial trunk musculature and the tail at sacral levels); and finally, preganglionic autonomic cells (PGC), sympathetic at Th13 and L1/2 and parasympathetic at S1. We analyzed 4 to 9 motoneurons per animal. Initially, we kept the data separated by mouse identity to check for possible differences due to mouse and/or genetics (Figure 5D). A mixed-effects nested ANOVA revealed significant differences in V1 synapse density over different types of motoneurons (p<0.0001), but no influence of mouse or genetics (statistics details in Supplemental Table S2 and Figure 5D table). Post-hoc Bonferroni t-tests demonstrated that HMC and lower lumbar LMC motoneurons receive significantly more V1 synapses than MMC motoneurons, while LMC motoneurons in L1/2 and L6 had V1 synaptic densities not significantly different to MMC motoneurons. PGC neurons received very low densities of V1 input, significantly lower than LMC or MMC motoneurons.

Limb and axial motoneurons are densely innervated by Foxp2-V1s and Renshaw cells.

A, Motor column identification from lower thoracic to upper sacral spinal cord in P20 mice following labeling with ChAT antibodies: PGC = preganglionic cell column; MMC = medial motor column; HMC = hypaxial motor column; LMCd/v = lateral motor column (dorsal/ventral). B, Schematic representation of the rostro-caudal span of each motor column in the spinal segments studied. C, Synapse quantification. Axons of Foxp2 and non-Foxp2 V1 interneurons were respectively labeled with EGFP and tdT in en1Cre::foxp2flpo::R26 RCE:dualGFP/Ai9tdT mice. In en1Cre::R26 Ai9tdT mice we identified V1-Renshaw cell axons using calbindin antibodies. Synaptic locations were labeled with VGAT antibodies and the postsynaptic motoneurons with ChAT antibodies. Synapse densities were analyzed in a ribbon of membrane at mid-cell body level (7 optical planes, 1 μm z-step). C1, Single optical plane of a L4/5 LMCv motoneuron surrounded by genetically labeled Foxp2-V1 and non-Foxp2-V1 axons. Inhibitory synapses on ChAT-IR motoneurons are VGAT+. C2, Single optical image of a L4/5 LMCv ChAT-IR motoneuron receiving synapses from V1 Renshaw cells (genetically labeled V1 axons with calbindin-IR and VGAT). C3-4, Method for estimating synapse densities on motoneuron cell bodies using C2 as example. C3, V1-VGAT (red arrowheads) and V1-CB-VGAT synapses (yellow arrowhead) are marked (VGAT-IR is not shown for clarity), and the cell body contour annotated with regions corresponding to dendrite exits. This process was repeated in 7 consecutive mid-cell body optical planes (cross-sections with well-defined nucleus and nucleolus). C4, A membrane surface slab is reconstructed in 3D (two different rotations shown). The surface area corresponding to dendrite exits is subtracted from the total surface area of the slab to calculate the available surface area on the motoneuron cell body. V1-VGAT synapses (red), V1-CB-VGAT synapses (yellow), and CB-VGAT synapses (green) are marked. A similar process was followed for calculating Foxp2-V1 synapse density. D, Quantification of total V1-VGAT synapse densities on motoneuron cell bodies in different motor columns (n = 21-30 motoneurons per motor column, n = 5 animals with 4-9 motoneurons per animal per motor column). Each data point is one motoneuron color coded by mouse origin. Average synaptic densities ±SD indicated to the right of scatter plots. A nested ANOVA found significant differences among motor column/segments (p < 0.0001) with no inter-animal variability (p = 0.4768). The table summarizes all post-hoc pairwise comparisons for average V1 synaptic densities of each motor column and segment (Bonferroni corrected t-tests) (Statistical details are in Table S2). Colors indicate increased (>1, red) or decreased ratios (<1, blue) of column motoneurons vs row motoneurons. PGC neurons receive significantly fewer V1 synapses than MMC or LMC motoneurons. The LMC (ventral and dorsal) in lower lumbar (L4/L5) had significantly more V1 contacts than MMC motoneurons or L6 dorsal LMC. E, Comparison of synaptic densities from Foxp2-V1 and non-Foxp2-V1 neurons (top) or Renshaw cells (bottom). All motoneurons sampled in 2 to 3 animals for each comparison were pooled together. Densities of V1-VGAT synapses from Foxp2-V1s, non-Foxp2 V1s, or calbindin (CB)+ V1s (Renshaw cells) (n = 6-17 motoneurons sampled per motor column/segments, average = 12.1 ±2.9 SD) were compared using a two-way ANOVA for axon type vs motor column and segment. Foxp2-V1 vs non-Foxp2-V1 synapses: significant differences in density were found for type of synapse (p=0.001), motor column location (p<0.0001), and their interaction (p<0.0001). Significant differences after post-hoc Bonferroni tests are indicated (*p<0.05; ****p<0.0001). In general, synapses from Foxp2-V1 axons have higher density than non-Foxp2-V1 axons on HMC and LMC columns at all spinal segments except for L1/L2 LMC. MMC motoneurons receive similar synaptic densities from both types of V1 axons, except at the sacral level in which non-Foxp2 V1 synapses predominate. PGC neurons receive very low densities of V1 axons and there are no significant differences between either type in any region. Foxp2-V1 vs CB+V1 synapses: significant density differences were found for type of synapse (p<0.0001), motor column location (p<0.0001), and their interaction (p<0.0001). Significant differences between Foxp2-V1 and CB+V1 synapses after post-hoc Bonferroni tests are indicated (*p<0.05; ****p<0.0001). Synapses from Foxp2-V1 axons have higher density than CB+V1 axons in HMC and LMC columns at all spinal segments except for L1/L2 LMC. MMC motoneurons receive similar synaptic densities from both types of V1 axons in upper lumbar regions, but Foxp2-V1 synapse predominate in lower lumbar. In S1 the density of CB+/V1 synapses is significantly higher. The low synaptic densities estimated in PGC neurons for Foxp2-V1s and CB+ V1s are not significantly different. Details of all statistical comparisons are Supplementary tables S3 and S4. F, Comparing the numbers of Foxp2 and CB+ (Renshaw) V1 synapses to the total number of V1 synapses, we estimated their respective percentages. From these estimates we calculated that the remainder belongs to non-Foxp2 and non-CB+ Renshaw cells. The large majority of V1 synapses on the cell bodies of LMC, HMC and MMC motoneurons are either from Renshaw cells or Foxp2-V1s. Asterisks denote significant differences as found in E. G, Summary diagram of major V1 clade connectivity to motoneuron cell bodies. Foxp2-V1s and Renshaw cells form the majority of inhibitory V1 contacts on LMC and HMC motoneurons, with slightly higher density from Foxp2-V1s. The MMC receives roughly equal portions of V1 contacts from Foxp2-V1s and Renshaw cells. V1s provide only sparse inhibition on preganglionic sympathetic neurons and most originate in V1 clades other than Renshaw cells and Foxp2-V1s.

Next, we examined possible differences between Foxp2-V1 and non-Foxp2-V1 neurons in motoneuron targeting (Figure 5E, top graph). In this case we pooled all motoneurons from 3 mice (n=6 to 16 motoneurons per motor column/segment) and found significant differences according to motoneuron identity (p<0.0001), type of V1 axon (p=0.0107), and their interaction (p<0.0001) (two-way ANOVA, statistics details in supplemental Table S3). This was followed by pair-wise comparisons of synaptic densities according to the type of V1 axon for each motoneuron type (Figure 5E, top graph). HMC and LMC motoneurons receive significantly higher synaptic densities of Foxp2-V1 axons compared to non-Foxp2-V1 axons in all segments, except for L1/2. The synaptic densities of both types of axons are not significantly different in MMC motoneurons, except for S1 MMC motoneurons that received significantly higher density of non-Foxp2-V1 synapses. We conclude that motoneurons controlling the hindlimb receive more synapses from Foxp2-V1 interneurons, while Foxp2-V1 and non-Foxp2-V1 interneurons equally contact cell bodies of motoneurons controlling axial musculature. Synapses on PGC neurons were always at low density and highly variable, with most originating from non-Foxp2 V1 interneurons.

We then compared synapse densities from Foxp2-V1 interneurons to those from Renshaw cells. Renshaw cell axons were identified by the presence of calbindin-IR in tdT V1 axons (in this analysis we pooled 10 to 17 motoneurons from two mice). Like above, we found significant differences according to motoneuron identity (p<0.0001), type of axon (p<0.0001) and their interaction (p<0.0001) (two-way ANOVA, statistics details in supplemental Table S4). Post-hoc pairwise comparisons revealed that Renshaw cell synapses occurred at significantly lower densities compared to Foxp2-V1s in all LMC motor groups, except for L1/2 (Figure 5E, lower graph). MMC motoneurons showed similar densities of Renshaw cell and Foxp2-V1 synapses in Th13 and L1/2, higher density of Foxp2-V1 synapses in L3/4, and much higher density of calbindin+ V1 synapses in S1. The identity of calbindin+ V1 axons in S1 is unclear because of the higher numbers of non-Renshaw calbindin-IR V1 interneurons in sacral segments (i.e., calbindin+ V1 interneurons not contacted by motor axons). Further confirmation of Renshaw cell identity for calbindin+ V1s in S1 is required. PGC neurons received almost no calbindin+ V1 axons (Figure 5E, bottom graph).

After calculating synaptic densities originating from Foxp2-V1 and Renshaw cell axons, we estimated the remaining V1 synapses. Plotting the percent contributions of three V1 synapse categories to all V1 synapses shows that most synapses on the cell bodies of HMC, LMC, and MMC motoneurons originate from either Foxp2-V1s or Renshaw cells (Figure 5F). PGC neurons mostly receive inputs from V1s that are non-Foxp2 and non-Renshaw cells, but the frequency of these synapses is very low and highly variable. The data suggest that the Foxp2-V1 clade is a major source of inhibitory inputs to motoneuron cell bodies where they likely strongly modulate motoneuron firing. Moreover, Foxp2-V1 synapses preferentially target limb motoneurons. Foxp2-V1 synapses on motoneurons likely originate from those Foxp2 V1 interneurons clustered spatially close to the LMC. Accordingly, motoneurons in spinal segments with Foxp2-V1 interneurons located further dorsally from motor pools (e.g., S1) receive a relatively lower density of synapses from Foxp2-V1 interneurons. Conversely, Pou6f2, Sp8, and other possible V1 clades either do not target motoneurons directly, or they target them sparsely or on distal dendrites. This suggests functional differences among V1 clades in their strength modulating motoneuron firing directly. V1-to-motoneuron soma connectivity is summarized in Figure 5G, highlighting the preferred targets of each V1 clade.

Foxp2-V1 interneurons clustered near LMC motoneurons are genetically distinct

Birthdates, spatial organization, and synapse densities on different motor columns all suggest that Foxp2-V1 interneurons are heterogeneous, and that a laterally located group close to the LMC, might modulate the output of limb motoneurons. To identify potential genetic differences among Foxp2-V1 interneurons, we selected for further study two TFs previously found highly enriched in Foxp2-V1 interneurons: Orthopedia homeobox (Otp) and Foxp4 (Bikoff et al., 2016). We used antibodies to reveal V1 interneurons expressing Otp and/or Foxp4 in the P5 spinal cord in Lumbar 4 and 5 segments. We first used two dual color animals with Foxp2 and non-Foxp2 V1 interneurons labeled with EGFP and tdT, respectively. Both Otp and Foxp4 are almost exclusively expressed by Foxp2-V1 interneurons at P5 with negligible expression in non-Foxp2 V1 interneurons (Figure 6A). Otp was expressed in around 50% of lineage-labeled Foxp2-V1 interneurons, and Foxp4 in 20%. To examine the relationship between these groups with V1 cells that retain Foxp2 expression at P5, we generated different combinations of paired immunolabelings for Otp, Foxp4, and Foxp2 (Figure 6B) in 3 mice with EGFP expression in Foxp2-V1 interneurons. We constructed cell density contours and calculated the percentages of lineage-labeled Foxp2-V1 cells (EGFP) expressing different combinations of Otp, Foxp4, and Foxp2 (Figure 6C). Each TF combination is found in significantly different percentages of cells in the Foxp2-V1 lineage (one-way ANOVA followed by post-hoc Bonferroni corrected t-tests, statistical details in supplemental tables S5, S6 and S7). To summarize the most salient results, a large group of Foxp2-V1 cells that co-expresses Otp and Foxp2 at P5 (44% of Foxp2-V1 cells) is localized close to the LMC. It includes a smaller subgroup located more ventrally that also expresses Foxp4 (23% of Foxp2-V1 cells). Foxp4-IR / Foxp2-V1 interneurons always co-localized with Otp-IR, indicating they are a subpopulation of the Otp group. Lineage labeled Foxp2-V1 cells expressing only Foxp2 preferentially occupy a ventromedial location typical of late-born Foxp2-V1 cells (Figure 2C), or a dorsal location typical of the earliest-born Foxp2-V1 cells.

Subgroups of Foxp2-V1 interneurons defined by transcription factor expression at P5 and birthdate.

A, Otp (blue) and Foxp4 (white) expression in lineage-labeled Foxp2-V1s (EGFP, green) and non-Foxp2-V1s (tdT, red). The boxed area is shown at higher magnification with different color combinations for clarity. It shows that Foxp4-IR cells in the Foxp2-V1 population always expressed Otp (arrows). Quantification: 49.8-51.2% of Foxp2-V1s express Otp and 20.7-21.7% Foxp4 (n=2 mice each examined in 3 ventral horns in L4/5). Very few non-Foxp2-V1 cells express either TF (Otp: 1.3-4.2% and Foxp2: 0.5-2.9%). B, Images of P5 spinal cords containing Foxp2-V1 lineage labeling (EGFP, omitted for clarity) and double immunolabeled for Otp/Foxp2, Foxp4/Foxp2 and Otp/Foxp4. C, Quantification of Foxp2-V1 interneurons with different combinations of TF expression at P5. For each combination, the left panels show cell distributions, and the right graphs the percentage of Foxp2-V1s with each combination (n=3 mice each analyzed in 6 ventral horns). The data was analyzed with one-way ANOVAS followed by Bonferroni corrected pair-wise comparisons (Statistical details in Supplemental Tables S5, S6 and S7. ****p<0.0001; ***p<0.001; **p<0.01; *p<0.05). There are five main groups defined by TF expression patterns: Otp-Foxp2 (44%), Otp-Foxp2-Foxp4 (23%), Otp only (15%), Foxp2 only (9%) and no TF labeling (32% by subtracting all other groups). Some groups associate with specific locations: Otp-Foxp2 cells and the smaller proportion of Otp only cells are located laterally; Otp-Foxp4-Foxp2 cells are located latero-ventrally; medial cells either contain only Foxp2 or nothing; some dorsal cells are either Foxp2 only or do not express any of these TFs. D, Example image of EdU birthdating in en1Cre::foxp2flpo::R26 RCE:dualGFP/Ai9tdT dual-color mice pulse labeled at E11. Foxp2-V1s (green and yellow cells) are born between E10 and E12, with peak birthdate around E11 and after non-Foxp2 V1s (tdTomato only). The lower graph’s data is normalized to highlight the time of peak neurogenesis for each population (n=15.0 ± 4.1 ventral horns from one mouse per time point from E10 to E11 and 2 mice per time from E11.5 to E12.5; total 9 mice, bars show SEM). There is no difference between green (EGFP only) and the smaller population of yellow cells (EGFP and tdTomato). E, EdU labeling in the Foxp2-V1 dual-color genetic model combined with Otp and Foxp4 antibody staining. F, V1s expressing Foxp2, Otp, and Foxp4 at P5 are mostly born between E10.5 and E11.5 with neurogenesis time courses largely overlapping. The normalized plot indicates that peak neurogenesis for all three populations occurs at E11.5, although a marginally higher number of Otp V1 cells are born earlier (n=4 ventral horns from 3.27 ±1.34 mice per timepoint; error bars show SEM). Contour plots to the right show settling locations of Otp and Foxp4-IR populations born at each time point. G, Schematic of the L4-5 ventral horn summarizing Foxp2-V1 subgroups according to location and combinatorial expression of Otp, Foxp4, and Foxp2 at P5. LMCD: dorsal lateral motor column; LMCV: ventral lateral motor column; MMC: medial motor column.

The birthdates of genetically labeled Foxp2-V1 (EGFP) and non-Foxp2-V1 cells (tdT) (Figure 6D) parallel the early (most non-Foxp2 V1s) and late (Foxp2 V1s) times described previously (Figure 1G-H). Like Foxp2-IR V1 cells, most cells in the genetically labeled Foxp2-V1 lineage were born between E10.5 and E12.0, with few born at E10.0 or E12.5 (n=1-2 mice per time point). The peak of neurogenesis for the lineage-labeled population occurred at E11.0; that is 12 hours earlier than the peak of neurogenesis for V1 cells expressing Foxp2 at P5. This can be explained by a proportion of the earliest born cells in the Foxp2-V1 clade downregulating Foxp2 expression by P5. Next, we used Otp and Foxp4 to subdivide lineage-labeled lateral Foxp2-V1 cells (Figure 6E). To increase sample size, we pooled spinal cords from mice in which we genetically labeled Foxp2-V1s (EGFP) and non-Foxp2-V1s (tdT) together with mice having all V1s lineage-labeled with tdT (n = 4 at E10, 5 at E10.5, 4 at E11, 3 at E11.5 and 2 at E12.5). Pooling data from both genetic models is justified by the above results showing that V1 cells expressing Otp or Foxp4 at P5 are all contained within the Foxp2-V1 lineage. The neurogenesis curves of V1 cells expressing Otp, Foxp4 and Foxp2 at P5 largely overlapped (Figure 6F). The locations of Otp and Foxp4 expressing V1s generated between E10.5 and E11.5 were lateral for Otp and ventral for Foxp4. The location of Foxp4 cells generated at E10.5, E11.0 and E11.5 did not change significantly, but the location of Otp cells generated at E11.5 shifted ventrally compared to those born at E10.5 or E11.0 (Figure 6F). We interpret this result as suggesting a shift in the balance towards generation of Foxp4-Otp Foxp2-V1 interneurons at later times.

The results suggest at least four groups of Foxp2-V1 cells according to location, TF expression, and birthdate (summarized in Figure 6G). Group I is located dorso-medially, lacks Otp, and has variable expression of Foxp2 at P5. Many are likely generated during early Foxp2-V1 neurogenesis (before E10.5). Groups II and III are laterally located and express Otp with or without Foxp2 (group II) or express Otp, Foxp4 and Foxp2 together (group III). These two groups together represent the largest class of Foxp2-V1 cells in L4/5 and are generated during a 24-hour period from E10.5 to E11.5 with a slight shift in the balance of group II vs III cells at later times of neurogenesis. Finally, ventro-medial Foxp2-V1 cells (group IV) are generated very late (after E12: see Figure 2C) and lack expression of Otp or Foxp4, but many retain expression of Foxp2 at P5. We failed to identify in this ventro-medial Foxp2-V1 group expression of late-born markers, like NeuroD2 and Prox1 (Delile et al., 2019; Osseward et al., 2021), despite the presence of many other neurons positive for these TFs in the vicinity.

Otp-expressing Foxp2-V1 cells receive proprioceptive synapses

Earlier studies reported that Foxp2-V1 interneurons include a large class of proprioceptive interneurons. Some of these could represent reciprocal Ia inhibitory interneurons (IaINs) because they receive convergent synapses from excitatory proprioceptive afferents (VGLUT1+) and inhibitory Renshaw cells (calbindin+ V1 axons) (Benito-Gonzalez and Alvarez, 2012). Likely candidates are Otp Foxp2-V1 interneurons in groups II and III because their localization matches that of electrophysiologically identified IaINs in the cat (Jankowska and Lindstrom, 1972; Alvarez et al., 1997). To examine this, we first analyzed the types of Foxp2-V1 interneurons receiving proprioceptive inputs. The spinal cords of two mice at P5 (to preserve TF expression) containing Foxp2-V1 lineage-labeled cells (EGFP) were dual or triple immunolabeled for VGLUT1 and Otp and/or Foxp2 (Figure 7A). VGLUT1 synapses in the ventral horn at P5 originate solely from parvalbumin+ proprioceptors, the majority being Ia afferents (Alvarez et al., 2004). We analyzed 6 ventral horns at L4/L5 in each animal, tiling the whole ventral region containing Foxp2-V1 cells using high magnification confocal microscopy (60x). Cells were categorized as receiving no synapses (I in Figure 7A, 60x inlay), or low/medium and high density of VGLUT1 synapses (respectively, II and III in Figure 7A, 60x inlay). We also noted whether these synapses were located proximally (on cell body and primary dendrites) or more distally. In general, cells with proximal VGLUT1+ synapses had higher densities than those with only distal synapses. Overall, we found that 63.0% and 74.3% of Foxp2-V1 interneurons received VGLUT1 synapses in each animal respectively (Figure 7C). Foxp2-V1 cells with no VGLUT1 synapses were found throughout the ventral horn, but those receiving VGLUT1 synapses had a bias towards lateral positioning (Figure 7B).

Proprioceptive (VGLUT1+) inputs preferentially target lateral Foxp2-V1 interneurons.

A, Left, low magnification confocal image of Foxp2-V1 lineage-labeling (EGFP) combined with VGLUT1 antibody staining to identify primary proprioceptive afferent synapses. Center, the boxed area is expanded and shown with the magenta OTP-IR channel overlaying the GFP channel instead of merging with the green to maximize visual discrimination of OTP-IR Foxp2-V1s (denoted by asterisks). The overlay method used here is fully described in Methods, “Figure composition”. Right, the center image’s inlay is shown at high magnification to demonstrate variability of VGLUT1 synapse density on Foxp2-V1 interneurons (I = absent, II = medium or low, III = high). For simplicity and rigor, we classified Foxp2-V1 interneurons as receiving or not receiving VGLUT1 synapses. Asterisks here indicate which of these Foxp2-V1s are OTP-IR based on the lower magnification images. OTP-IR Foxp2-V1s are laterally biased in their positioning and tend to have higher VGLUT1 synapse densities. B, Distribution of Foxp2-V1 interneurons with and without VGLUT1 synapses (green and magenta, respectively). The positioning of Foxp2-V1 interneurons receiving VGLUT1 synapses is laterally biased. C, Percentages of Foxp2-V1s (GFP+ interneurons), OTP-IR Foxp2-V1s, and Foxp2-IR Foxp2-V1s receiving VGLUT1 synapses (both proximal and/or distal). Each dot is an animal estimate from 6 or 7 ventral horns with respectively 591 and 525 Foxp2-V1 interneurons sampled. Lines indicate the averages of both animals. D, Percentages of lineage labeled Foxp2-V1 interneurons receiving VGLUT1 synapses and with Otp expression (left graph) or Foxp2 (right graph). Each dot represents one mouse, and the lines indicate averages. The numbers of sections and genetically labeled Foxp2-V1 interneurons sampled in each mouse are as in C. In each mouse, this includes 237 and 236 Otp-IR cells and 256 and 225 Foxp2-IR cells.

VGLUT1 synapses preferentially targeted genetically labeled Foxp2-V1 interneurons expressing Otp and Foxp2 at P5 (Figure 7C and 7D). On average (n=2 mice, 12 ventral horns and 1,116 Foxp2-V1 cells analyzed), 42.5% of Foxp2-V1 interneurons had VGLUT1 contacts and were Otp+, while 22.1% were Otp(-). Thus, 65.8% of interneurons receiving VGLUT1 synapses in the Foxp2-V1 lineage express Otp. Only 9.3% of Foxp2-V1 interneurons that were Otp+ lacked VGLUT1 synapses (Figure 7C). Thus, VGLUT1 synapses contacted 82.0% of Otp+ Foxp2-V1 interneurons and 64.9% received these synapses proximally (cell body and primary dendrites) in addition to also having VGLUT1 synapses on more distal dendrites. Sections immunolabeled for Foxp2 revealed 38.2% of lineage labeled Foxp2-V1 interneurons with VGLUT1 contacts expressed Foxp2 while 17.4% were Foxp2(-). Only 14.0% of Foxp2-V1 interneurons retaining Foxp2 expression at P5 lacked VGLUT1 synapses (Figure 7C). This suggests that 68.7% of interneurons in the Foxp2-V1 lineage that were targeted by VGLUT1 synapses had Foxp2 expression at P5, and therefore 73.3% of Foxp2-V1 interneurons with genetic and antibody labeling received VGLUT1 synapses. Of these cells, 60.2% received VGLUT1 inputs proximally. In one of the two animals we also analyzed Otp and Foxp2 co-localization in Foxp2-V1 interneurons receiving VGLUT1+ synapses. We found that 85.6% of Otp+ cells receiving VGLUT1 synapses in the Foxp2-V1 lineage also maintain Foxp2 expression at P5, while 89.8% of genetic and Foxp2 antibody labeled V1s receiving VGLUT1 synapses are also Otp+. In conclusion, laterally positioned Foxp2-V1 interneurons that express Otp and Foxp2 at P5 are preferentially targeted by VGLUT1 synapses.

We attempted to genetically target these cells using an intersection between otp and en1 with a new otp-flpo mouse (see supplemental Figure S5A,B for generation and validation). We found double number of lineage labeled Otp-V1 cells in the spinal cord of dual color mice (en1cre/+::Ai9-R26tdTomato Otpflpo/+::RCE:dual-EGFP) compared to those V1 cells expressing Otp protein at P5 (supplemental Figure S5C,D). Moreover, only 11.4% of V1 cells expressed only tdT (en1 and no otp). Both results together suggest widespread transient expression of Otp in V1 cells before P5. Consequently, markers of non-Foxp2-V1 clades, such as calbindin and Pou6f2, were found in significant numbers within lineage labeled Otp-V1 cells (supplemental Figure S5E,F). This demonstrates that Otp is expressed by subpopulations of cells in all V1 clades during embryonic development. Unfortunately, currently there is no genetic model to temporally control recombination from the otp locus at P5 and restrict expression to postnatally Otp-expressing Foxp2-V1 cells.

Foxp-V1 interneurons form reciprocal inhibitory circuits between antagonistic ankle muscles

Finally, to examine whether Foxp2-V1 interneurons are embedded in reciprocal inhibitory circuits, we combined EGFP lineage labeling of Foxp2-V1 interneurons with anterograde labeling of tibialis anterior (TA) muscle sensory afferents (via cholera toxin subunit B (CTB) muscle injections) and retrograde monosynaptic labeling of interneurons premotor to the antagonistic lateral gastrocnemius muscle (LG) (using glycoprotein (G) deleted mCherry rabies virus (RVΔG-mCherry) (Figure 8A). The sections were immunolabeled for CTB, VGLUT1, mCherry and EGFP (Figure 8B). To obtain transcomplementation of RVΔG-mCherry with glycoprotein in LG motoneurons, we first injected the LG muscle with an AAV1 expressing B19-G at P4. We then performed RVΔG and CTB injections at P15 to optimize muscle targeting and avoid cross-contamination of nearby muscles. Muscle specificity was confirmed post-hoc by dissection of all muscles below the knee. Analyses were done at P22, a timepoint after developmental critical windows through which Ia (VGLUT1+) synaptic numbers increase and mature on V1-IaINs (Siembab et al., 2010).

Some Foxp2-V1 interneurons are reciprocal Ia inhibitory interneurons (IaINs).

A, Experimental design to label spinal neurons that receive inputs from tibialis anterior (TA) muscle primary afferents and connect to lateral gastrocnemius (LG) motoneurons, forming Ia reciprocal inhibitory connections from TA to LG. TA sensory afferents are labeled anterogradely with CTB followed by antibody detection of CTB and the presynaptic marker VGLUT1. Interneurons premotor to LG motoneurons are labeled by monosynaptic retrograde labeling with RVΔG-mCherry. B, Foxp2-V1 IaIN with the most TA/VGLUT1 contacts (31) in our sample (n = 5). Left, low magnification image of Foxp2-V1 interneurons (EGFP, green), RV-mCherry labeling (red) of LG muscle afferents in the dorsal horn and interneurons presynaptic to the LG and of TA afferents anterogradely labeled with CTB (white). The Foxp2-V1 interneuron contains mCherry (yellow cell, inside box). This cell is magnified in two panels to the right, one showing Foxp2-V1 and RV-mCherry and the other RV-mCherry and CTB labeling. Arrows in the zoomed image show examples of CTB synapses (confirmed with VGLUT1) on the dendrites of this neuron. Far right image is the 3D reconstructed cell (Neurolucida) with CTB/VGLUT1 synapses indicated on its dendrites by yellow stars. The axon initial trajectory is indicated (the axon is lost at the section cut surface). The blue area highlights lamina IX. C, Low magnification of section serial to B, showing TA-CTB afferents (white) and LG-RV-mCherry labeled interneurons (red). Transsynaptically labeled interneurons are categorized according to position and Foxp2-V1 lineage labeling: images below show superimposed Foxp2-V1 EGFP (green) and additional VGLUT1 immunolabeling (blue). The location of the LMC is indicated. Contralateral interneurons were found in LX (as the one in this section) and in LVIII (the other two in this animal, not shown). D, High magnification images of two LVII LG-coupled interneurons (RV-mCherry, red) receiving synapses from TA afferents (dendrites in boxed regions are shown at high magnification demonstrating CTB-TA labeling and VGLUT1 content). The most medial interneuron belongs to the Foxp2-V1 lineage (see inset with +EGFP). Insets shown VGLUT1 synapses with (arrows) and without CTB from the TA muscle. E, Neurolucida neuronal reconstructions showed that the Foxp2-V1 interneuron contained a medium number of TA/VGLUT1 synapses (13 contacts) in our sample of putative IaINs derived from the Foxp2-V1 lineage (n=5), while the non-Foxp2-V1 interneuron contained the largest (115) of any LVII reconstructed interneuron with mCherry, including many proximal synapses. For further examples of labeling in serial section from this animal, see Supplementary Figure 6.

Unfortunately, motoneuron infection from muscle and transsynaptic retrograde labeling using RVΔG is known to be inefficient after P10 (Stepien et al., 2010). Additionally, at older ages transsynaptic transport is slower and more temporally spread such that fewer interneurons are recovered at single time points after injection. We chose 7 days post-injection for analyses to avoid as much as possible cell degeneration that frequently occurs at longer survival times after RV infection. We traded the low yield of these experiments for higher specificity when identifying synaptic inputs from TA sensory afferents onto Foxp2-V1 interneurons that are premotor to the LG motor pool. We injected 5 animals that were analyzed in serial sections from segments L2 to L6. All mice showed consistent TA anterograde labeling that occupied the dorsal third of LVII and LIX in the ventral horn of the L4/L5 segments (Figures 8B and 8C; supplemental Figure S6A-C). This distribution matches the well-known musculotopic trajectories of central Ia afferents axons in the ventral horn (Ishizuka et al., 1979). In agreement with the known rostro-caudal trajectories of Ia afferent terminal axon collaterals, TA-CTB VGLUT1+ synapses were found in all lumbar segments, but caudal lumbar segments had the largest density in the ventral horn. Additionally, there were dense projections to medial LV and to discreet regions in LIV in all segments and in all animals. Projections to superficial laminae (I to III) were more common in upper lumbar segments. Only 3 mice showed transsynaptic transport of RVΔG-mCherry from the LG motor pool to interneurons, with large variability from animal to animal. In the best animal, we recovered 51 transsynaptically labeled interneurons with no evidence of degenerative phenotypes (examples in Figure 8 and supplemental Figure S6). These cells were found at the same locations, and in similar proportions as was reported by other groups using injections in younger animals with more cells were labeled (Stepien et al., 2010; Tripodi et al., 2011; Ronzano et al., 2022). The interneuron sample included cells in the ipsilateral Renshaw area (n=6 or 11.8%), LVII (15, 29.4%), medial LV (7, 13.7%), LI to LIV (20, 39.2%), and the contralateral spinal cord (3, 5.9%: 1 in LX and 2 in LVIII). Pooling cells with transsynaptic labeling from all three animals, we identified 8 out of 15 LVII as Foxp2-V1 interneurons. Their dendritic arbors were reconstructed in Neurolucida following mCherry labeling. Five of these cells received more than 1 TA synaptic contact (CTB+ and VGLUT1+) but with large differences in number (5, 9, 12, 17, and 31 synapses). Most synaptic contacts occurred on dendrites, particularly those crossing areas with many CTB-labeled TA afferents. Thus, the direction of dendrites in the section strongly influenced the total number of synapses detected. VGLUT1 densities on spinal interneurons are known to depend on dendritic trajectories with respect to VGLUT1 synaptic fields (Siembab et al., 2016). The Foxp2-V1 cell shown in Figure 8B received the most contacts, which were concentrated on a dendritic segment crossing a field with a high density of TA afferent synapses. Within single sections we found LG-coupled Foxp2-V1 and non-Foxp2 LVII interneurons receiving TA synapses on their dendrites (Figure 8C and 8D). Non-Foxp2 IaINs could be derived from V2b or even non-Foxp2-V1 cells. It is known that several genetic subclasses contribute to the full repertoire of IaINs controlling different leg joints (Zhang et al., 2014). These results provide proof-of-principle that some Foxp2-V1 interneurons are in synaptic circuits capable of exerting reciprocal Ia inhibition between antagonistic muscles. Clearly, a technique with higher yield and that also maintains high specificity is necessary. Additionally, analysis of further extensor-flexor pairs in different joints in both directions will need to be performed to reveal a complete picture of IaIN organization.

Other cell types transsynaptically labeled from the LG included V1-Renshaw cells which did not receive any TA synapses since their dendrites are far away from TA projection areas (Supplemental Figure S6B). Another group were medial LV LG-coupled interneurons (possibly Ia/Ib interneurons not derived from either V2b or V1 classes) which are in a region with high density of TA/VGLUT1+ synapses (Supplemental Figure S6A and S6C). A few reconstructed neurons at this location (n=3) received the highest densities of TA synapses with more than 50 contacts on relatively smaller dendrites. A few other LG-coupled interneurons reside in superficial laminae of the dorsal horn (Figure 8D) where they are also contacted by CTB-labeled TA afferent synapses. Muscle afferents ending in superficial laminae are likely non-proprioceptive (i.e., Type III(Aδ) and IV(C) afferents (Ling et al., 2003)). It has also been shown that interneurons at this location can be transsynaptically labeled in the anterograde direction from sensory afferents (Zampieri et al., 2014). However, rabies virus tropism towards primary afferents depends on developmental age, with abundant infection between P0 and P5 and reduced infection by P15. In our experiments, motoneurons in the location of the LG pool were mCherry-labeled, but there was very limited mCherry labeling in primary afferents in dorsal roots or dorsal columns. These labeled axons were restricted to upper lumbar segments (Supplementary Figure S6, see also Figure 8B and 8C). Thus, ventral mCherry labeled Foxp2-V1 interneurons in our experiment are most likely transsynaptically labeled in the retrograde direction from LG motor pools.

Discussion

A comprehensive inventory of spinal premotor interneurons and circuits requires detailed cataloging of their core components, ideally using multimodal information such as genetic subtypes, timing of neurogenesis, settling positions, incorporation within spinal circuits, and their electrophysiological properties and functional roles during motor behaviors. In the present work we focus on V1 inhibitory interneurons, a large heterogenous group of inhibitory interneurons with ipsilateral synaptic projections throughout the ventral horn (Alvarez et al., 2005). Previously, the diversity of V1 interneurons in mice was organized into at least four clades deduced from Bayesian statistical analyses of combinatorial expression of 19 TFs and their cell locations in the spinal cord (Bikoff et al., 2016; Gabitto et al., 2016). Each group divides into further subgroupings organized in a hierarchical fashion. Here we show that these four V1 clades differ in neurogenesis times and targeting of motoneurons. This result strengthens the conclusion that these V1 clades defined by their genetic make-up might represent distinct functional subtypes, although further validation is necessary in more functionally focused studies.

Sequential neurogenesis of V1 clades reflects their heterogeneous microcircuits

Two V1 clades, Renshaw cells and Pou6f2-V1 cells, have narrow windows of neurogenesis with almost all cells generated before E11, while the Foxp2-V1 and Sp8-V1 clades are generated in a wider temporal window with most born after E11. Within the “early group” many Renshaw cells are generated before Pou6f2-V1 cells and within the “late group” Sp8-V1 cells lag behind Foxp2-V1 cells. These data confirm a previous report on the earlier generation of MafB-V1 Renshaw cells compared to Foxp2-V1 IaINs (Benito-Gonzalez and Alvarez, 2012). Sequential generation of these two V1 cell types seem to be an intrinsic property of p1 progenitors and can be replicated in vitro using mouse embryonic stem cells (mESC) to derive p1 and V1 cells (Hoang et al., 2018). The present data extend the view of sequential determination of cell fate by time of neurogenesis to all V1 clades. In addition, the results suggest that clades with few neurons, limited diversity and specific locations have narrow windows of neurogenesis, while larger V1 clades, like Foxp2-V1 interneurons, are generated through longer periods of neurogenesis and include subtypes that differ in time of neurogenesis and location.

Overall, our data agree with a previous report that examined spinal cord subtypes based on the intersection of neurogenesis and transcriptomics by analyzing mouse spinal cords from E9.5 to E13.5 (Delile et al., 2019). This study divided V1 interneurons into seven groups. Groups V1.5 and V1.2 had early birthdates. The V1.5 group’s gene expression profile includes neurod1,2 and 4, neurod1, prox1, tcf4, lhx1 and 5, pou2f2 and hes1, and although it does not clearly match the TF repertoire of the Pou6f2-V1 clade, it occupies a similar birthdate window. In contrast, the V1.2 group shares the gene expression profile of Renshaw cells which includes calb1, mafb and onecut2. We consistently found Renshaw cells to be the first born V1 cells, although this study suggests they are generated after the V1.5 group. One explanation is that some of the TFs used to define V1.2 cells, like mafb, are upregulated sometime after neurogenesis (Benito-Gonzalez and Alvarez, 2012). The V1.1, V1.4 and V1.3 groups follow in neurogenesis timing, and all three express foxp2; V1.1 and V1.4 have similar birthdates while V1.3 neurogenesis is slightly delayed. We show similar heterogeneity of birthdates in Foxp2-V1 interneurons. Two subgroups expressing the TF Otp at P5 with or without co-expression of Foxp4 and/or Foxp2 have intermediate times of neurogenesis and form a lateral group closely related to LMC motoneurons. They both receive VGLUT1/proprioceptive inputs, and some may form reciprocal inhibitory circuits between antagonistic motor pools. Many Foxp2-V1s located dorsomedially are generated earlier, and Foxp2-V1s located ventromedially are generated last. Many of these cells are not targeted by VGLUT1/proprioceptive inputs which indicates different circuit roles. Finally, groups V1.6 and V1.7 are generated at the end of V1 neurogenesis, and it is tempting to speculate that they include the Sp8-V1 clade and the late generated ventromedial group of Foxp2-V1 interneurons. However, it is presently difficult to match gene expression profiles of V1.6 and V1.7 (neurod1, 2, and 6, Nfia, Nfib, Nfix, prox1, tcf4, hex6, cbln2 and slit2) to our V1 clades. One problem is that the high throughput sequencing in Delile’s paper is based on samples collected at early embryonic times (E9.5 to E13.5), but gene expression in V1 groups changes throughout embryonic and postnatal development. For example, our efforts to generate a genetic model for Otp-V1s demonstrate that the postnatal restriction of this TF to the Foxp2-V1 clade does not occur until after embryogenesis. A similar situation occurs with the Sp8-V1 clade that is best defined by V1 interneurons retaining Sp8 expression at P0 (Bikoff et al., 2016), because Sp8 expression is widely expressed in ventral spinal progenitors during embryogenesis. Conversely, some TFs are not expressed in certain cell types until after neurogenesis or even until late embryonic stages. For example, we obtained preliminary evidence that the medial Foxp2-V1 group upregulates Foxp2 expression after E14 by difference to the lateral group which upregulates Foxp2 expression as they emerge from progenitors (Benito-Gonzalez and Alvarez, 2012). Thus, Foxp2 expression might not be captured in medial Foxp2-V1 interneurons when examining gene expression from E9.5 to E13.5. Despite these differences, there is general agreement that Renshaw cells are early born and that several Foxp2-V1 groups are generated later and can be distinguished by differences in neurogenesis.

More generally, our results agree with the pattern of sequentially generated neurons that is consistent throughout the neural tube (summarized by Sagner, 2024). Frequently, three neurogenesis waves are described, each defined by specific temporally restricted TFs. The earliest neurons are defined by TFs onecut1 and onecut2 which occur in all Renshaw cells, some Pou6f2-V1, a few Foxp2-V1 interneurons and even a subpopulation of Sp8-V1 cells (Bikoff et al., 2016). Intermediate neurons express Pou2f2 and Zfhx3 and 4 that are expressed by Pou6f2-V1 and Foxp2-V1 interneurons (Alexandra Trevisan and Jay Bikoff, unpublished data). Finally, late cells express TFs like Nfia, Nfib, Neurod2 and Neurod6 and their location is restricted to medial regions of the neural tube. These correspond to ventromedial subpopulations of Foxp2-V1 cells and most medially located Sp8-V1 interneurons. Thus, expression of temporarily restricted TFs and the locations of cells from the different V1-clades aligns well with the conclusions of our EdU birthdating study: Renshaw cells are earliest generated, Pou6f2-V1 include cells with early and intermediate generation, while Foxp2-V1 and Sp8-V1 cells contain cells that are generated, early, intermediate or late, at different proportions in each clade.

In addition to differences in neurogenesis timing, we uncovered differences among V1-clades on motoneuron synaptic targeting and this supports the view that these clades constitute unique functional subsets forming distinct ventral horn circuits. Renshaw cells and Foxp2-V1 interneurons are major sources of V1 synapses on motoneuron cell bodies and proximal dendrites. Pou6f2-V1 and Sp8-V1 interneurons either do not provide a significant direct input to the motoneuron or this occurs quite distally. We previously reported that the density of Sp8-V1 synapses on motoneuron cell bodies and proximal dendrites is between one and two orders of magnitude less dense than Renshaw cell synapses at this location (Bikoff et al., 2016). This analysis included a variety of motoneurons innervating flexor and extensor muscles at various hindlimb joints, ruling out the possibility that each interneuron preferentially targets specific motor pools. Moreover, unpublished data from our group using intracellular fills of dorsal MafB-V1 interneurons belonging to the Pou6f2-V1 clade also show a lack of axon projections towards LIX. Given that V1 axons are largely restricted to the ventral horn (Alvarez et al., 2005), it is fair to conclude that Pou6f2-V1 and Sp8-V1 interneurons likely modulate activity in ventral premotor spinal networks and/or perhaps synaptic integration in motoneuron distal dendrites traversing LVII, but do not establish proximal synapses that effectively modulate motoneuron firing and excitability directly. For V1 interneurons that role seems exclusive of Renshaw cells and Foxp2-V1s, although these cells may also target premotor network elements. Within the Foxp2-V1 clade it is not possible at present to define which subgroup provides proximal synapses to motoneurons. However, given the known placement of IaINs (Jankowska and Lindstrom, 1972) and the known proximal location of reciprocal inhibitory synaptic inputs (Burke et al., 1971), it is reasonable to expect that many Foxp2-V1 synapses on the cell bodies and proximal dendrites of motoneurons arise from lateral Foxp2-V1 groups (II and III) which occupy spinal cord locations typical of IaINs and preferentially receive VGLUT1/proprioceptive projections.

It is notable that birthdate order was not predictive of presynaptic coupling to motoneurons or phylogenetic relations. Both early born Renshaw cells and late born Foxp2-V1 cells were preferentially associated with synapses on motoneuron cell bodies. Moreover, the circuits they form, recurrent and reciprocal inhibition of motoneurons, are both phylogenetically recent. Reciprocal inhibition of muscle flexors and extensors at limb joints first appears in limbed Amphibia, a subclass of tetrapodal vertebrates that lacks recurrent inhibitory circuits (Czeh, 1977). At present recurrent inhibition of motoneurons by Renshaw cells has only been detected in mammals and in the developing hindlimb motor pools of chick embryos (Wenner and O’Donovan, 1999; Alvarez and Fyffe, 2007). It is thus possible that Renshaw cells and many Foxp2 V1 interneurons are either phylogenetically recent or that these cells acquired novel connectivity in parallel with the evolution of limb motor control. V1 interneurons are found in all vertebrate species ranging from fishes with swimming locomotion to mammals with limbed terrestrial locomotion, but while V1 interneurons show low diversity in zebrafish (Kimura and Higashijima, 2019) they display large heterogeneity in mice (Bikoff et al., 2016). This could suggest the emergence of novel cell types with the evolution of limbed species. However, the appearance of novel circuits is also supported by observations in the axial MMC, a motor column shared with fishes. These motoneurons are densely innervated by Foxp2-V1s and Renshaw cells and correspondingly, thoracic axial motoneurons in the cat are modulated by Renshaw cell mediated recurrent inhibition (Saywell et al., 2013). This suggests that axial motoneurons in limbed mammals acquired inhibitory controls absent in aquatic vertebrates. Whether this is due to the development of new cell types or the reassignment of existing cells to novel circuits remains to be determined.

The close functional relationship between Renshaw cells and subgroups of Foxp2-V1s involved in reciprocal inhibitory circuits is further emphasized by their intricate connectivity in the mammalian spinal cord. Together with flexor and extensor motoneurons, these cells form a canonical motor output module that controls the muscles around individual joints. This canonical microcircuit was first proposed by Hultborn and colleagues (Baldissera et al., 1981) as an end-stage modulator of the spatio-temporal properties of the motor output across all spinal segments in all mammalian species (human and cats at the time, and now also rats and mice). Furthermore, the significance of Renshaw cell and Foxp2-V1 synapses on motoneurons is highlighted by recent studies in ALS mouse models. These studies showed that V1 synapses located on the cell bodies and proximal dendrites of motoneurons are preferentially affected during early stages of the disease, and in addition Foxp2-V1 interneurons have a high susceptibility for cell death (Wootz et al., 2013; Salamatina et al., 2020; Montaña-Rosell et al., 2024). Despite the close relationship between Renshaw cells and Foxp2-V1 interneurons they greatly differ in firing properties and circuit function. Renshaw cells are burst-tonic firing cells, while Foxp2-V1 interneurons are tonic fast-spiking (Bikoff et al., 2016) with these properties furrther diverging during postnatal maturation. For example, mature Renshaw cell firing properties in postnatal mice (Bikoff et al., 2016) greatly differ from early embryos (E12) (Boeri et al., 2018; Boeri et al., 2021).

In conclusion, previously defined V1 clades exhibit differences in birthdate, heterogeneity, and projections to motoneurons and/or premotor networks. Collectively, these differences may reflect the appearance of novel circuits associated with limb function and overground locomotion. Additionally, early- and late-born V1 interneurons in the mammalian spinal cord are intricately interconnected and together form critical circuits that modulate motor output.

Molecular determinants of V1 diversity

In a remarkable in vitro replication of the V1 neurogenesis sequence using mouse embryonic stem cells (mESC), it was found that 24-hour treatment with a high concentration of retinoic acid and a low concentration of Smoothened agonist induced mESC derivation of p1 progenitors (Hoang et al., 2018). These progenitors sequentially generated V1 interneurons with genetic profiles of Renshaw cells at in vitro day 5 and Foxp2-V1s at day 8. Lengthening the retinoic acid treatment favored differentiation into neurons with Renshaw cell characteristics at the expense of Foxp2-V1s. This suggests that Renshaw cells and Foxp2-V1s derive from a similar pool of p1 progenitors, and that their fates might depend on differences in the morphogenetic signals present at the embryonic times when they are born. For example, in early embryos retinoic acid is highly expressed by the mesoderm adjacent to the developing spinal cord (Novitch et al., 2003) and by motoneurons inside the spinal cord (Sockanathan et al., 2003), but retinoic acid signals are later attenuated by opposing actions from fibroblast growth factor family members (Diez del Corral et al., 2003). A final consideration is that, although Hoang et al. also differentiated V1 cells with gene expression profiles characteristic of Pou6f2 and Sp8 cells, these clades were underrepresented suggesting that p1 progenitors giving rise to Pou6f2 and Sp8 V1 cells may have distinct signaling requirements during embryogenesis.

Significance of Foxp2-V1 interneuron diversity

Genetic lineage labeling of Foxp2-V1 interneurons reveals approximately twice as many V1 interneurons compared to the number identified when this clade is defined by postnatal Foxp2 protein expression, because downregulation of Foxp2. Not surprisingly, we uncovered heterogeneity in this clade. We identified four groups of Foxp2-V1s according to TF combinatorial expression and position (Figure 6G), and this classification scheme reflected the sequential birthdate order of the groups. Of special interest were groups II and III, which were born at mid-to-late neurogenesis times and expressed Otp at P5. These cells were closely associated with the limb-controlling LMC. Foxp2-V1 interneurons at this location also receive proprioceptive VGLUT1 inputs, and some were found in reciprocal inhibitory circuits between the TA and LG. Proprioceptive TA input densities varied across different dendrites of individual Foxp2-V1 interneurons according to dendrite locations relative to the trajectories of specific primary afferents—a concept already established for the major inputs of Renshaw cells (both proprioceptive and motor) (Mentis et al., 2006; Benito-Gonzalez and Alvarez, 2012; Siembab et al., 2016). Similarly, motoneurons with genetically altered dendritic arbor structure exhibit changes in their proprioceptive inputs (Vrieseling and Arber, 2006). In this context, the two groups of Otp-Foxp2-V1 interneurons, defined by Foxp4 expression and dorso-ventral location (groups II and III) are likely connected to primary afferents from different muscle groups. In conclusion, the Foxp2-V1 clade represents a diverse group of interneurons. Types II and III may be closely related to proprioceptive pathways, and differ in specificities for particular muscles, joints, and/or flexors and extensors, while types I and IV may be broadly associated with other motor functions.

Materials and methods

All animal procedures were carried out in accordance with NIH guidelines and were approved by the Emory University Institutional Animal Care and Use Committee (Atlanta, GA). Animal experimentation also follow ARRIVE guidelines.

Animal models

To genetically lineage-label all or subclasses of V1-interneurons we used eight transgenic mouse models (Table 1). These mice were crossed for intersectional genetic labeling combining a line in which all En1-expressing V1 interneurons express Cre with lines expressing Flpo dependent on Foxp2 or Otp or in which MafB expressing cells express GFP.

V1 and Foxp2-V1 model

en1Cre/+ heterozygotes (Sapir et al 2004) were crossed with Rosa26-Frt-lox-STOP-lox-tdTomato-WPRE-frt homozygotes (Ai9R26-lsl-tdT, JAX#007909; B6;129S6-Gt(ROSA)26Sort M14(CAG-tdtomato)Hze/J) to obtain en1cre/+::Rosa26lsl-tdT/+ and after backcrossing, en1cre/+::Rosa26lsl-tdT/lsl-tdT animals (in both animals all V1 cells are lineage-labeled with tdT). Similarly, foxp2flpo/+animals (Bikoff et al 2015) were crossed with RCE:dual-EGFP homozygotes (RCE:FRT JAX#010675; Gt(ROSA)26Sortm1(CAG-EGFP)Fsh; initially donated by Dr. Gordon Fishell, Harvard University) to produce foxp2flpo/+::RosadualEGFPl/+and foxp2flpo/+::RosadualEGFP/dualEGFP mice. Crossing en1Cre/+::Rosa26lsl-tdT/+ or en1cre/+::Rosa26lsl-tdT/lsl-tdT mice with foxp2flpo/+ or foxp2flpo/+::RosadualEGFP/+ or foxp2flpo/+::RosadualEGFP/dualEGFP we obtained mice for experiments with the following genotypes: en1cre/+::foxp2flpo/+::Rosa26lsl-tdT/+(V1 red), en1cre/+::foxp2flpo/+::Rosa26+/dualEGFP (Foxp2-V1 green) or en1cre/+::foxp2flp/+::Rosa26lsl-tdT/ dualEGFP (dual color, Foxp2-V1 green and non-Foxp2-V1 red). In some experiments we substituted the reporter lines Ai9-lsl-tdT and RCE:dualEGFP for the RC::FLTG line. This line has in the R26 locus a frt-flanked STOP and loxP-flanked tdTomato::STOP preventing transcription of EGFP (JAX#026932,B6.Cg-Gt(ROSA)26Sortm1.3(CAG-tdTomato,-EGFP)Pjen/J). In this line, Flpo recombination in cells with Foxp2 expression induces expression of tdT, while cells with additional Cre recombination (V1 expressing Foxp2 cells) will express EGFP and remove the tdT reporter.

V1-Otp model

Similar breeding schemes and reporter lines were used to combine en1cre/+ and otpflpo/+ mice to study Otp-V1 interneurons. These mice were generated as described previously (Bikoff et al., 2016) and summarized in Supplementary Figure 5. Briefly, Flpo, a codon-optimized version of Flp recombinase, was inserted into the ATG in the 1st exon of the Otp genomic locus, generating a null allele. Positive ES cell clones were screened by Southern blot and microinjected into blastocysts, and the resulting chimeric mice were crossed to C57BL/6J females. The neomycin selectable cassette was removed using Protamine::Cre mice (Jax#03328).

V1-MafB model

en1cre/+::Rosa26lsl-tdT/lsl-tdTanimals were crossed with mafbGFP/+ knock-in mice (Moriguchi et al 2006) to label cells expressing MafB and assess their overlap with V1 interneurons.

The en1, foxp2, otp and mafb genes carry the specified transgene knocked in into the gene coding sequence resulting in a null allele. The animals are maintained and bred in heterozygosis, with homozygotes being knockouts for each of these genes. We generated foxp2flpo/flpo and mafbgfp/gfp homozygotes to test antibody specificities. foxp2 knockout mice survive postnatally, but mafb knockouts die at birth due to respiratory failure (Blanchi et al., 2003). Thus, foxp2 knockout mice were harvested at P5 and mafb knockout mice as late embryos. All animals were bred in our colonies at Emory University, and the resulting litters were genotyped using a combination of standard tail PCR (Transnetyx) and fluorescent phenotyping of neonates (each gene combination results in specific patterns of labeling along the body).

Timed pregnancies

Female mice were caged with males at the beginning of the dark period (7:00 PM), and the next morning (7:00 AM) vaginal plugs were checked. A positive plug was considered E0.5; however, since the exact time of mating is unknown, this procedure has an approximate error of 12 hours. Moreover, we found embryos within single litters that differ by 6-12 hour in developmental stage.

Tissue preparation

Mouse pups of different postnatal (P) ages (P0, P5, P15, P30, adult) were anesthetized with an overdose of Euthasol (>100 mg/Kg i.p.) and, after transcranial vascular rinsing with phosphate buffer saline (PBS) and heparin they were perfusion-fixed with 4% paraformaldehyde in 0.1M phosphate buffer (PB). The spinal cords were then dissected, removed, and postfixed in 4% paraformaldehyde for either 2 hours, 4 hours, or overnight. Post-fixation times depend on antigens targeted in immunocytochemistry experiments. In general, calcium buffering proteins, choline acetyltransferase, and synaptic vesicle markers require longer post-fixation times while transcription factors require shorter post-fixation times. After post-fixation, the tissues were cryoprotected in 0.1M PB with 30% sucrose and prepared for sectioning. Transverse spinal cord sections were obtained in a sliding freezing microtome at 50 μm thickness and collected free-floating. Embryos, P0, and P5 spinal cords were cut in a cryostat at 20 or 30 μm thickness from tissue blocks snap-frozen in OCT.

Birthdating experiments

5-ethynyl-2’-deoxyuridine (EdU; Invitrogen) was injected i.p. at a dose of 50 mg/kg weight in timed-pregnant females. The data reported were obtained from 17 pregnant females successfully injected at gestation days E9.5, 10, 10.5, 11, 11.5, 12m or 12.5 after crossing with appropriate males to generate pups with genetic labels for all V1, Foxp2-V1 or MafB-V1 interneurons. The spinal cords were collected after P5 perfusion-fixation with paraformaldehyde as above. P5 was chosen for analyses to maximize transcription factor antigenicity.

EdU Click-iT reaction

Fifty micrometer thick transverse spinal cord sections were obtained in a freezing, sliding microtome from lower lumbar segments (4 and 5) and processed free floating with Click-iT Alexa Fluor 488 (C10337, Invitrogen) or Click-iT Alexa Fluor 647 Imaging Kits (C10340, Invitrogen) depending on other genetic fluorophores present in the animal. Sections were washed twice (5 min) with 3% bovine serum albumin (BSA, Fisher) in 0.01M phosphate buffered saline (PBS) and then then permeabilized with a solution of 0.5% Triton X-100 (Fisher) in 0.01M PBS at room temperature for 20 min. During this time, the Click-iT reaction cocktail was prepared per manufacturer instructions and applied to the sections for 30 min at room temperature protected from light. Sections were washed with 3% BSA in 0.01M PBS at the conclusion of the incubation. Antibody labeling followed the EdU Click-iT reaction. In one pup EdU labeling did not correspond to the target time point and was discarded (#459.2 in Figure 1D). The littermate (#459.1 in Figure 1D) displayed correct EdU labeling for the injection time.

Immunohistochemistry

The characteristics, RRID numbers, and dilutions of all primary antibodies used are summarized in Table 2. Genetic labels, tdT and EGFP, were always amplified with antibodies to aid in visualization. After blocking the sections with normal donkey serum (1:10 in PBS + 0.3% of Triton-X-100; PBS-TX), we incubated the section in different primary antibody cocktails diluted in PBS-TX. Chicken antibodies were used to detect EGFP, and mouse or rabbit antibodies were used to detect tdT, depending on the hosts of primary antibody combinations. In birthdating experiments we used in serial sections either rabbit anti-MafB (Sigma), goat anti-Foxp2 (Sant Cruz), rabbit anti-Pou6f2 (Sigma), goat anti-Sp8 (Santa Cruz), guinea pig anti-Otp (Jessell lab; Bikoff et al., 2016), rabbit anti-Otp (Jessell lab; Bikoff et al., 2016), rabbit anti-Foxp4 (Jessell lab; Bikoff et al., 2016), or rabbit anti-calbindin (Swant). Depending on color combination for triple or quadruple fluorescent labeling with genetic reporters (EGFP or tdTomato) and the EdU Click-it reaction (Alexa 488 or Alexa 647), these markers were revealed with either FITC, Cy3 or Cy5 conjugated species-specific donkey-raised secondary antibodies, or with biotinylated secondary antibodies followed by streptavidin Alexa-405 (for all secondary antibodies dilution was from 1:100 to 1:200; all secondary reagents were obtained from Jackson ImmunoResearch). After immunoreactions, the sections were mounted on slides and cover-slipped with Vectashield (Vector Laboratories). Similar immunocytochemical protocols were followed in other experiments.

Analysis

Confocal images (10X and 20X) were obtained with an Olympus FV1000 microscope. Image confocal stacks were fed into Neurolucida (MicroBrightField) for counting and plotting cells. We analyzed 4 ventral horns per animal in lower lumbar segments (L4-L5). Cells were classified according to genetic labeling, TF immunoreactivity, and EdU labeling. EdU labeled cells were classified as strongly labeled (at least two thirds of the nucleus uniformly labeled) or weakly labeled (speckles or partial nuclear labeling). From Neurolucida plots we estimated: (1) the percentage of V1 INs labeled with EdU (strongly or weakly); (2) the percentage of V1 cells labeled with TF antibodies; (3) the percentage of V1 cells genetically labeled with MafB-GFP; (4) the percentage of V1 INs genetically labeled with Foxp2; (5) the percentage of V1 cells with different genetic or immunocytochemical markers and incorporating weak or strong EdU at different time points; (6) the cumulative numbers of EdU weakly or strongly labeled cells for all V1 cells and for each marker across all embryonic times.

Cell location/density analyses

From Neurolucida plots we constructed cell density profiles for each V1 interneuron type and birthdate, assigning cartesian coordinates to the nucleus location with respect to the dorsal edge of the central canal which was defined as position (0,0). Coordinates were exported as .csv files and plotted using custom Matlab scripts to display the position of each individual cell (Bikoff et al., 2016). Distribution contours were constructed in Matlab using the kde2d function (Matlab File Exchange), which estimates a bivariate kernel density over a set of grid points. We plotted density contours containing from 5% (inner contours) to 95% (most outer contour) of the cell population in 10% increments. Density contours were superimposed onto hemicord schematic diagrams in which the distance from central canal to lateral, dorsal, or ventral boundaries was adjusted depending on age and segment from direct measurements obtained in Neurolucida.

Analyses of lineage-labeled Foxp2-V1, Otp-V1, and MafB(GFP)-V1 cells across ages and markers

The spinal cords of animals carrying different combinations of genetic labels were prepared as above for amplification with immunolabeling of their fluorescent reporters and combination with different markers (for the characteristics of the different samples in terms of number of animals and sections analyzed see the text in results).

Analysis of Foxp2-V1 interneurons in different segments of the adult spinal cord

We used en1cre/+, foxp2flpo/+, Rosa26+/dualEGFP mice and amplified the EGFP signal as above. The position of motoneurons was revealed using a goat anti-choline acetyltransferase (ChAT, Millipore) antibody. Laminae cytoarchitectonics were assessed with a rabbit recombinant anti-NeuN antibody (Novus) (not shown in figures). ChAT and NeuN immunoreactivities were detected with Cy3 and Cy5 conjugated donkey anti-goat IgG antibodies, respectively (Jackson ImmunoResearch). Confocal microscopy images were obtained (as above), plotted in Neurolucida and cell density contours generated. We analyzed the number of Foxp2-V1 cells and cholinergic motoneurons per section and calculated their ratios (a motoneuron was defined as any cholinergic immunoreactive cell in Lamina IX). Density contours were used to compare distributions in different segments.

Analyses of transcription factor expression in Foxp2-V1 interneurons at P5

To examine Otp and Foxp4 immunoreactivity in lineage-labeled Foxp2 vs. non-Foxp2 V1 cells, we used dual color en1cre/+::foxp2flp/+::Rosa26lsl-tdT/dualEGFPmice. The sections were immunolabeled with guinea pig anti-Otp (Jessell Lab, CU) and rabbit anti-Foxp4 (Jessell Lab, VU). Otp immunoreactive sites were revealed with biotinylated donkey anti-guinea pig followed by streptavidin Alexa-405 and Foxp4 immunoreactivity was detected with donkey anti-rabbit IgG secondary antibody conjugated to Cy5. These antibodies were combined with immunocytochemical amplification of the EGFP signal (chicken anti-EGFP) and tdTomato (mouse anti-RFP) respectively with FITC and Cy3 conjugated species-specific IgG antibodies raised in donkey. To analyze the populations of lineage-labeled Foxp2-V1 cells expressing Otp, Foxp4 and Foxp2 at P5, we used single color en1cre/+::foxp2flp/+::Rosa26+/dualEGFPmice and amplified EGFP fluorescence with antibodies as above, combined with guinea pig anti-Otp (Jessell, CU)/rabbit anti-Foxp4 (Jessell, CU), rabbit anti-Otp (ThemoFisher)/goat anti-Foxp2 (Santa Cruz) and rabbit anti-Foxp4 (Jessell, CU)/goat anti-Foxp2 (Santa Cruz) antibodies. Triple immunostains were revealed with species-specific secondary antibodies as above using the Cy3 and Cy5 channels for transcription factors. A few sections were immunolabelled with rabbit antibodies against NeuroD2 (Abcam) or Prox1 (Millipore/Sigma). In this case immunoreactivities were revealed with Cy3-conjugated anti-rabbit IgG antibodies. All sections were imaged using confocal microscopy and confocal stacks analyzed in Neurolucida as above. All analyses were done in Lumbar 4 and 5 segments.

Analyses of V1 clade markers in MafB-V1 and Otp-V1 interneurons at P5

For these analyses we used en1cre/+::otpflpo/+::Rosa26+/dualEGFP or en1cre/+::Rosa26lsl-tdT/+::mafbGFP/+mice. To analyze Otp-V1 interneurons, EGFP fluorescence was amplified with chicken anti-EGFP antibodies as above, and one of the following additional primary antibodies was added in serial sections: rabbit anti-calbindin (Swant), guinea pig anti-Otp (Jessell CU), guinea pig anti-Pou6f2 (Jessell CU), goat anti-Foxp2 (Santa Cruz), or goat anti-Sp8 (Santa Cruz). All markers were detected in the Cy3 channel. When possible, they were combined with goat anti-ChAT antibodies (EMD Millipore) or mouse anti-NeuN antibodies (Millipore) labeled in the Cy5 channel. ChAT and NeuN immunoreactivities are not shown in results but were used to identify laminae and spinal cord segments according to motor column organization and cytoarchitectonics. In MafB-V1 mice we amplified both EGFP and tdT as above and combined with the following antibodies: rabbit anti-calbindin (Swant), guinea pig anti-Pou6f2 (Jessell CU), goat anti-Foxp2 (Santa Cruz) and goat anti-Sp8 (Santa Cruz). All marker antibodies were revealed using Cy5 conjugated donkey species-specific anti-IgG secondary antibodies as above. Analyses were done as above: confocal images were imported into Neurolucida for cell plotting and the results expressed as number or proportion of neurons and their position analyzed using cell distribution density profiles. All analyses were done in Lumbar 4 and 5 segments.

Analyses of Foxp2-V1 and Foxp2-non-V1 cells at P5

For these analyses we generated two en1cre/+::foxp2flpo/+::Rosa26+/FLTG mice. EGFP (Foxp2-V1 cells) and tdTomato (Foxp2-non-V1 cells) were amplified with antibodies as above. The sections were counterstained with mouse anti-NeuN antibodies (Millipore) for segment and laminar localization. Sections from Lumbar 4 and 5 segments were imaged with confocal miscopy and analyzed in Neurolucida. Cell distribution plots, cell numbers, and cell density curves were obtained as above.

Analyses of VGLUT1 inputs on Foxp2-V1 interneurons

Analyses were done at P5 in en1cre/+::foxp2flp/+::Rosa26dualEGFP/+mice. The P5 age was selected to preserve TF immunoreactivity. Moreover, at this age VGLUT1 synapses in the ventral horn are specifically contributed by proprioceptive afferents, most likely Ia afferents (Alvarez et al., 2004). For these analyses spinal cord sections were obtained in a cryostat (20 μm thickness). EGFP was amplified with chicken anti-GFP (Aves) and combined with goat anti-Foxp2 (Santa Cruz), rabbit anti-Otp (ThermoFisher) and guinea-pig anti-VGLUT1 (Synaptic Systems). EGFP, Foxp2 and Otp immunoreactivities were revealed respectively with FITC, Cy3 and Cy5 conjugated species-specific secondary antibodies. VGLUT1 was revealed with biotinylated anti-guinea-pig IgG antibodies followed by streptavidin-Alexa 405. The sections were imaged at 10X and 60X with confocal microscopy. VGLUT1 contacts were analyzed only in high magnification images in which we tiled the whole ventral horn to sample every Foxp2-V1 cell present. Using Neurolucida, we plotted all Foxp2-V1 cells in each section and classified them according to the presence of VGLUT1 inputs and whether they were at high or low density in a qualitative assessment. Later we pooled all data by the presence or absence of VGLUT1 contacts. From these plots we estimated: (1) the percentage of lineage-labeled Foxp2-V1 interneurons with VGLUT1 contacts; (2) the percentage of these cells with Otp, Foxp2 or both TFs co-localized (for sample attributes with respect to number of animals, sections and cells analyzed, see results).

Identification of Foxp2-V1 interneurons interposed in reciprocal connections between the tibialis anterior muscle and the lateral gastrocnemius motor pool

For these analyses we combined retrograde monosynaptic tracing of RVΔG-mCherry from the LG muscles with anterograde tracing of muscle sensory afferent synapses using cholera toxin subunit b (CTB) from the TA. RVΔG-mCherry and CTB intramuscular injections were done at P15 to avoid critical windows of synaptic reorganization. To facilitate tracing at this age we applied RVΔG-mCherry at high titer (>109 TU/ml) and CTB was injected at high concentration (1%).

Production of RVΔG-mCherry

SADB19ΔG-mCherry rabies virus (RVΔG-mCherry) and the BHK-B7GG (B7GG) cells expressing B19 RV glycoprotein (RV-G) and nuclear GFP for virus propagation and amplification were donated by Dr. Edward Callaway (Salk Institute, La Jolla, CA). Sterile cell culture technique without antibiotics was used throughout all procedures. B7GG cells were placed in cell culture dishes containing DMEM in 10% fetal bovine serum (FBS) (culture medium) and incubated in 5%CO2 in humid air at 37°C for two hours. Once they adhered to the substrate, the medium was removed, the cultures washed with 10 ml warm PBS and fresh medium applied. Four plates were grown to 90% confluency, and then 4 ml of virus stock was added to each culture and incubated at 37°C/5%CO2 for 4 hours. After washing three times in warm PBS (to remove as much virus as possible) the cultures were incubated in fresh medium at 35°C/3%CO2 and monitored daily for fluorescence. Four plates of fresh B7GG cells were grown to 90% confluency, washed with 10ml PBS and detached with 6 ml of 0.25% trypsin for five minutes at room temperature (RT) with gentle rocking. A warm culture medium with 10% FBS was added to quench trypsin activity and the cells were dissociated by gentle trituration (approximately 12 passes). The cell suspensions were centrifuged at 800G for 3 mins to pellet cells. After removing the trypsin/culture medium they were re-suspended in warm culture medium. Twelve culture dishes containing 18 ml of culture medium were simultaneously inoculated with 2ml each of the B7GG cell suspension and incubated at 37°C/5%CO2. The culture medium was changed after 80-90% confluency and 4ml of viral supernatant added to each plate and incubated for 4 hours at 37°C/5%CO2., then culture medium was removed, the cells washed three times in PBS and 20ml fresh medium added to each plate. The cultures were incubated at 35°C/3%CO2 and monitored daily for expression of cytoplasmic mCherry and nuclear GFP. The cell culture medium containing RVΔG-mCherry was collected after 4 days, filtered through 0.45 μm membranes and placed on ice. RVΔG-mCherry was concentrated from 180 ml (10 plates) of cell supernatant via ultra-centrifugation, 2 hours at 20,000G at 4°C. The supernatant was aspirated, and the six viral pellets were re-suspended in 200 μL each of cold Hanks Balanced Salt Solution (HBSS). All supernatants were combined and layered over 1.8 ml of 20% sucrose in HBSS in a 3 ml tube. The virus was pelleted through the sucrose cushion by ultra-centrifugation, at 20,000G for 2 hours at 4°C. The supernatant and sucrose cushion were gently poured off and any remining fluid aspirated. The pellet was re-suspended in 105 μL of HBSS by gentle agitation and 5 μL aliquots frozen at -80°C for later use.

Virus Titer

Serial dilutions of RVmCH (from one 5 μL aliquot of virus) from the preparation were inoculated onto 1 105 293T cells in a 48 well plate. As soon as mCherry was detectable (usually 2 days), the positive cells in the well (dilution) containing 10-100 cells were counted and a titer obtained. For these experiments we used the rabies virus at a titer of 1.92 E10 transfection units (TU)/ml.

Production of AAV1-G

Adeno-associated viruses were produced by the Emory Virus Core from AAV2 plasmids expressing the B19 RV glycoprotein under a CMV promoter. This plasmid was donated by Dr. Silvia Arber (Biozentrum, Basel)(Stepien et al., 2010). Virus titer was expressed in genomic copies (GC) and was 2.5XE12 vg/ml by qPCR for the lot used here.

Viral and tracer intramuscular injections

AAV1-G was injected at P4 in the LG (1-2 μL, undiluted), RVΔG-mCherry was injected in the same muscle at P15 (2-3 μL, undiluted) and CTB was injected in the TA at the same time (0.5 μL, 1% diluted is sterile saline). The animals were allowed to survive 7 days after the last injection (P22) at which time they were perfusion-fixed, and 50 μm thick sections were prepared as above.

Immunocytochemistry

Sections were incubated overnight in a cocktail of primary antibodies that included chicken anti-GFP (Aves), rabbit anti-DsRed (Clontech), guinea-pig anti VGLUT1 (Synaptic Systems) and goat anti-CTb (List labs). Primary antibodies were detected with species-specific antibodies coupled to FITC (for EGFP), Cy3 (for tdT), Cy5 (for CTb) or biotin (for VGLUT1). Biotinylated antibodies were exposed with streptavidin-Alexa 405.

Analyses

All sections with mCherry labeled cells were imaged at low (10X), medium (20X) and high (60X) magnification using confocal microscopy. Reciprocal Ia inhibitory interneurons were defined as neurons retrogradely labeled from the LG by RVΔG-mCherry and receiving inputs from TA CTb/VGLUT1 labeled boutons. Because this technique was relatively low yield. All analyses were qualitative. The numbers of animals and yields are reported in results.

Antibody specificities

The most frequently used antibodies in this study were tested in knockout mice: guinea-pig anti-VGLUT1 and rabbit anti-calbindin (Siembab et al., 2010), rabbit anti-Pou6f2, goat anti-Sp8, and guinea-pig anti-Otp (Bikoff et al., 2016), and goat anti-Foxp2 and two rabbit anti-MafB antibodies (this study, Supplemental Figures S1 and S2). In addition, Foxp2 and MafB antibodies used here were further characterized using western blots (see below). Alternative antibodies against Pou6f2 or Otp were first confirmed in dual immunolabeling with validated antibodies. NeuroD2 and Prox1 antibodies were tested and used in recent literature (Osseward et al., 2021). Goat anti-ChAT antibodies exhibited the well-known patterns of cholinergic immunoreactive neurons in the spinal cord and coincided with genetically labeled neurons in ChAT-IRES-Cre-tdTomato mice. GFP, DsRed and RFP antibodies did not result in any immunolabeling in sections not expressing any of these fluorescent proteins. CTb antibodies resulted in no staining in naïve sections. Rabbit and mouse anti-NeuN antibodies gave identical results. The immunostaining of the mouse anti-NeuN monoclonal in the spinal cord has been amply characterized (Alvarez et al., 2004 and 2005).

Western Blots. Foxp2 and MafB antibodies were further characterized in western blots from spinal cord samples collected from wildtypes, heterozygotes (one null allele) and homozygous knockouts (both null alleles) (supplemental Figures S1 and S2). Lumbar spinal cords were dissected in oxygenated artificial cerebrospinal fluid and immediately homogenized using Cytoplasmic Extraction Reagent Kit from the NE-PERTM nuclear extraction kit (ThermoFisher) with a protease inhibitor cocktail added (5mg/ml. Complete Mini, Roche). The manufacturer’s instructions were followed to isolate nuclei and yield aliquots of nuclear proteins. A Bio-Rad DC protein assay was used to determine total nuclear protein content. Protein standards (Bio-Rad) were prepared in NER buffer from the NE-PER kit. Sample absorbance was read on a plate reader at 750 nm. Samples were stored at -80°C until use. For immunoblotting, the samples were prepared in standard SDS-PAGE sample buffer (5X) and 30 μL of nuclear protein from each spinal cord was added to each lane of a Bio-Rad precast 10% polyacrylamide gel. Bio-Rad Kaleidoscope molecular weight markers were added to one lane. Electrophoresis was carried out in Tris buffer (Bio-Rad) at 180V until the dye front reached the bottom of the gel. The proteins were then then transferred overnight onto PVDF membranes (Bio-Rad) using standard SDS-PAGE transfer buffer (Bio-Rad) and a constant 0.15 A current with gentle stirring at 4°C. For MafB antibodies, three immunostaining procedures were carried out successively on the same membrane. The membrane was washed three times with Tris Buffered Saline and Triton-X-100 (TBST) for 15 minutes and blocked with non-fat dry milk (Carnation, Nestle) for 1 hour at room temp on a rocker. The first immunostaining employed a primary antibody against MafB (Novus) at a dilution 1:500 and worked best with 2.5% NFDM in TBST and incubated overnight at 4°C. The primary antibody was detected with donkey anti-rabbit IgG secondary antibodies conjugated to horseradish peroxidase (HRP, GE Health Sciences) and immunostained bands were revealed using enhanced chemiluminescence (ECL). The blot was then stripped for 20 minutes (Reblot Plus), washed 3 times for 15 minutes in TBS, blocked in 5% NFDM for one hour and re-probed with a primary antibody against MafA (Novus) at 1:500 in TBST and the immunoreactive bands detected via ECL as above. Finally, the blot was stripped a final time using ReBlot Plus (Millipore) for 20 minutes, blocked in 5% NFDM and re-probed with an antibody against MafB (Sigma). Secondary antibodies and ECL were identical to the previous two primary antibodies. In a different sequence we substituted the MafB (Sigma) antibody for c-Maf (Novus). Similar procedures were used to detect Foxp2 immunoreactivity (Santa Cruz) except that in this case the blot was probed only once, and we used anti-goat secondary antibodies coupled to HRP. Images of Western blots are not cropped and show the full gel.

Statistical Analyses

All statistical analyses were performed in Prism (GraphPad v.9). In all cases the samples passed the normality test. When comparing multiple groups, we used one or two-way ANOVAs depending on the sample structure. To include consideration of repetitive measures in single animals (for example different motoneurons from single animals) we used Nested one-way ANOVAs. Post-hoc pair wise comparisons were always done using Bonferroni-corrected t-tests. When comparisons involved only two groups we used standard t-tests. All statistical details are provided in supplementary tables.

Figure composition

All images for presentation were obtained with an Olympus FV1000 confocal microscope and processed with Image-Pro Plus (Media Cybernetics) for optimization of image brightness and contrast. Frequently we used a high gaussian filter to increase sharpness. Figures were composed using CorelDraw (version X6 and CDR2023). Graphs were generated in Prism (GraphPad, ver 9). Color choices were always selected to optimal visualization by people with different degrees of color detection capacity. For Figure 7 we decided to superimpose the OTP signal in magenta and avoid merging the magenta color with the green EGFP which have resulted in a difficult to detect whitish green. Instead, we added the magenta on top of the green neurons. These image manipulations were performed in Image Pro-Plus by first subtracting the OTP+ labeled nucleus from the EGFP image and then adding the magenta OTP signal. Because magenta is restricted to the nucleus the EGFP (green) cells are easily detectable and the OTP+ cells within this group are best visualized.